### 2022-01-09 README.Rmd
first human commit
- Lots of changes

---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# ASQ3

<!-- badges: start -->
[![R-CMD-check](https://github.com/nim-ach/ASQ3/workflows/R-CMD-check/badge.svg)](https://github.com/nim-ach/ASQ3/actions)
[![Lifecycle: experimental](https://img.shields.io/badge/lifecycle-experimental-orange.svg)](https://lifecycle.r-lib.org/articles/stages.html#experimental)
<!-- badges: end -->

The purpose of the {ASQ3} package is to provide access to the data collected by the Centro Asistencial Docente e Investigación de la Universidad de Magallanes (CADI-UMAG).

The data were treated, eliminating any sensitive information that could compromise the identity of the participants or linked individuals.

## Installation

You can install the development version of ASQ3 from [GitHub](https://github.com/) with:

``` r
# install.packages("devtools")
devtools::install_github("nim-ach/ASQ3")
```

## Example

This is a basic example which shows you how to use the dataset from your <i class="fab fa-r-project"></i> console.

```{r example}
library(ASQ3) # Load the package

head(dataset[, 1:5], n = 5) # Print the first 5 rows and columns
```

For more information about the dataset, you can consult the package documentation.

### 2022-01-09 R/ASQ3-package.R

@@ -1,0 +1,8 @@
+#' data from the ASQ3 study from CADI-UMAG
+#' @keywords internal
+"_PACKAGE"
+
+## usethis namespace: start
+#' @importFrom lifecycle deprecated
+## usethis namespace: end
+NULL

### 2022-01-09 R/dataset.R

@@ -1,0 +1,44 @@
+#' Data from the ASQ-3 study from CADI-UMAG
+#'
+#'
+#' @format A data frame with 1266 rows y 53 variables:
+#'
+#' \itemize{
+#'   \item{\code{profesional_id}: Numeric. Identifying number of the assessing professional.}
+#'   \item{\code{profesional_especialidad}: Character. Profession of the professional evaluator.}
+#'   \item{\code{paciente_id}: Numeric. Patient identifier number.}
+#'   \item{\code{evaluacion_numero}: Numeric. Patient assessment number.}
+#'   \item{\code{evaluacion_fecha}: Date. Date of patient assessment.}
+#'   \item{\code{respondedor_vinculo}: Character. Type of relationship between the respondent and the patient.}
+#'   \item{\code{fecha_nacimiento}: Date. Patient's date of birth.}
+#'   \item{\code{edad_cronologica_meses}: Numeric. Chronological age of the patient in months.}
+#'   \item{\code{semanas_prematurez}: Numeric. Weeks of prematurity.}
+#'   \item{\code{edad_corregida_meses}: Numeric. Chronological age of the patient in months corrected for weeks of prematurity.}
+#'   \item{\code{agrupacion}: Character. Patient's diagnostic group.}
+#'   \item{\code{diagnostico_agrupado}: Character. Patient's diagnostic group.}
+#'   \item{\code{asq3_meses}: Numeric. Age in months of patient development according to ASQ 3.}
+#'   \item{\code{comunicacion_q1 - comunicacion_q6}: Numeric. Scoring of the individual questions of the communication item.}
+#'   \item{\code{comunicacion_total}: Numeric. Total score of the communication item.}
+#'   \item{\code{comunicacion_interpretacion}: Character. Interpretation of the communication item.}
+#'   \item{\code{motora_gruesa_q1 - motora_gruesa_q6}: Numeric. Scoring of the individual questions of the gross motor skills item.}
+#'   \item{\code{motora_gruesa_total}: Numeric. Total score of the gross motor skills item.}
+#'   \item{\code{motora_gruesa_interpretacion}: Character. Interpretation of the gross motor skills item.}
+#'   \item{\code{motora_fina_q1 - motora_fina_q6}: Numeric. Scoring of the individual questions of the fine motor skills item.}
+#'   \item{\code{motora_fina_total}: Numeric. Total score of the fine motor skills item.}
[…]

### 2022-01-09 README.Rmd
update README
- Only 4 columns in head(dataset)

@@
 
-head(dataset[, 1:5], n = 5) # Print the first 5 rows and columns
+head(dataset[, 1:4], n = 5) # Print the first 5 rows of the first 4 columns
 ```
 
-For more information about the dataset, you can consult the package documentation.
+For more information about the dataset, you can consult the package [documentation](https://nim-ach.github.io/ASQ3/reference/dataset.html).

### 2022-01-09 R/ASQ3-package.R
Imports data.table

@@ -4,5 +4,6 @@
 
 ## usethis namespace: start
 #' @importFrom lifecycle deprecated
+#' @import data.table
 ## usethis namespace: end
 NULL

### 2022-01-13 README.Rmd
Imports data.table
pkgdown upgrades
- Navbar now includes articles section with links to data preparation process
- `data-prep` and `dataset` now include 'seealso' section
- Hyperlinks fixed

@@
 The data were treated, eliminating any sensitive information that could compromise the identity of the participants or linked individuals.
+Data preparation process can be seen [here](reference/data-prep.html).
 
@@
 
-For more information about the dataset, you can consult the package [documentation](https://nim-ach.github.io/ASQ3/reference/dataset.html).
+For more information about the dataset, you can consult the dataset [documentation](https://nim-ach.github.io/ASQ3/reference/dataset.html).

### 2022-01-13 R/data-prep.R

@@ -1,0 +1,135 @@
+#' Data preparation process
+#'
+#' @description
+#'
+#' The \link[=dataset]{dataset} preparation
+#' process occurs in several steps, some of which involve manual processing of the
+#' data using Excel and programmatic preparation using R. The programmatic part and
+#' the steps used for data treatment are detailed here.
+#'
+#' @section From Excel to R:
+#'
+#' Once the data entry using Excel was complete, it was imported into R using the
+#' `read_excel` function of the [readxl](https://readxl.tidyverse.org/) package.
+#'
+#' ```r
+#' dataset <- readxl::read_excel("path/to/excel-file.xlsx")
+#' ```
+#'
+#' Subsequently, the column names were changed to syntactically valid versions, in
+#' this case in camel case (e.g. "Column A" to "column_a").
+#'
+#' ```r
+#' names(dataset) <- c("evaluacion_numero", "evaluacion_fecha", "respondedor_nombre", ...)
+#' ```
+#'
+#' Once completed, the data was transformed into a `data.table' format, which allows
+#' for the efficient implementation of a function (or functions) across multiple columns
+#' (using selection criteria for greater precision).
+#'
+#' ```r
+#' dataset <- data.table::as.data.table(dataset)
+#' ```
+#'
+#' @section Formatting variables:
+#'
+#' Once the import was complete, we proceeded to standardise the formatting of the
+#' columns by using regular expressions (RegEx) to select the columns, starting with
+#' the modification of the text variables.
+#'
+#' ```r
+#' col_names <- names(dataset) # for later use
+#'
+#' # Labels to "title-case"
+#' ind <- grep("nombre|interpretacion|especialidad", col_names, value = TRUE)
+#' dataset[, (ind) := lapply(.SD, stringr::str_to_title), .SDcols = ind]
+#'
+#' # Labels to "lower-case"
+#' ind <- grep("agrupacion|diagnostico", col_names, value = TRUE)
+#' dataset[, (ind) := lapply(.SD, tolower), .SDcols = ind]
+#' ```
+#'
+#' We perform a similar process with the numeric and date variables, assigning them
+#' the necessary properties to be analysed later.
+#'
+#' ```r
+#' # numerical variables
+#' ind <- grep("total", col_names, value = TRUE)
+#' dataset[, (ind) := lapply(.SD, as.numeric), .SDcols = ind]
+#'
[…]

### 2022-01-13 R/dataset.R

@@ -1,12 +1,13 @@
 #' Data from the ASQ-3 study from CADI-UMAG
 #'
 #'
-#' @format A data frame with 1266 rows y 53 variables:
+#' @format A data frame with 1266 rows y 54 variables:
 #'
 #' \itemize{
 #'   \item{\code{profesional_id}: Numeric. Identifying number of the assessing professional.}
 #'   \item{\code{profesional_especialidad}: Character. Profession of the professional evaluator.}
 #'   \item{\code{paciente_id}: Numeric. Patient identifier number.}
+#'   \item{\code{sexo}: Character. Patient sex.}
 #'   \item{\code{evaluacion_numero}: Numeric. Patient assessment number.}
 #'   \item{\code{evaluacion_fecha}: Date. Date of patient assessment.}
 #'   \item{\code{respondedor_vinculo}: Character. Type of relationship between the respondent and the patient.}
@@ -40,5 +41,9 @@
 #'   progress in children between the ages of one month to 5.5 years. For more information please see the homepage of the
 #'   questionnaire [ASQ-3](https://agesandstages.com/products-pricing/asq3/).
 #'
+#' @seealso
+#' - \link[=data-prep]{Data preparation process}
+#'
 #' @source CADI-UMAG, Punta Arenas, Chile.
 "dataset"
+

### 2022-01-17 README.Rmd
Add umag hyperlink and about section

@@
 
-The purpose of the {ASQ3} package is to provide access to the data collected by the Centro Asistencial Docente e Investigación de la Universidad de Magallanes (CADI-UMAG).
+The purpose of the {ASQ3} package is to provide access to the data collected by the Centro Asistencial Docente e Investigación de la [Universidad de Magallanes](http://www.umag.cl/) (CADI-UMAG).
 
@@
 For more information about the dataset, you can consult the dataset [documentation](https://nim-ach.github.io/ASQ3/reference/dataset.html).
+
+## About <i class="fab fa-r-project"></i>
+
+R is a high-level statistical programming language used for data analysis, as well as for the development of software tools such as websites, reports, applications, digital books, etc.
+For more information you can visit the [R-project](https://www.r-project.org/about.html) website.

### 2022-04-25 R/dataset.R
updates on data

@@ -1,22 +1,20 @@
 #' Data from the ASQ-3 study from CADI-UMAG
 #'
 #'
-#' @format A data frame with 1266 rows y 54 variables:
+#' @format A data frame with 1266 rows y 52 variables:
 #'
 #' \itemize{
 #'   \item{\code{profesional_id}: Numeric. Identifying number of the assessing professional.}
 #'   \item{\code{profesional_especialidad}: Character. Profession of the professional evaluator.}
 #'   \item{\code{paciente_id}: Numeric. Patient identifier number.}
-#'   \item{\code{sexo}: Character. Patient sex.}
-#'   \item{\code{evaluacion_numero}: Numeric. Patient assessment number.}
-#'   \item{\code{evaluacion_fecha}: Date. Date of patient assessment.}
 #'   \item{\code{respondedor_vinculo}: Character. Type of relationship between the respondent and the patient.}
+#'   \item{\code{evaluacion_tipo}: Numeric. Type of assessment.}
+#'   \item{\code{fecha_evaluacion}: Date. Date of patient assessment.}
 #'   \item{\code{fecha_nacimiento}: Date. Patient's date of birth.}
 #'   \item{\code{edad_cronologica_meses}: Numeric. Chronological age of the patient in months.}
 #'   \item{\code{semanas_prematurez}: Numeric. Weeks of prematurity.}
 #'   \item{\code{edad_corregida_meses}: Numeric. Chronological age of the patient in months corrected for weeks of prematurity.}
-#'   \item{\code{agrupacion}: Character. Patient's diagnostic group.}
-#'   \item{\code{diagnostico_agrupado}: Character. Patient's diagnostic group.}
+#'   \item{\code{diagnostico}: Character. Patient's diagnostic group.}
 #'   \item{\code{asq3_meses}: Numeric. Age in months of patient development according to ASQ 3.}
 #'   \item{\code{comunicacion_q1 - comunicacion_q6}: Numeric. Scoring of the individual questions of the communication item.}
 #'   \item{\code{comunicacion_total}: Numeric. Total score of the communication item.}

### 2022-04-25 data-raw/clean_dataset.R
data preparation files except raw-data

@@ -1,0 +1,109 @@
+# Load libraries
+library(readxl)
+library(data.table)
+
+# Import data ---------------------------------------------------------------------------------
+
+  # Import the data directly from Excel
+  dataset <- read_excel(path = "data-raw/raw-data.xlsx", sheet = 2)[,-1L]
+
+  ## Expect warning message:
+  ## Expecting date in B1415 / R1415C2: got '2/15/0222'
+  ##
+  ## Note: Expected and assumed as a data-entry error
+
+  # Modify variable names
+  names(dataset) <- readLines("data-raw/helpers/colnames")
+
+  # Set data to data.table
+  setDT(dataset)
+
+
+# Treat data ----------------------------------------------------------------------------------
+
+
+  col_names <- names(dataset) # for later use
+
+  # Labels to "title-case"
+  ind <- grep("nombre|interpretacion|especialidad", col_names, value = TRUE)
+  dataset[, (ind) := lapply(.SD, stringr::str_to_title), .SDcols = ind]
+
+  # Labels to "lower-case"
+  ind <- grep("agrupacion|diagnostico", col_names, value = TRUE)
+  dataset[, (ind) := lapply(.SD, tolower), .SDcols = ind]
+
+  # numerical variables
+  ind <- grep("total", col_names, value = TRUE)
+  dataset[, (ind) := lapply(.SD, as.numeric), .SDcols = ind]
+
+  # asq3_meses to numeric
+  dataset[, asq3_meses := as.numeric(x = gsub("MESES", "", asq3_meses))]
+
+  # Processing dates
+  ind <- grep("fecha", col_names, value = TRUE)
+
+  # Fix individual cases for dates issues (data-entry errors)
+  source("data-raw/helpers/fix_dates.R")
+
+  dataset[, fecha_nacimiento := as.numeric(fecha_nacimiento)]
+  dataset[, (ind) := lapply(.SD, data.table::as.IDate, origin = "1899-12-30"), .SDcols = ind]
+
+  # Removing ind and col_names objects
+  rm(ind, col_names)
+
+
+# Subject anonymisation -----------------------------------------------------------------------
+
+
+  # Drop missing values
+  dataset <- dataset[!is.na(rut_paciente) & !is.na(nombre_paciente)]
+
+  # ID for each professional
+  dataset[, profesional_id := as.numeric(x = factor(profesional_nombre))]
+  # and then we eliminate the column with their name
[…]

### 2022-04-25 data-raw/helpers/fix_dates.R

@@ -1,0 +1,7 @@
+fe_ind <- c(1414)
+fn_ind <- c(483, 647, 962, 1187, 1330)
+
+dataset[fe_ind, fecha_evaluacion := as.Date("2022-02-15")]
+dataset[fn_ind, fecha_nacimiento := c("42483", "43530", "43483", "43599", "43862")]
+
+rm(fe_ind, fn_ind)

### 2022-04-26 R/dataset.R
update on data format
- transform some variables to factor (k = 10)

@@ -4,11 +4,11 @@
 #' @format A data frame with 1266 rows y 52 variables:
 #'
 #' \itemize{
-#'   \item{\code{profesional_id}: Numeric. Identifying number of the assessing professional.}
-#'   \item{\code{profesional_especialidad}: Character. Profession of the professional evaluator.}
-#'   \item{\code{paciente_id}: Numeric. Patient identifier number.}
-#'   \item{\code{respondedor_vinculo}: Character. Type of relationship between the respondent and the patient.}
-#'   \item{\code{evaluacion_tipo}: Numeric. Type of assessment.}
+#'   \item{\code{profesional_id}: Factor. Identifying number of the assessing professional.}
+#'   \item{\code{profesional_especialidad}: Factor. Profession of the professional evaluator.}
+#'   \item{\code{paciente_id}: Factor. Patient identifier number.}
+#'   \item{\code{respondedor_vinculo}: Factor. Type of relationship between the respondent and the patient.}
+#'   \item{\code{evaluacion_tipo}: Factor. Type of assessment.}
 #'   \item{\code{fecha_evaluacion}: Date. Date of patient assessment.}
 #'   \item{\code{fecha_nacimiento}: Date. Patient's date of birth.}
 #'   \item{\code{edad_cronologica_meses}: Numeric. Chronological age of the patient in months.}
@@ -18,19 +18,19 @@
 #'   \item{\code{asq3_meses}: Numeric. Age in months of patient development according to ASQ 3.}
 #'   \item{\code{comunicacion_q1 - comunicacion_q6}: Numeric. Scoring of the individual questions of the communication item.}
 #'   \item{\code{comunicacion_total}: Numeric. Total score of the communication item.}
-#'   \item{\code{comunicacion_interpretacion}: Character. Interpretation of the communication item.}
+#'   \item{\code{comunicacion_interpretacion}: Factor. Interpretation of the communication item.}
 #'   \item{\code{motora_gruesa_q1 - motora_gruesa_q6}: Numeric. Scoring of the individual questions of the gross motor skills item.}
 #'   \item{\code{motora_gruesa_total}: Numeric. Total score of the gross motor skills item.}
-#'   \item{\code{motora_gruesa_interpretacion}: Character. Interpretation of the gross motor skills item.}
+#'   \item{\code{motora_gruesa_interpretacion}: Factor. Interpretation of the gross motor skills item.}
[…]

### 2022-04-26 data-raw/clean_dataset.R

@@ -10,7 +10,8 @@ library(data.table)
   ## Expect warning message:
   ## Expecting date in B1415 / R1415C2: got '2/15/0222'
   ##
-  ## Note: Expected and assumed as a data-entry error
+  ## Note: Expected and assumed as a data-entry error.
+  ## These error were corrected in helpers/fix_dates.R
 
   # Modify variable names
   names(dataset) <- readLines("data-raw/helpers/colnames")
@@ -98,10 +99,24 @@ library(data.table)
   rm(edad_corregida)
 
 
+# Set factor variables ------------------------------------------------------------------------
+
+  dataset[, `:=`(
+    profesional_id = factor(profesional_id),
+    profesional_especialidad = factor(profesional_especialidad),
+    paciente_id = factor(paciente_id),
+    respondedor_vinculo = factor(respondedor_vinculo),
+    evaluacion_tipo = factor(evaluacion_tipo),
+    comunicacion_interpretacion = factor(comunicacion_interpretacion),
+    motora_fina_interpretacion = factor(motora_fina_interpretacion),
+    motora_gruesa_interpretacion = factor(motora_gruesa_interpretacion),
+    socio_individual_interpretacion = factor(socio_individual_interpretacion),
+    resolucion_problemas_interpretacion = factor(resolucion_problemas_interpretacion)
+  )]
+
 # Reorder columns -----------------------------------------------------------------------------
 
   setcolorder(dataset, neworder = readLines("data-raw/helpers/colorder"))
-
 
 # Export dataset ------------------------------------------------------------------------------
 

### 2022-04-27 data-raw/clean_dataset.R
update on data import process

@@ -11,7 +11,7 @@ library(data.table)
   ## Expecting date in B1415 / R1415C2: got '2/15/0222'
   ##
   ## Note: Expected and assumed as a data-entry error.
-  ## These error were corrected in helpers/fix_dates.R
+  ## These errors will be corrected later on.
 
   # Modify variable names
   names(dataset) <- readLines("data-raw/helpers/colnames")
@@ -20,10 +20,40 @@ library(data.table)
   setDT(dataset)
 
 
+# Fix data-entry errors -----------------------------------------------------------------------
+
+  corrections <- read.csv("data-raw/helpers/corrections.csv")
+  n <- nrow(corrections)
+
+  # `fecha_evaluacion` as character (latter on will be transformed to date format)
+  dataset[, fecha_evaluacion := as.character(fecha_evaluacion)]
+
+  for (i in seq_len(n)) {
+    i_var  <- corrections[i, "param_in"]
+
+    ind <- like(
+      vector = dataset[[i_var]],
+      pattern = corrections[i, "search_for"]
+    )
+
+    set(
+      x = dataset,
+      i = which(ind),
+      j = corrections[i, "param_out"],
+      value = corrections[i, "replace"]
+    )
+  }
+
+  rm(corrections, i, i_var, ind, n)
+
 # Treat data ----------------------------------------------------------------------------------
 
-
   col_names <- names(dataset) # for later use
+
+  # Processing date type variables
+  ind <- grep("fecha", col_names, value = TRUE)
+  dataset[, fecha_nacimiento := as.numeric(fecha_nacimiento)]
+  dataset[, (ind) := lapply(.SD, data.table::as.IDate, origin = "1899-12-30"), .SDcols = ind]
 
   # Labels to "title-case"
   ind <- grep("nombre|interpretacion|especialidad", col_names, value = TRUE)
@@ -40,15 +70,6 @@ library(data.table)
   # asq3_meses to numeric
   dataset[, asq3_meses := as.numeric(x = gsub("MESES", "", asq3_meses))]
 
-  # Processing dates
-  ind <- grep("fecha", col_names, value = TRUE)
-
-  # Fix individual cases for dates issues (data-entry errors)
-  source("data-raw/helpers/fix_dates.R")
-
-  dataset[, fecha_nacimiento := as.numeric(fecha_nacimiento)]
-  dataset[, (ind) := lapply(.SD, data.table::as.IDate, origin = "1899-12-30"), .SDcols = ind]
-
[…]

### 2022-04-27 data-raw/helpers/fix_dates.R

@@ -1,7 +1,0 @@
-fe_ind <- c(1414)
-fn_ind <- c(483, 647, 962, 1187, 1330)
-
-dataset[fe_ind, fecha_evaluacion := as.Date("2022-02-15")]
-dataset[fn_ind, fecha_nacimiento := c("42483", "43530", "43483", "43599", "43862")]
-
-rm(fe_ind, fn_ind)

### 2022-04-27 data-raw/clean_dataset.R
data is now sorted on patient id and evaluation date

@@ -141,6 +141,7 @@ library(data.table)
 # Reorder columns -----------------------------------------------------------------------------
 
   setcolorder(dataset, neworder = readLines("data-raw/helpers/colorder"))
+  setkey(dataset, paciente_id, fecha_evaluacion)
 
 # Export dataset ------------------------------------------------------------------------------
 

### 2022-05-05 R/dataset.R
update on raw-data cleaning
- Missing values on sex
- Update on `vinculo_paciente`

@@ -1,7 +1,7 @@
 #' Data from the ASQ-3 study from CADI-UMAG
 #'
 #'
-#' @format A data frame with 1266 rows y 53 variables:
+#' @format A data frame with 1523 rows y 53 variables:
 #'
 #' \itemize{
 #'   \item{\code{profesional_id}: Factor. Identifying number of the assessing professional.}

### 2022-05-05 data-raw/clean_dataset.R

@@ -61,7 +61,7 @@ library(data.table)
   dataset[, (ind) := lapply(.SD, stringr::str_to_title), .SDcols = ind]
 
   # Labels to "lower-case"
-  ind <- grep("agrupacion|diagnostico", col_names, value = TRUE)
+  ind <- grep("diagnostico", col_names, value = TRUE)
   dataset[, (ind) := lapply(.SD, tolower), .SDcols = ind]
 
   # numerical variables
@@ -77,6 +77,8 @@ library(data.table)
     y = fread("data-raw/helpers/rut_sex.csv"),
     by = "rut_paciente",
   )
+  # Removing missing values from sex
+  dataset[sexo_paciente == "", sexo_paciente := NA]
 
   # Removing ind and col_names objects
   rm(ind, col_names)

### 2022-05-09 R/dataset.R
add new vignette
- non-linear analysis using generalized additive models (GAMs)

@@ -9,6 +9,7 @@
 #'   \item{\code{paciente_id}: Factor. Patient identifier number.}
 #'   \item{\code{sexo_paciente}: Factor. Patient's sex.}
 #'   \item{\code{respondedor_vinculo}: Factor. Type of relationship between the respondent and the patient.}
+#'   \item{\code{n_evaluacion}: Factor. Number of assessment for each subject.}
 #'   \item{\code{evaluacion_tipo}: Factor. Type of assessment.}
 #'   \item{\code{fecha_evaluacion}: Date. Date of patient assessment.}
 #'   \item{\code{fecha_nacimiento}: Date. Patient's date of birth.}

### 2022-05-09 data-raw/clean_dataset.R

@@ -80,6 +80,10 @@ library(data.table)
   # Removing missing values from sex
   dataset[sexo_paciente == "", sexo_paciente := NA]
 
+  # Including evaluation number
+  dataset[, n_evaluacion := rleid(fecha_evaluacion), keyby = rut_paciente]
+
+
   # Removing ind and col_names objects
   rm(ind, col_names)
 
@@ -136,7 +140,7 @@ library(data.table)
 # Set factor variables ------------------------------------------------------------------------
 
   ind <- grep(
-    pattern = "profesional|paciente|vinculo|tipo|interpret",
+    pattern = "profesional|paciente|vinculo|evaluacion|interpret",
     x = names(dataset),
     value = TRUE
   )

### 2022-05-19 R/dataset.R
website update using only those with a diagnosis of congenital hypotonia.

@@ -1,7 +1,7 @@
 #' Data from the ASQ-3 study from CADI-UMAG
 #'
 #'
-#' @format A data frame with 1523 rows y 53 variables:
+#' @format A data frame with 1523 rows y 55 variables:
 #'
 #' \itemize{
 #'   \item{\code{profesional_id}: Factor. Identifying number of the assessing professional.}
@@ -17,6 +17,7 @@
 #'   \item{\code{semanas_prematurez}: Numeric. Weeks of prematurity.}
 #'   \item{\code{edad_corregida_meses}: Numeric. Chronological age of the patient in months corrected for weeks of prematurity.}
 #'   \item{\code{diagnostico}: Character. Patient's diagnostic group.}
+#'   \item{\code{alteracion}: Character. Patient's alteration group.}
 #'   \item{\code{asq3_meses}: Numeric. Age in months of patient development according to ASQ 3.}
 #'   \item{\code{comunicacion_q1 - comunicacion_q6}: Numeric. Scoring of the individual questions of the communication item.}
 #'   \item{\code{comunicacion_total}: Numeric. Total score of the communication item.}

### 2022-05-19 data-raw/clean_dataset.R

@@ -76,6 +76,7 @@ library(data.table)
     x = dataset,
     y = fread("data-raw/helpers/rut_sex.csv"),
     by = "rut_paciente",
+    all = TRUE
   )
   # Removing missing values from sex
   dataset[sexo_paciente == "", sexo_paciente := NA]
@@ -136,6 +137,15 @@ library(data.table)
   # Removing function already used
   rm(edad_corregida)
 
+
+# create clusters for diagnostics -------------------------------------------------------------
+
+  dataset <- merge(
+    x = dataset,
+    y = fread("data-raw/helpers/diagnostic_group.csv"),
+    by = "diagnostico",
+    all = TRUE
+  )
 
 # Set factor variables ------------------------------------------------------------------------
 

### 2022-06-08 manuscript/manuscript.Rmd
add manuscript files

---
title: ""
bibliography: "misc/ref.bib"
csl: "misc/vancouver.csl"
output: 
  word_document:
    reference_docx: "ref.docx"
---

**Título**: [título]

**Title**: [title]

**Autores**: Primer autor,^1^ Segundo autor,^2^ Tercer autor,^3^ Cuarto autor,^4^ Quinto autor,^5^ Sexto autor.^6^ \*

::: {custom-style="Afiliation"}
^1^ Afiliación, lugar.

^2^ Afiliación, lugar.

^3^ Afiliación, lugar.

^4^ Afiliación, lugar.

^5^ Afiliación, lugar.

^6^ Afiliación, lugar.
:::

### \*Autor de correspondencia

Autor, Afiliación, lugar. e-mail: autor\@mail.com. 
Dirección: XXXX. Teléfono: XXXX.

\newpage

## Resumen

**Objetivo**: [...]. **Material y métodos**: [...]. **Resultados**: [...]. **Conclusión**: [...].

**Palabras clave**: [...].

\newpage

## Abstract

**Objective**: [...]. **Material and methods**: [...]. **Results**: [...]. **Conclusion**: [...].

**Keywords**: [...].

\newpage

# Introducción

<!---- Antecedentes históricos ---->
[...].

<!---- Conceptualización del problema ---->
[...].

<!---- Hipótesis y objetivos ---->
[...].

# Material y métodos

## Participantes

[...].

## Instrumentos

### Instrumento 1

[...].

### Instrumento 2

[...].

## Procedimiento

[...].

[...].

## Análisis estadístico

[...].

[...], using the *R* programming language for statistical computing, [@rlanguage].

# Resultados

[...].

[...].

# Discusión

[...].

[...].

[...].

# Conclusión

[...].

# Agradecimientos

[...].

# Conflictos de interés

[...].

# Referencias

<div id="refs"></div>

### 2022-06-19 manuscript/manuscript.Rmd
major features added for gam models reporting and functionalities

@@
 bibliography: "misc/ref.bib"
-csl: "misc/vancouver.csl"
+#csl: "misc/vancouver.csl"
 output: 
   word_document:
-    reference_docx: "ref.docx"
+    reference_docx: "misc/ref.docx"
 ---
 
-**Título**: [título]
+```{r setup, include=FALSE}
+knitr::opts_chunk$set(
+  echo = FALSE,
+  include = FALSE
+)
+```
 
@@
 
-**Autores**: Primer autor,^1^ Segundo autor,^2^ Tercer autor,^3^ Cuarto autor,^4^ Quinto autor,^5^ Sexto autor.^6^ \*
+**Authors**: Primer autor,^1^ Segundo autor,^2^ Tercer autor,^3^ Cuarto autor,^4^ Quinto autor,^5^ Sexto autor.^6^ \*
 
@@
 
-## Resumen
-
-**Objetivo**: [...]. **Material y métodos**: [...]. **Resultados**: [...]. **Conclusión**: [...].
-
-**Palabras clave**: [...].
-
-\newpage
-
 ## Abstract
@@
 
-# Introducción
+# Introduction
 
@@
 
-# Material y métodos
+# Material y methods
 
-## Participantes
+## Participants
 
@@
 
-## Instrumentos
+## Instruments
 
-### Instrumento 1
+### Instrument 1
 
@@
 
-### Instrumento 2
+### Instrument 2
 
@@
 
-## Procedimiento
+## Procedures
 
@@
 
-## Análisis estadístico
+## Statistical analysis
 
-[...].
+Data is presented as median (*Mdn*) and interquartile range (*IQR*) for continuous variables, and as absolute and relative sample size for categorical/discrete variables.
 
-[...], using the *R* programming language for statistical computing, [@rlanguage].
+A non-parametric approach was used since the underlying distribution of continuous outcomes, evaluated through analytical and graphical methods, did not follow a Gaussian distribution.
+
+In order assess the differences in developmental scores between males and females, the *Wilcoxon* rank-sum test was used, meanwhile the chi-square test ($\chi^2$) was used to evaluate the goodness-of-fit ($\chi^2_{gof}$) and the independence of factors ($\chi^2_{Pearson}$).
+
+Generalized additive models (GAM) were used to describe linear and non-linear relationships between developmental characteristics through smooth terms that are represented using penalized regression splines [@wood2011fast].
+The restricted maximum likelihood estimation method was used for the selection of the smoothing parameter, and the thin-plate regression spline as the smoothing basis method, as they are the optimal smoother of any given basis dimension/rank [@wood2003thin].
+GAMs were fitted using the mcgv package [@wood2017generalized].
+To describe the smooth term, we use approximative derivative to summarise de trend in terms of linear segments.
+
+[...], using the *R* programming language for statistical computing, [@rlanguage] and complementary R packages for model fitting, visualization and complementary computations [@wood2017generalized; @hadley2016ggplot2; @daniel2021see; @dominique2020estimation].
 
@@
 
-[...].
+```{r load-packages}
+source(file = "manuscript/misc/setup.R")
+```
 
-[...].
+```{r}
+# Sample
+n_sample <- dataset[, .N]
+#> 234
 
-# Discusión
+# Sample by sex
+n_sex <- dataset[j = .(stat = paste0(.N, " (", round(.N*100/n_sample, 1), "%)")), 
+                 keyby = .(sex = sexo_paciente)]
+#>    sex        stat
+#> 1:   F  94 (40.2%)
+#> 2:   M 140 (59.8%)
+
+# Prop test on sex
+prop_sex <- dataset[, table(sexo_paciente)] |> 
+  chisq.test() |> 
+  report_chi(effsize = TRUE)
+#> [1] "$\\chi^2_{gof}$ (1) = 9.04, *p* = 0.003, $C_{Pearson}$ = 0.19, CI~0.95%~[0.09, 1]"
+
+```
+
+From a total of `r n_sample` subjects with congenital hypotonia, `r n_sex["F", stat]` were females and `r n_sex["M", stat]` males (`r prop_sex`).
+The developmental characteristics of the sample can be seen in [Table 1](#tab1).
+
+```{r}
+mod <- gam(comunicacion_total ~ s(edad_cronologica_meses), 
+    data = dataset,
+    method = "REML")
+
+# Overall slope
+gam_coef <- report_slope(mod, trend = c("edad_cronologica_meses"))
+
+# Linear segments
+modelbased::estimate_smooth(
+  data = modelbased::estimate_relation(mod, length = 100),
+  x = "edad_cronologica_meses"
+)
+
+plot_gam(dataset, edad_cronologica_meses, comunicacion_total, round_to = 5,
+         "Chronological age (months)", "Communication score", method = "REML")
+```
+
+When modelling the effect of chronological age on developmental skills, we found an averaged negative marginal effect on the communication score (`r gam_coef`), whereas from 1 to 5.7 moths old
+
+# Discusion
 
@@
 
-# Conclusión
+# Conclusion
 
@@
 
-# Agradecimientos
+# Acknowledgment
 

### 2022-06-19 R/plot_gam.R

@@ -1,0 +1,59 @@
+#' Round to nearest multiple
+#'
+#' Internal function used inside `plot_gam()` function.
+#' It takes a `i` number and rounds it to the near `j` multiple.
+#'
+#' @param i Numeric. Number to be rounded.
+#' @param j Numeric. Multiple to round.
+#'
+#' @export
+
+near <- function(i, j) {
+  stopifnot("'i' must be a number" = is.numeric(i),
+            "'j' must be a number" = is.numeric(j))
+
+  i_hat = i / j
+  output = round(i_hat) * j
+  return(output)
+}
+
+#' Plot GAM model of bivariate model
+#'
+#' Display graphical representation of a gam model for bivariate
+#' relationships. Model specification can be set with the ellipsis
+#' argument (dot-dot-dot).
+#'
+#' @param data Dataset from where to extract x and y variables.
+#' @param x Numeric. Variable for the x-axis.
+#' @param y Numeric. Variable for the y-axis.
+#' @param xlab Character. Label for x in the plot.
+#' @param ylab Character. Label for y in the plot.
+#' @param round_to Numeric. Intervals for mean and SE calculations.
+#' @param ... Further arguments for `method.args` in `geom_smooth`.
+#'
+#' @export
+
+plot_gam <- function(data, x, y, xlab, ylab, round_to = 1, ...) {
+
+  if (missing(xlab)) xlab <- deparse(substitute(x))
+  if (missing(ylab)) ylab <- deparse(substitute(y))
+
+  scaled = copy(data)
+  if (round_to > 1) {
+    j_exec = substitute(near(x, round_to))
+    x_var = deparse(substitute(x))
+    scaled[, (x_var) := eval(j_exec)]
+  }
+
+  ggplot2::ggplot(data, ggplot2::aes({{x}}, {{y}})) +
+    ggplot2::geom_count(alpha = 0.1) +
+    ggplot2::stat_summary(data = scaled, fun.data = ggplot2::mean_se) +
+    ggplot2::geom_smooth(method = "gam",
+                         formula = y ~ s(x),
+                         col = "red4",
+                         method.args = list(...)) +
+    ggplot2::labs(x = xlab, y = ylab, caption = paste0("Mean ",
+                                                       rawToChar(x = as.raw(x = c(0xc2, 0xb1))),
+                                                       " SE in ", round_to, "-unit intervals")) +
+    see::theme_lucid()
+}

### 2022-06-19 R/report_stats.R

@@ -1,0 +1,146 @@
+#' Format p-values
+#'
+#' Internal function to round p-values
+#'
+#' @param x Numeric. P-value to be formated.
+
+format_p <- function(x) {
+  stopifnot(
+    "x is not numeric" = is.numeric(x)
+  )
+
+  x <- data.table::fifelse(
+    test = x < 0.001,
+    yes = "< 0.001",
+    no = paste("=", round(x, 3))
+  )
+
+  return(x)
+}
+
+#' Report chi-square test
+#'
+#' This function lets you pass a chi-square function and report it.
+#'
+#' @param chi A chi-square object.
+#' @param effsize Logical. Whether to report effect sizes or not.
+#' @param ci Numeric. Confidence interval to effect size calculation.
+#'
+#' @name report_chi
+#' @export
+
+report_chi <- function(chi, effsize = TRUE, ci = 0.95) {
+
+  stopifnot(
+    "Is not an object of class \"htest\"" = inherits(chi, "htest")
+  )
+
+  if (isTRUE(effsize)) {
+    efs <- effectsize::pearsons_c(chi, ci = ci)
+    efs <- lapply(efs, round, 2)
+  }
+
+  chi$statistic <- round(chi$statistic, 2)
+
+  chi$p.value <- format_p(chi$p.value)
+
+  chi$method <- data.table::fifelse(
+    test = grepl("pearson", x = tolower(chi$method)),
+    yes = "Pearson",
+    no = "gof"
+  )
+
+  expr <- paste(
+    paste0("$\\chi^2_{",chi$method,"}$ (",chi$parameter,") = ",chi$statistic),
+    paste0("*p* ",chi$p.value),
+    sep = ", "
+  )
+
+  if (isTRUE(effsize)) {
+    expr <- paste(
+      expr,
+      paste0("$C_{Pearson}$ = ", efs[[1]]),
+      paste0("CI~", efs$CI,"%~[", efs$CI_low,", ",efs$CI_high,"]"),
+      sep = ", "
+    )
+  }
+
+  return(expr)
+}
+
+#' Report GAM model overall slope
+#'
+#' Report the overall slopes for a GAM model fitted through the `mgcv` package.
+#'
+#' @param model Object of class "gam".
+#' @param ci Numeric. Confidence interval for coefficients. Defaults to 0.95.
[…]

### 2022-06-19 manuscript/misc/setup.R

@@ -1,0 +1,24 @@
+# Load the necessary packages
+library(ASQ3)
+library(data.table)
+
+# For tables
+library(gtsummary)
+library(gt)
+
+# For GAMS
+library(mgcv)
+
+# Load the data
+data("dataset", package = "ASQ3")
+
+# Work only with a subset of data whom:
+# 1. Includes only observations from assessment 1
+# 2. Includes subjects with diagnosis of congenital hypotonia
+# 3. Subjects whose age is not greater than 60 months old (chronological age)
+dataset <- dataset[i = n_evaluacion == 1 & diagnostico %like% "hipotonia" & edad_cronologica_meses <= 60]
+
+# Drop levels of unused factors
+fv <- dataset[, names(.SD), .SDcols = is.factor]
+dataset[, (fv) := lapply(.SD, droplevels), .SDcols = fv]
+rm(fv)

### 2022-06-19 manuscript/misc/table.R

@@ -1,0 +1,44 @@
+
+# Table 1 -------------------------------------------------------------------------------------
+
+tbl_data <- copy(dataset)[, sexo_paciente := `levels<-`(sexo_paciente, c("Female", "Male"))]
+
+tbl1 <- tbl_summary(tbl_data,
+                    by = "sexo_paciente",
+                    include = c(
+                      "semanas_prematurez",
+                      "edad_cronologica_meses",
+                      "edad_corregida_meses",
+                      "asq3_meses",
+                      "comunicacion_total",
+                      "motora_gruesa_total",
+                      "motora_fina_total",
+                      "resolucion_problemas_total",
+                      "socio_individual_total",
+                    ),
+                    label = list(
+                      semanas_prematurez ~ "Prematurity weeks",
+                      edad_cronologica_meses ~ "Chronological age",
+                      edad_corregida_meses ~ "Corrected age",
+                      asq3_meses ~ "Developmental age (ASQ-3)",
+                      comunicacion_total ~ "Communication score",
+                      motora_gruesa_total ~ "Gross motor score",
+                      motora_fina_total ~ "Fine motor score",
+                      resolucion_problemas_total ~ "Problem solving score",
+                      socio_individual_total ~ "Socio-individual score"
+                    ), digits = everything() ~ 1)
+
+tbl1 <- add_overall(x = tbl1)
+
+tbl1 <- add_p(
+  x = tbl1,
+  test = everything() ~ "wilcox.test",
+  pvalue_fun = function(x) style_pvalue(x, digits = 3)
+)
+
+tbl1 <- modify_header(
+  x = tbl1,
+  all_stat_cols(stat_0 = FALSE) ~ "**{level}**<br>N = {n}",
+  stat_0 = "**Overall**<br>N = {N}"
+)
+

### 2022-06-20 R/report_stats.R
more GAM functionalities

@@ -2,7 +2,7 @@
 #'
 #' Internal function to round p-values
 #'
-#' @param x Numeric. P-value to be formated.
+#' @param x Numeric. p-value to be formatted.
 
 format_p <- function(x) {
   stopifnot(
@@ -78,7 +78,7 @@ report_chi <- function(chi, effsize = TRUE, ci = 0.95) {
 #'
 #' @export
 
-report_slope <- function(model, ..., ci = 0.95) {
+report_overall <- function(model, ..., ci = 0.95) {
 
   stopifnot(
     "Is not a GAM model" = inherits(model, "gam")
@@ -90,10 +90,10 @@ report_slope <- function(model, ..., ci = 0.95) {
   slopes$t <- round(slopes$t, 2)
 
   expr <- paste(
-    paste0("$t_{student}$ (", slopes$df_error, ") = ", slopes$t),
-    paste0("*p* ", format_p(slopes$p)),
     paste0("$\\beta$ = ", round(slopes$Coefficient, 2)),
     paste0("CI~",slopes$CI*100,"%~[", round(slopes$CI_low, 2), ", ", round(slopes$CI_high, 2), "]"),
+    paste0("$t_{student}$ (", slopes$df_error, ") = ", slopes$t),
+    paste0("*p* ", format_p(slopes$p)),
     sep = ", "
   )
 
@@ -109,12 +109,69 @@ report_slope <- function(model, ..., ci = 0.95) {
   return(expr)
 }
 
-#' Report smooth term from a GAM model
+#' Report linear slopes from a GAM model
+#'
+#' Report the linear combinations (i.e., slopes) for a GAM model fitted through effect derivatives.
+#'
+#' @param model Object of class "gam".
+#' @param term Character. Term present in the model for which estimate an effect
+#' @param ci Numeric. Confidence interval for coefficients. Defaults to 0.95.
+#' @param length Numeric. This arguments controls the number of (equally spread) values that will be taken to represent the continuous variable.
+#' @param k Numeric. Number of decimal places when printing numeric variables (Defaults to 1). Except for expression column.
+#' @param ... Further parameters passed to the `estimate_sloples()` from `modelbased` package.
+#'
+#' @export
+
+report_slopes <- function(model, term, ci = 0.95, length = 100, k = 1, ...) {
[…]

### 2022-06-20 manuscript/manuscript.Rmd
advances with manuscript

@@
 
-Data is presented as median (*Mdn*) and interquartile range (*IQR*) for continuous variables, and as absolute and relative sample size for categorical/discrete variables.
+Data is presented as median (*Mdn*) and interquartile range (*IQR*) for continuous variables; for categorical/discrete variables, the absolute and relative sample size was reported.
 
-A non-parametric approach was used since the underlying distribution of continuous outcomes, evaluated through analytical and graphical methods, did not follow a Gaussian distribution.
+A non-parametric approach was used since the underlying distribution of continuous measured outcomes, assessed through analytical and graphical methods, did not follow a Gaussian distribution.
 
@@
 
-Generalized additive models (GAM) were used to describe linear and non-linear relationships between developmental characteristics through smooth terms that are represented using penalized regression splines [@wood2011fast].
-The restricted maximum likelihood estimation method was used for the selection of the smoothing parameter, and the thin-plate regression spline as the smoothing basis method, as they are the optimal smoother of any given basis dimension/rank [@wood2003thin].
-GAMs were fitted using the mcgv package [@wood2017generalized].
-To describe the smooth term, we use approximative derivative to summarise de trend in terms of linear segments.
+Generalized additive models (GAM) were used to describe linear and non-linear relationships in the form of smooth terms between developmental characteristics, represented through penalized regression splines [@wood2011fast].
+Restricted maximum likelihood (i.e., REML) method was used for the estimation of the smoothing parameters, and thin-plate regression splines for smoothing basis, as they are the optimal smoother of any given basis dimension/rank [@wood2003thin].
+To describe the smooth terms by means of quasi-linear segments, we used approximative effect derivatives to summarise the trend with 95% confidence intervals.
 
-[...], using the *R* programming language for statistical computing, [@rlanguage] and complementary R packages for model fitting, visualization and complementary computations [@wood2017generalized; @hadley2016ggplot2; @daniel2021see; @dominique2020estimation].
+A probability of committing a type I ($\alpha$) error of less than 5% (*p* < 0.05), was considered sufficient evidence for statistical significance in hypothesis testing.
+The *R* programming language was used for statistical computing [@rlanguage].
+GAMs and the corresponding model estimates were computed using the *R* packages *mcgv* and *modelbased* [@wood2017generalized; @dominique2020estimation].
+Complementary R packages were used for visualization purposes [@hadley2016ggplot2; @daniel2021see].
 
@@
 ```{r}
-mod <- gam(comunicacion_total ~ s(edad_cronologica_meses), 
-    data = dataset,
-    method = "REML")
+mod <- gam(comunicacion_total ~ s(edad_cronologica_meses), data = dataset, method = "REML")
 
 # Overall slope
-gam_coef <- report_slope(mod, trend = c("edad_cronologica_meses"))
-
-# Linear segments
-modelbased::estimate_smooth(
-  data = modelbased::estimate_relation(mod, length = 100),
-  x = "edad_cronologica_meses"
-)
+(gam_coeffs <- report_overall(mod, "edad_cronologica_meses"))
+(gam_slopes <- report_slopes(mod, "edad_cronologica_meses"))
+(gam_smooth <- report_smooth(mod))
 
@@
 
-When modelling the effect of chronological age on developmental skills, we found an averaged negative marginal effect on the communication score (`r gam_coef`), whereas from 1 to 5.7 moths old
+When modelling the effect of chronological age on developmental skills, we found a significant non-linear effect (`r gam_smooth`), that reflect an averaged negative marginal effect on communication skills (`r gam_coeffs`), however, this wasn't true when assessing the direction of the effect in the age range between 1 to 7.6 (`r gam_slopes[1, expr]`), and neither in the 18 to 48 months old group (`r gam_slopes[3, expr]`), whereas the effect tend to be positive but non-significant.
+
+```{r eval=FALSE}
+mod <- gam(comunicacion_total ~ s(edad_corregida_meses), data = dataset, method = "REML")
+
+# Overall slope
+(gam_coeffs <- report_overall(mod, "edad_corregida_meses"))
+(gam_slopes <- report_slopes(mod, "edad_corregida_meses"))
+(gam_smooth <- report_smooth(mod))
+
+plot_gam(dataset, edad_corregida_meses, comunicacion_total, round_to = 5,
+         "Corrected age (months)", "Communication score", method = "REML")
+```
 

### 2022-06-20 R/plot_gam.R
Allow strings in `plot_gam()`

@@ -35,6 +35,8 @@ near <- function(i, j) {
 
 plot_gam <- function(data, x, y, xlab, ylab, round_to = 1, ...) {
 
+  if (is.character(substitute(x))) x <- as.name(x)
+  if (is.character(substitute(y))) y <- as.name(y)
   if (missing(xlab)) xlab <- deparse(substitute(x))
   if (missing(ylab)) ylab <- deparse(substitute(y))
 

### 2022-06-20 manuscript/manuscript.Rmd
figs, tables y ms update

@@
 bibliography: "misc/ref.bib"
-#csl: "misc/vancouver.csl"
+csl: "misc/vancouver.csl"
+link-citations: true
 output: 
@@
 
+To account for any source of variability coming from subject's sex, evaluators and the type of relationship between the infants with the respondents, we incorporate them as random effects in the fitted models in the form of penalized parametric terms [@wood2016smoothing].
+
 A probability of committing a type I ($\alpha$) error of less than 5% (*p* < 0.05), was considered sufficient evidence for statistical significance in hypothesis testing.
-The *R* programming language was used for statistical computing [@rlanguage].
-GAMs and the corresponding model estimates were computed using the *R* packages *mcgv* and *modelbased* [@wood2017generalized; @dominique2020estimation].
+All the statistical analyses were computed and implemented in the R programming language [@rlanguage].
+GAMs and the corresponding model estimates were computed using the *mgcv* and *modelbased* packages with their corresponding documentation [@wood2017generalized; @dominique2020estimation].
 Complementary R packages were used for visualization purposes [@hadley2016ggplot2; @daniel2021see].
 
-# Resultados
+# Results
 
@@
 From a total of `r n_sample` subjects with congenital hypotonia, `r n_sex["F", stat]` were females and `r n_sex["M", stat]` males (`r prop_sex`).
-The developmental characteristics of the sample can be seen in [Table 1](#tab1).
+The developmental characteristics of the sample can be seen in Table 1.
 
 ```{r}
-mod <- gam(comunicacion_total ~ s(edad_cronologica_meses), data = dataset, method = "REML")
+mod <- gam(
+  comunicacion_total ~ 
+    s(profesional_id, sexo_paciente, respondedor_vinculo, bs = "re") + 
+    s(edad_cronologica_meses), 
+  data = dataset, 
+  method = "REML"
+)
+
+gam.check(mod)
 
@@
 
-When modelling the effect of chronological age on developmental skills, we found a significant non-linear effect (`r gam_smooth`), that reflect an averaged negative marginal effect on communication skills (`r gam_coeffs`), however, this wasn't true when assessing the direction of the effect in the age range between 1 to 7.6 (`r gam_slopes[1, expr]`), and neither in the 18 to 48 months old group (`r gam_slopes[3, expr]`), whereas the effect tend to be positive but non-significant.
+When modelling the effect of chronological age on developmental skills, we observed a significant non-linear relationship on communication scores (`r gam_smooth[1, expr]`), that reflect an overall negative marginal effect (`r gam_coeffs`), however, this was not true when assessing the direction of the effect in the age range between 1 to 7 (`r gam_slopes[1, expr]`), and neither in the 18 to 48 months old group (`r gam_slopes[3, expr]`), whereas the effect tend to be positive but non-significant.
+The relationship between developmental skills and corrected age can be seen in Figure 1.
 
 ```{r eval=FALSE}
-mod <- gam(comunicacion_total ~ s(edad_corregida_meses), data = dataset, method = "REML")
+mod <- gam(
+  comunicacion_total ~ 
+    s(profesional_id, respondedor_vinculo, sexo_paciente, bs = "re") + 
+    s(edad_corregida_meses), 
+  data = dataset, 
+  method = "REML"
+)
+
+gam.check(mod)
 
@@
 
-# Discusion
+# Discussion
 
@@
 
+# Data Availability Statement
+
+[...].
+
 # Acknowledgment
@@
 
-# Conflictos de interés
+# Conflicts of interest
 
@@
 
-# Referencias
+# References
 
 <div id="refs"></div>
+
+\newpage
+
+**Figure 1**.
+Relationship between corrected age (in months) and developmental domains.
+Regression lines represent fitted values estimated from GAM models.
+Points and error bars represent the mean and standard error at 5-month age intervals.

### 2022-06-20 manuscript/misc/figures.R

@@ -1,0 +1,57 @@
+
+# Load packages and scripts -------------------------------------------------------------------
+
+# Setup script
+source("manuscript/misc/setup.R")
+
+# R Packages
+library(ggplot2)
+library(see)
+
+# Figure 1 ------------------------------------------------------------------------------------
+
+# Enviromental variable
+.save_plot <- TRUE
+
+# Effect of corrected age on developmental skills (many smooths)
+# Communication score ~ corrected age
+# Gross motor score ~ corrected age
+# Fine motor score ~ corrected age
+# Problem solving score ~ corrected age
+# Socio-individual score ~ corrected age
+
+y_vars <- c("Communication" = "comunicacion_total"
+            , "Gross motor" = "motora_gruesa_total"
+            , "Fine motor" = "motora_fina_total"
+            , "Problem solving" = "resolucion_problemas_total"
+          # , "Socio-individual" = "socio_individual_total"
+          )
+
+fig_data <- melt(
+  data = dataset,
+  measure.vars = y_vars
+)
+
+fig_data[, variable := `levels<-`(variable, names(y_vars))][]
+
+scaled <- copy(fig_data)
+scaled[, edad_corregida_meses := ASQ3:::near(edad_corregida_meses, 5)][]
+
+fig_1 <- ggplot(fig_data, aes(x = edad_corregida_meses, y = value)) +
+  facet_wrap(~variable, ncol = 2) +
+  geom_count(alpha = 0.1, na.rm = TRUE) +
+  stat_summary(data = scaled, fun.data = mean_se, na.rm = TRUE) +
+  geom_smooth(method = "gam",
+              formula = y ~ s(x),
+              col = "red4",
+              method.args = list(method = "REML"),
+              na.rm = TRUE) +
+  labs(x = "Corrected age (months)", y = "Score",
+       caption = "Mean ± SE in 5-unit intervals") +
+  theme_lucid(legend.position = "top")
+
+if (isTRUE(x = .save_plot)) {
+  pdf("manuscript/fig-1.pdf", width = 7, height = 7);
+  print(fig_1);
+  dev.off()
+}

### 2022-06-20 manuscript/misc/setup.R

@@ -2,10 +2,6 @@
 library(ASQ3)
 library(data.table)
 
-# For tables
-library(gtsummary)
-library(gt)
-
 # For GAMS
 library(mgcv)
 

### 2022-06-20 manuscript/misc/table.R

@@ -1,4 +1,12 @@
 
+# Load packages and scripts -------------------------------------------------------------------
+
+# Setup script
+source("manuscript/misc/setup.R")
+
+# R Packages
+library(gtsummary)
+
 # Table 1 -------------------------------------------------------------------------------------
 
 tbl_data <- copy(dataset)[, sexo_paciente := `levels<-`(sexo_paciente, c("Female", "Male"))]

### 2022-06-20 manuscript/manuscript.Rmd
update on table and ms

@@
 
-```{r}
+```{r eval=FALSE}
 # Sample
@@
 
-From a total of `r n_sample` subjects with congenital hypotonia, `r n_sex["F", stat]` were females and `r n_sex["M", stat]` males (`r prop_sex`).
-The developmental characteristics of the sample can be seen in Table 1.
-
-```{r}
-mod <- gam(
-  comunicacion_total ~ 
-    s(profesional_id, sexo_paciente, respondedor_vinculo, bs = "re") + 
-    s(edad_cronologica_meses), 
-  data = dataset, 
-  method = "REML"
-)
-
-gam.check(mod)
-
-# Overall slope
-(gam_coeffs <- report_overall(mod, "edad_cronologica_meses"))
-(gam_slopes <- report_slopes(mod, "edad_cronologica_meses"))
-(gam_smooth <- report_smooth(mod))
-
-plot_gam(dataset, edad_cronologica_meses, comunicacion_total, round_to = 5,
-         "Chronological age (months)", "Communication score", method = "REML")
-```
-
-When modelling the effect of chronological age on developmental skills, we observed a significant non-linear relationship on communication scores (`r gam_smooth[1, expr]`), that reflect an overall negative marginal effect (`r gam_coeffs`), however, this was not true when assessing the direction of the effect in the age range between 1 to 7 (`r gam_slopes[1, expr]`), and neither in the 18 to 48 months old group (`r gam_slopes[3, expr]`), whereas the effect tend to be positive but non-significant.
-The relationship between developmental skills and corrected age can be seen in Figure 1.
+From a total of 234 subjects with congenital hypotonia, 94 (40.2%) were females and 140 (59.8%) males ($\chi^2_{gof}$ (1) = 9.04, *p* = 0.003, $C_{Pearson}$ = 0.19, CI~0.95%~[0.09, 1]).
+The developmental characteristics of the sample can be seen in [Table 1](#tab1).
 
@@
   comunicacion_total ~ 
-    s(profesional_id, respondedor_vinculo, sexo_paciente, bs = "re") + 
+    s(profesional_id, sexo_paciente, respondedor_vinculo, bs = "re") + 
     s(edad_corregida_meses), 
@@
 
-gam.check(mod)
+report_overall(mod, "edad_corregida_meses")
+#> [1] "$\\beta$ = -2.36, CI~95%~[-3.47, -1.25], $t_{student}$ (224.04) = -4.2, *p* < 0.001"
 
-# Overall slope
-(gam_coeffs <- report_overall(mod, "edad_corregida_meses"))
-(gam_slopes <- report_slopes(mod, "edad_corregida_meses"))
-(gam_smooth <- report_smooth(mod))
+report_slopes(mod, "edad_corregida_meses")
+#>    Start  End edad_corregida_meses                                                                                 expr
+#> 1:   0.0  6.8                  3.4    $\\beta$ = 0.49, CI~95%~[-0.89, 1.86], $t_{student}$ (224.04) = 0.45, *p* = 0.319
+#> 2:   7.3 17.9                 12.6 $\\beta$ = -2.08, CI~95%~[-3.17, -0.99], $t_{student}$ (224.04) = -3.79, *p* = 0.004
+#> 3:  18.4 48.0                 33.2    $\\beta$ = 0.45, CI~95%~[-1.32, 2.23], $t_{student}$ (224.04) = 0.42, *p* = 0.593
 
-plot_gam(dataset, edad_corregida_meses, comunicacion_total, round_to = 5,
-         "Corrected age (months)", "Communication score", method = "REML")
+report_smooth(mod)
+#>                                                   term                                            expr
+#> 1:                             s(edad_corregida_meses) $F_{smooth}$ (5.2, 224.04) = 13.43, *p* < 0.001
+#> 2: s(profesional_id,sexo_paciente,respondedor_vinculo)  $F_{smooth}$ (3.76, 224.04) = 0.72, *p* = 0.02
 ```
+
+When modelling the effect of chronological age corrected for prematurity on developmental domains, we observed a significant non-linear relationship on communication scores ($F_{smooth}$ (5.2, 224.04) = 13.43, *p* < 0.001), that reflect an overall negative marginal effect ($\beta$ = -2.36, CI~95%~[-3.47, -1.25], $t_{student}$ (224.04) = -4.2, *p* < 0.001), however, this was not true when assessing the direction of the effect in the age range between 1 to 7 ($\beta$ = 0.49, CI~95%~[-0.89, 1.86], $t_{student}$ (224.04) = 0.45, *p* = 0.319), and neither in the 18 to 48 months old group ($\beta$ = 0.45, CI~95%~[-1.32, 2.23], $t_{student}$ (224.04) = 0.42, *p* = 0.593), whereas the effect tend to be positive but non-significant.
+
+```{r eval=FALSE}
+mod <- gam(
+  motora_gruesa_total ~ 
+    s(profesional_id, sexo_paciente, respondedor_vinculo, bs = "re") + 
+    s(edad_corregida_meses), 
+  data = dataset, 
+  method = "REML"
+)
+
+report_overall(mod, "edad_corregida_meses")
+#> [1] "$\\beta$ = 1.95, CI~95%~[0.66, 3.25], $t_{student}$ (226.75) = 2.97, *p* = 0.003"
+
+report_slopes(mod, "edad_corregida_meses")
+#>    Start  End edad_corregida_meses                                                                                expr
+#> 1:   0.0  6.8                  3.4 $\\beta$ = -2.94, CI~95%~[-4.55, -1.34], $t_{student}$ (226.75) = -3.7, *p* = 0.004
+#> 2:   7.3  9.2                  8.2    $\\beta$ = 0.02, CI~95%~[-1.12, 1.17], $t_{student}$ (226.75) = 0.06, *p* = 0.45
+#> 3:   9.7 15.5                 12.6    $\\beta$ = 1.86, CI~95%~[0.61, 3.11], $t_{student}$ (226.75) = 2.93, *p* = 0.009
+#> 4:  16.0 48.0                 32.0  $\\beta$ = -0.02, CI~95%~[-2.05, 2.01], $t_{student}$ (226.75) = 0.07, *p* = 0.646
+
+report_smooth(mod)
+#>                                                   term                                            expr
+#> 1:                             s(edad_corregida_meses) $F_{smooth}$ (5.24, 226.75) = 6.19, *p* < 0.001
+#> 2: s(profesional_id,sexo_paciente,respondedor_vinculo)  $F_{smooth}$ (1.01, 226.75) = 0.1, *p* = 0.244
+```
+
+When analyzing the motor skills domain we found a significant non-linear effect of corrected age on gross motor scores ($F_{smooth}$ (5.24, 226.75) = 6.19, *p* < 0.001), which had an overall positive effect ($\beta$ = 1.95, CI~95%~[0.66, 3.25], $t_{student}$ (226.75) = 2.97, *p* = 0.003), nevertheless, the slope varied across age, whereas between the 0 to 7 age range this has a negative effect on gross motor score $\beta$ = -2.94, CI~95%~[-4.55, -1.34], $t_{student}$ (226.75) = -3.7, *p* = 0.004, but in the 9.7 to 15.5 a positive effect was observed ($\beta$ = 1.86, CI~95%~[0.61, 3.11], $t_{student}$ (226.75) = 2.93, *p* = 0.009), meanwhile in the 7.3 to 9.2 ($\beta$ = 0.02, CI~95%~[-1.12, 1.17]) and 16 to 48 months old ($\beta$ = -0.02, CI~95%~[-2.05, 2.01]) the slope was non significant and practically cero ($t_{student}$ (226.75) = 0.06, *p* = 0.45, and $t_{student}$ (226.75) = 0.07, *p* = 0.646 respectively).
+
+The relationship between developmental domains and corrected age can be seen in [Figure 1](#fig1).
 
@@
 # Conclusion
-
-[...].
-
-# Data Availability Statement
 
@@
 
-**Figure 1**.
+[**Table 1**]{#tab1}.
+Overall baseline and developmental characteristics of the sample and grouped by sex.
+^1^ Data is presented as sample size, and *Mdn* (*IQR*); ^2^ P-values derived from the *Wilcoxon* rank-sum test.
+
+[**Figure 1**]{#fig1}.
 Relationship between corrected age (in months) and developmental domains.
-Regression lines represent fitted values estimated from GAM models.
+Regression lines and shaded area represent predicted values estimated from GAM models and their 95% confidence interval.
 Points and error bars represent the mean and standard error at 5-month age intervals.

### 2022-06-21 manuscript/manuscript.Rmd
ms update

@@
 
-When analyzing the motor skills domain we found a significant non-linear effect of corrected age on gross motor scores ($F_{smooth}$ (5.24, 226.75) = 6.19, *p* < 0.001), which had an overall positive effect ($\beta$ = 1.95, CI~95%~[0.66, 3.25], $t_{student}$ (226.75) = 2.97, *p* = 0.003), nevertheless, the slope varied across age, whereas between the 0 to 7 age range this has a negative effect on gross motor score $\beta$ = -2.94, CI~95%~[-4.55, -1.34], $t_{student}$ (226.75) = -3.7, *p* = 0.004, but in the 9.7 to 15.5 a positive effect was observed ($\beta$ = 1.86, CI~95%~[0.61, 3.11], $t_{student}$ (226.75) = 2.93, *p* = 0.009), meanwhile in the 7.3 to 9.2 ($\beta$ = 0.02, CI~95%~[-1.12, 1.17]) and 16 to 48 months old ($\beta$ = -0.02, CI~95%~[-2.05, 2.01]) the slope was non significant and practically cero ($t_{student}$ (226.75) = 0.06, *p* = 0.45, and $t_{student}$ (226.75) = 0.07, *p* = 0.646 respectively).
+When analysing the motor skills domain, we found a significant non-linear effect of corrected age on gross motor scores ($F_{smooth}$ (5.24, 226.75) = 6.19, *p* < 0.001), which had an overall positive effect ($\beta$ = 1.
+95, CI~95%~[0.66, 3.25], $t_{student}$ (226.75) = 2.97, *p* = 0.003), however, the slope varied as a function of age, with a negative effect on gross motor score $\beta$ = -2.94, CI~95%~[-4.55, -1.
+34], $t_{student}$$ (226.75) = -3.7, *p* = 0.004, but in the range 9.7 to 15.5 a positive effect was observed ($\beta$ = 1.86, CI~95%~[0.61, 3.11], $t_{student}$ (226.75) = 2.93, *p* = 0.009), while in the range 7.3 to 9.2 ($\beta$ = 0.02, CI~95%~[-1.
+12, 1.17]) and 16 to 48 months ($\beta$ = -0.02, CI~95%~[-2.05, 2.01]) the slope was non-significant and virtually zero ($t_{student}$ (226.75) = 0.06, *p* = 0.45, and $t_{student}$ (226.75) = 0.07, *p* = 0.646 respectively).
 

### 2022-06-22 manuscript/manuscript.Rmd
further updates to manuscript
- gams + effect derivatives in fine and gross motor domains

@@
 
+## Communication
+
 ```{r eval=FALSE}
@@
 
-When modelling the effect of chronological age corrected for prematurity on developmental domains, we observed a significant non-linear relationship on communication scores ($F_{smooth}$ (5.2, 224.04) = 13.43, *p* < 0.001), that reflect an overall negative marginal effect ($\beta$ = -2.36, CI~95%~[-3.47, -1.25], $t_{student}$ (224.04) = -4.2, *p* < 0.001), however, this was not true when assessing the direction of the effect in the age range between 1 to 7 ($\beta$ = 0.49, CI~95%~[-0.89, 1.86], $t_{student}$ (224.04) = 0.45, *p* = 0.319), and neither in the 18 to 48 months old group ($\beta$ = 0.45, CI~95%~[-1.32, 2.23], $t_{student}$ (224.04) = 0.42, *p* = 0.593), whereas the effect tend to be positive but non-significant.
+When modelling the effect of chronological age corrected for prematurity on developmental domains, we observed a significant non-linear relationship on communication scores ($F_{smooth}$ (5.2, 224.04) = 13.43, *p* < 0.001), that reflect an overall negative marginal effect ($\beta$ = -2.36, CI~95%~[-3.47, -1.25], $t_{student}$ (224.04) = -4.2, *p* < 0.001), however, this was not true when assessing the direction of the effect in the age range between 0 to 6.8 ($\beta$ = 0.49, CI~95%~[-0.89, 1.86], $t_{student}$ (224.04) = 0.45, *p* = 0.319), and neither in the 18.4 to 48 months old group ($\beta$ = 0.45, CI~95%~[-1.32, 2.23], $t_{student}$ (224.04) = 0.42, *p* = 0.593), whereas the effect tend to be positive but non-significant.
+
+## Motor skills
 
@@
 
-When analysing the motor skills domain, we found a significant non-linear effect of corrected age on gross motor scores ($F_{smooth}$ (5.24, 226.75) = 6.19, *p* < 0.001), which had an overall positive effect ($\beta$ = 1.
-95, CI~95%~[0.66, 3.25], $t_{student}$ (226.75) = 2.97, *p* = 0.003), however, the slope varied as a function of age, with a negative effect on gross motor score $\beta$ = -2.94, CI~95%~[-4.55, -1.
-34], $t_{student}$$ (226.75) = -3.7, *p* = 0.004, but in the range 9.7 to 15.5 a positive effect was observed ($\beta$ = 1.86, CI~95%~[0.61, 3.11], $t_{student}$ (226.75) = 2.93, *p* = 0.009), while in the range 7.3 to 9.2 ($\beta$ = 0.02, CI~95%~[-1.
-12, 1.17]) and 16 to 48 months ($\beta$ = -0.02, CI~95%~[-2.05, 2.01]) the slope was non-significant and virtually zero ($t_{student}$ (226.75) = 0.06, *p* = 0.45, and $t_{student}$ (226.75) = 0.07, *p* = 0.646 respectively).
+When analysing the motor skills domain, we found a significant non-linear effect of corrected age on gross motor scores, $F_{smooth}$ (5.24, 226.75) = 6.19, *p* < 0.001, which had an overall positive effect ($\beta$ = 1.95, CI~95%~[0.66, 3.25], $t_{student}$ (226.75) = 2.97, *p* = 0.003), however, the slope varied as a function of age, with a negative effect in the 0 to 6.8 age range ($\beta$ = -2.94, CI~95%~[-4.55, -1.34], $t_{student}$ (226.75) = -3.7, *p* = 0.004), but in the 9.7 to 15.5 interval, this relationship was inverted ($\beta$ = 1.86, CI~95%~[0.61, 3.11], $t_{student}$ (226.75) = 2.93, *p* = 0.009), although, in the 7.3 to 9.2 ($\beta$ = 0.02, CI~95%~[-1.12, 1.17]) and 16 to 48 age range ($\beta$ = -0.02, CI~95%~[-2.05, 2.01]) the slope was non-significant and virtually zero ($t_{student}$ (226.75) = 0.06, *p* = 0.45, and $t_{student}$ (226.75) = 0.07, *p* = 0.646 respectively).
+
+```{r eval=FALSE}
+mod <- gam(
+  motora_fina_total ~ 
+    s(profesional_id, sexo_paciente, respondedor_vinculo, bs = "re") + 
+    s(edad_corregida_meses), 
+  data = dataset, 
+  method = "REML"
+)
+
+report_overall(mod, "edad_corregida_meses")
+#> [1] "$\\beta$ = 0.04, CI~95%~[-0.45, 0.52], $t_{student}$ (226.77) = 0.14, *p* = 0.886"
+
+report_slopes(mod, "edad_corregida_meses")
+#>    Start  End edad_corregida_meses                                                                                 expr
+#> 1:   0.0 21.8                 10.9    $\\beta$ = 0.03, CI~95%~[-0.54, 0.61], $t_{student}$ (226.77) = 0.04, *p* = 0.434
+#> 2:  22.3 38.3                 30.3 $\\beta$ = -0.79, CI~95%~[-1.45, -0.12], $t_{student}$ (226.77) = -2.34, *p* = 0.022
+#> 3:  38.8 48.0                 43.4  $\\beta$ = -0.88, CI~95%~[-1.97, 0.21], $t_{student}$ (226.77) = -1.62, *p* = 0.113
+
+report_smooth(mod)
+#>                                                   term                                           expr
+#> 1:                             s(edad_corregida_meses) $F_{smooth}$ (2.59, 226.77) = 4.2, *p* = 0.005
+#> 2: s(profesional_id,sexo_paciente,respondedor_vinculo) $F_{smooth}$ (3.63, 226.77) = 0.7, *p* = 0.021
+```
+
+The same was observed when inspecting the effect of corrected age in the fine motor domain scores ($F_{smooth}$ (2.59, 226.77) = 4.2, *p* = 0.005), however it was not possible to estimate a significant overall effect ($\beta$ = 0.04, CI~95%~[-0.45, 0.52], $t_{student}$ (226.77) = 0.14, *p* = 0.886), nevertheless, it was only in the 22.3 to 38.3 age range where a significant negative effect was observed ($\beta$ = -0.79, CI~95%~[-1.45, -0.12], $t_{student}$ (226.77) = -2.34, *p* = 0.022).
 

### 2022-06-24 data-raw/clean_dataset.R
modify data processing
- reorder factor levels from every `*_interpretacion` variable and set them as an ordered factor

@@ -157,6 +157,14 @@ library(data.table)
 
   dataset[, (ind) := lapply(.SD, factor), .SDcols = ind]
 
+  # Reorder interpretation variables
+  ind <- grep(
+    pattern = "interpretacion",
+    names(dataset)
+  )
+
+  dataset[, (ind) := lapply(.SD, factor, levels = c("Debajo De Las Expectativas", "Apenas Sobre Las Expectativas", "Sobre Las Expectativas"), ordered = TRUE), .SDcols = ind]
+
   rm(ind)
 
 # Reorder columns -----------------------------------------------------------------------------

### 2022-06-24 manuscript/misc/figures.R
New figure 2 and update theme of fig 1

@@ -21,8 +21,8 @@ library(see)
 # Socio-individual score ~ corrected age
 
 y_vars <- c("Communication" = "comunicacion_total"
-            , "Gross motor" = "motora_gruesa_total"
             , "Fine motor" = "motora_fina_total"
+            , "Gross motor" = "motora_gruesa_total"
             , "Problem solving" = "resolucion_problemas_total"
           # , "Socio-individual" = "socio_individual_total"
           )
@@ -48,10 +48,77 @@ fig_1 <- ggplot(fig_data, aes(x = edad_corregida_meses, y = value)) +
               na.rm = TRUE) +
   labs(x = "Corrected age (months)", y = "Score",
        caption = "Mean ± SE in 5-unit intervals") +
-  theme_lucid(legend.position = "top")
+  ggpubr::theme_pubr()
 
 if (isTRUE(x = .save_plot)) {
   pdf("manuscript/fig-1.pdf", width = 7, height = 7);
   print(fig_1);
+  dev.off()
+}
+
+# Figure 2 ------------------------------------------------------------------------------------
+
+models <- list("Communication" = NULL,
+               "Gross motor" = NULL,
+               "Fine motor" = NULL,
+               "Problem solving" = NULL)
+
+models[[1]] <- gam(
+  comunicacion_total ~
+    s(profesional_id, sexo_paciente, respondedor_vinculo, bs = "re") +
+    s(edad_corregida_meses),
+  data = dataset,
+  method = "REML"
+  )
+
+models[[2]] <- gam(
+  motora_gruesa_total ~
+    s(profesional_id, sexo_paciente, respondedor_vinculo, bs = "re") +
+    s(edad_corregida_meses),
+  data = dataset,
+  method = "REML"
+  )
+
+models[[3]] <- gam(
+  motora_fina_total ~
+    s(profesional_id, sexo_paciente, respondedor_vinculo, bs = "re") +
+    s(edad_corregida_meses),
+  data = dataset,
+  method = "REML"
+  )
+
+models[[4]] <- gam(
+  resolucion_problemas_total ~
+    s(profesional_id, sexo_paciente, respondedor_vinculo, bs = "re") +
+    s(edad_corregida_meses),
+  data = dataset,
+  method = "REML"
+  )
+
+slopes <- sapply(
+  X = models,
+  FUN = modelbased::estimate_slopes,
[…]

### 2022-06-24 manuscript/manuscript.Rmd
problem solving and fine motor domains updated

@@
 Generalized additive models (GAM) were used to describe linear and non-linear relationships in the form of smooth terms between developmental characteristics, represented through penalized regression splines [@wood2011fast].
-Restricted maximum likelihood (i.e., REML) method was used for the estimation of the smoothing parameters, and thin-plate regression splines for smoothing basis, as they are the optimal smoother of any given basis dimension/rank [@wood2003thin].
-To describe the smooth terms by means of quasi-linear segments, we used approximative effect derivatives to summarise the trend with 95% confidence intervals.
+Restricted maximum likelihood method was used for the estimation of the smoothing parameters, and thin-plate regression splines for smoothing basis, as they are the optimal smoother of any given basis dimension/rank [@wood2003thin].
+To describe the smooth terms by means of quasi-linear segments, we used approximative effect derivatives with 95% confidence intervals.
 
@@
 All the statistical analyses were computed and implemented in the R programming language [@rlanguage].
-GAMs and the corresponding model estimates were computed using the *mgcv* and *modelbased* packages with their corresponding documentation [@wood2017generalized; @dominique2020estimation].
+GAMs and the corresponding model estimates were computed using the *mgcv* and *modelbased* packages, each with well documented functions and methods [@wood2017generalized; @dominique2020estimation].
 Complementary R packages were used for visualization purposes [@hadley2016ggplot2; @daniel2021see].
@@
 
-## Communication
-
 ```{r eval=FALSE}
@@
 #> 2: s(profesional_id,sexo_paciente,respondedor_vinculo)  $F_{smooth}$ (3.76, 224.04) = 0.72, *p* = 0.02
+
+mod <- gam(
+  I(as.integer(comunicacion_interpretacion)) ~ 
+    s(profesional_id, sexo_paciente, respondedor_vinculo, bs = "re") + 
+    s(edad_corregida_meses), 
+  family = ocat(R = 3),
+  data = dataset, 
+  method = "REML"
+)
 ```
@@
 When modelling the effect of chronological age corrected for prematurity on developmental domains, we observed a significant non-linear relationship on communication scores ($F_{smooth}$ (5.2, 224.04) = 13.43, *p* < 0.001), that reflect an overall negative marginal effect ($\beta$ = -2.36, CI~95%~[-3.47, -1.25], $t_{student}$ (224.04) = -4.2, *p* < 0.001), however, this was not true when assessing the direction of the effect in the age range between 0 to 6.8 ($\beta$ = 0.49, CI~95%~[-0.89, 1.86], $t_{student}$ (224.04) = 0.45, *p* = 0.319), and neither in the 18.4 to 48 months old group ($\beta$ = 0.45, CI~95%~[-1.32, 2.23], $t_{student}$ (224.04) = 0.42, *p* = 0.593), whereas the effect tend to be positive but non-significant.
-
-## Motor skills
+The relationship between developmental domains and corrected age can be seen in [Figure 1](#fig1).
 
@@
 
-The same was observed when inspecting the effect of corrected age in the fine motor domain scores ($F_{smooth}$ (2.59, 226.77) = 4.2, *p* = 0.005), however it was not possible to estimate a significant overall effect ($\beta$ = 0.04, CI~95%~[-0.45, 0.52], $t_{student}$ (226.77) = 0.14, *p* = 0.886), nevertheless, it was only in the 22.3 to 38.3 age range where a significant negative effect was observed ($\beta$ = -0.79, CI~95%~[-1.45, -0.12], $t_{student}$ (226.77) = -2.34, *p* = 0.022).
+Despite the fact that a similar non-linear effect was observed when inspecting the influence of corrected age in the fine motor domain scores ($F_{smooth}$ (2.59, 226.77) = 4.2, *p* = 0.005), it was not possible to estimate a significant overall effect different from zero ($\beta$ = 0.04, CI~95%~[-0.45, 0.52], $t_{student}$ (226.77) = 0.14, *p* = 0.886), nevertheless, it was only in the 22.3 to 38.3 age range where a significant negative effect was observed ($\beta$ = -0.79, CI~95%~[-1.45, -0.12], $t_{student}$ (226.77) = -2.34, *p* = 0.022).
 
-The relationship between developmental domains and corrected age can be seen in [Figure 1](#fig1).
+```{r eval=FALSE}
+mod <- gam(
+  resolucion_problemas_total ~ 
+    s(profesional_id, sexo_paciente, respondedor_vinculo, bs = "re") + 
+    s(edad_corregida_meses), 
+  data = dataset, 
+  method = "REML"
+)
+
+report_overall(mod, "edad_corregida_meses")
+#> [1] "$\\beta$ = -1.87, CI~95%~[-3.17, -0.57], $t_{student}$ (227.01) = -2.83, *p* = 0.005"
+
+report_slopes(mod, "edad_corregida_meses")
+#>    Start  End edad_corregida_meses                                                                                 expr
+#> 1:   0.0  5.8                  2.9     $\\beta$ = 2.81, CI~95%~[1.18, 4.44], $t_{student}$ (227.01) = 3.49, *p* = 0.002
+#> 2:   6.3  8.7                  7.5      $\\beta$ = 0.05, CI~95%~[-1.1, 1.2], $t_{student}$ (227.01) = 0.06, *p* = 0.395
+#> 3:   9.2 14.1                 11.6 $\\beta$ = -1.59, CI~95%~[-2.82, -0.37], $t_{student}$ (227.01) = -2.55, *p* = 0.015
+#> 4:  14.5 48.0                 31.3    $\\beta$ = 0.03, CI~95%~[-1.99, 2.04], $t_{student}$ (227.01) = -0.06, *p* = 0.55
+
+report_smooth(mod)
+#>                                                   term                                            expr
+#> 1:                             s(edad_corregida_meses) $F_{smooth}$ (5.66, 227.01) = 3.65, *p* = 0.001
+#> 2: s(profesional_id,sexo_paciente,respondedor_vinculo) $F_{smooth}$ (0.34, 227.01) = 0.03, *p* = 0.339
+```
+
+Problem solving abilities were significantly influenced by corrected age ($F_{smooth}$ (5.66, 227.01) = 3.65, *p* = 0.001), with an overall negative effect ($\beta$ = -1.87, CI~95%~[-3.17, -0.57], $t_{student}$ (227.01) = -2.83, *p* = 0.005), and just like the other domains, this relationship was modified across corrected age.
+In this sense, from the 0 to 5.8 age interval, we found that for every increase in one month in corrected age, we can expect a proportional increase in 2.81 points (CI~95%~[1.18, 4.44], $t_{student}$ (227.01) = 3.49, *p* = 0.002) in the problem solving domain, while in the 9.2 to 14.1 age range this relationship is modified inversely, mainly because in this range we observed that for every increase in one month in corrected age, we could expect a decrease in 1.59 points (CI~95%~[-2.82, -0.37], $t_{student}$ (227.01) = -2.55, *p* = 0.015) in the same domain.
+Other age intervals had no significant slope different from zero (age range [6.3, 8.7], $\beta$ = 0.05, CI~95%~[-1.1, 1.2], $t_{student}$ (227.01) = 0.06, *p* = 0.395; age range [14.5, 48.0], $\beta$ = 0.03, CI~95%~[-1.99, 2.04], $t_{student}$ (227.01) = -0.06, *p* = 0.55).
+
+```{r eval=FALSE}
+mod <- gam(
+  socio_individual_total ~ 
+    s(profesional_id, sexo_paciente, respondedor_vinculo, bs = "re") + 
+    s(edad_corregida_meses), 
+  data = dataset, 
+  method = "REML"
+)
+
+report_overall(mod, "edad_corregida_meses")
+#> [1] "$\\beta$ = -0.09, CI~95%~[-0.26, 0.08], $t_{student}$ (231.58) = -1.08, *p* = 0.282"
+
+report_slopes(mod, "edad_corregida_meses")
+#> [1] "$\\beta$ = -0.09, CI~95%~[-0.26, 0.08], $t_{student}$ (231.58) = -1.08, *p* = 0.282"
+
+report_smooth(mod)
+#>                                                   term                                            expr
+#> 1:                             s(edad_corregida_meses)    $F_{smooth}$ (1, 231.58) = 1.16, *p* = 0.282
+#> 2: s(profesional_id,sexo_paciente,respondedor_vinculo) $F_{smooth}$ (0.42, 231.58) = 0.03, *p* = 0.338
+```
+
+Unlike the others, socio-individual domain was not influenced by corrected age ($F_{smooth}$ (1, 231.58) = 1.16, *p* = 0.282).
 
@@
 Overall baseline and developmental characteristics of the sample and grouped by sex.
-^1^ Data is presented as sample size, and *Mdn* (*IQR*); ^2^ P-values derived from the *Wilcoxon* rank-sum test.
+^1^ Data is presented as sample size, and *Mdn* (*IQR*); ^2^ p-values are computed from the *Wilcoxon* rank-sum test.
 

### 2022-06-25 manuscript/misc/figures.R
Update figures
- Removing fig 2 (now is part of fig 1 as a side panel)
- Add `socio_individual` variable to plots (even though is not significant)

@@ -8,7 +8,7 @@ source("manuscript/misc/setup.R")
 library(ggplot2)
 library(see)
 
-# Figure 1 ------------------------------------------------------------------------------------
+# Figure 1a ------------------------------------------------------------------------------------
 
 # Enviromental variable
 .save_plot <- TRUE
@@ -24,7 +24,7 @@ y_vars <- c("Communication" = "comunicacion_total"
             , "Fine motor" = "motora_fina_total"
             , "Gross motor" = "motora_gruesa_total"
             , "Problem solving" = "resolucion_problemas_total"
-          # , "Socio-individual" = "socio_individual_total"
+            , "Socio-individual" = "socio_individual_total"
           )
 
 fig_data <- melt(
@@ -37,31 +37,25 @@ fig_data[, variable := `levels<-`(variable, names(y_vars))][]
 scaled <- copy(fig_data)
 scaled[, edad_corregida_meses := ASQ3:::near(edad_corregida_meses, 5)][]
 
-fig_1 <- ggplot(fig_data, aes(x = edad_corregida_meses, y = value)) +
-  facet_wrap(~variable, ncol = 2) +
+fig_1a <- ggplot(fig_data, aes(x = edad_corregida_meses, y = value)) +
+  facet_grid(variable ~ ., scale = "free_y") +
   geom_count(alpha = 0.1, na.rm = TRUE) +
   stat_summary(data = scaled, fun.data = mean_se, na.rm = TRUE) +
   geom_smooth(method = "gam",
               formula = y ~ s(x),
-              col = "red4",
+              col = "red3",
               method.args = list(method = "REML"),
               na.rm = TRUE) +
-  labs(x = "Corrected age (months)", y = "Score",
-       caption = "Mean ± SE in 5-unit intervals") +
+  labs(x = "Corrected age (months)", y = "Score") +
   ggpubr::theme_pubr()
 
-if (isTRUE(x = .save_plot)) {
-  pdf("manuscript/fig-1.pdf", width = 7, height = 7);
-  print(fig_1);
-  dev.off()
-}
-
-# Figure 2 ------------------------------------------------------------------------------------
+# Figure 1b ------------------------------------------------------------------------------------
 
 models <- list("Communication" = NULL,
[…]

### 2022-06-25 manuscript/manuscript.Rmd
Improve redaction of statistical methods for GAMs

@@
 Generalized additive models (GAM) were used to describe linear and non-linear relationships in the form of smooth terms between developmental characteristics, represented through penalized regression splines [@wood2011fast].
-Restricted maximum likelihood method was used for the estimation of the smoothing parameters, and thin-plate regression splines for smoothing basis, as they are the optimal smoother of any given basis dimension/rank [@wood2003thin].
-To describe the smooth terms by means of quasi-linear segments, we used approximative effect derivatives with 95% confidence intervals.
+Restricted maximum likelihood method was used for the estimation of the smoothing parameters, and thin-plate regression splines as the smoothing basis, as they are the optimal smoother of any given basis dimension/rank [@wood2003thin].
+To describe the smooth terms by means of quasi-linear segments, we used approximative effect derivatives with 95% confidence intervals (CI~95%~).
 

### 2022-06-25 manuscript/manuscript.Rmd
remove redundant code (not in use)

@@
 #> 2: s(profesional_id,sexo_paciente,respondedor_vinculo)  $F_{smooth}$ (3.76, 224.04) = 0.72, *p* = 0.02
-
-mod <- gam(
-  I(as.integer(comunicacion_interpretacion)) ~ 
-    s(profesional_id, sexo_paciente, respondedor_vinculo, bs = "re") + 
-    s(edad_corregida_meses), 
-  family = ocat(R = 3),
-  data = dataset, 
-  method = "REML"
-)
 ```

### 2022-06-25 manuscript/manuscript.Rmd
improve redaction when refering to figure caption

@@
 When modelling the effect of chronological age corrected for prematurity on developmental domains, we observed a significant non-linear relationship on communication scores ($F_{smooth}$ (5.2, 224.04) = 13.43, *p* < 0.001), that reflect an overall negative marginal effect ($\beta$ = -2.36, CI~95%~[-3.47, -1.25], $t_{student}$ (224.04) = -4.2, *p* < 0.001), however, this was not true when assessing the direction of the effect in the age range between 0 to 6.8 ($\beta$ = 0.49, CI~95%~[-0.89, 1.86], $t_{student}$ (224.04) = 0.45, *p* = 0.319), and neither in the 18.4 to 48 months old group ($\beta$ = 0.45, CI~95%~[-1.32, 2.23], $t_{student}$ (224.04) = 0.42, *p* = 0.593), whereas the effect tend to be positive but non-significant.
-The relationship between developmental domains and corrected age can be seen in [Figure 1](#fig1).
+The relationship between developmental domains, corrected age and their effect derivatives can be seen in [Figure 1](#fig1).
 
@@
 Relationship between corrected age (in months) and developmental domains.
-Regression lines and shaded area represent predicted values estimated from GAM models and their 95% confidence interval.
-Points and error bars represent the mean and standard error at 5-month age intervals.
+Left panel: regression lines and shaded area represent predicted values estimated from GAM models and their CI~95%~, points and error bars represent the mean and standard error at 5-month age intervals.
+Right panel: effect derivatives representing how the effect of corrected age (in months) in developmental domains changes across corrected age.
+Significant areas consider CI~95%~ that did not cross zero.

### 2022-06-27 manuscript/manuscript.Rmd
minor fix
intro added to the ms
- There is a need for further details before the hypothesis (neurodevelopmental variations across age with ASQ-3)

@@
 bibliography: "misc/ref.bib"
-csl: "misc/vancouver.csl"
+csl: "misc/apa.csl"
 link-citations: true
@@
 
-**Title**: [title]
+**Title**: Neurodevelopmental features among infants with congenital hypotonia: An observational cross-sectional trial.
 
@@
 <!---- Antecedentes históricos ---->
-[...].
+Both physical and psychological signs of early childhood development have been shown to be representative and relevant markers for the identification and monitoring of overall growth in early life [@di2016predictive] and can, therefore, be used in the screening of children at risk of developmental delay to support early referral and need further assessment to determine if they are eligible for early intervention services [@guralnick2017early; @bruder2010early].
 
-<!---- Conceptualización del problema ---->
-[...].
+Currently, a plethora of tools have been proposed to assess the developmental continuum of infants.
+In this sense, the Ages and Stages Questionnaire, Third Edition (ASQ-3) has been proposed as a global screening tool, parent/caregivers-oriented, that assesses five domains of development in children aged 0 to 5.5 years [@singh2017ages].
+Current evidence suggests that the ASQ-3 is an accurate, cost-effective yet parent-friendly instrument for screening, monitoring children up to pre-school age, and can help identify and exclude neurodevelopmental impairments in very preterm-born children [@kerstjens2015ages; @ballantyne2016risk; @singh2017ages].
 
-<!---- Hipótesis y objetivos ---->
-[...].
+Hypotonia has been defined both as decreased muscle tone or floppiness, involving a wide range and levels of progression [@harris2008congenital; @gabis2021weak].
+There are multiple forms of neuromuscular, metabolic and genetic conditions associated with hypotonia and it may be a sign of neurodevelopmental delay, that may predispose to cognitive impairment [@riou2009global].
+Given that hypotonia, and the hyperlaxity and motor delay associated with it, may impair the infant's capacity to engage with its surroundings, critical visual cues may be ignored, potentially leading both to an impairment of learning and cognitive development [@harris2008congenital], hence the need to explore the neurodevelopmental attributes of infants with hypotonia.
+
+However, and to the best of our knowledge, there is no robust evidence characterising the observed variation in developmental traits in infants with known hypotonia across age.
+Therefore, our main objective in this study was to describe and model the relationship between sociodemographics, prematurity and neurodevelopmental levels based on ASQ-3 scores in infants with diagnosed hypotonia.
 
@@
 
-In order assess the differences in developmental scores between males and females, the *Wilcoxon* rank-sum test was used, meanwhile the chi-square test ($\chi^2$) was used to evaluate the goodness-of-fit ($\chi^2_{gof}$) and the independence of factors ($\chi^2_{Pearson}$).
+In order assess the differences in developmental scores between males and females, the *Wilcoxon* rank-sum test was used, meanwhile the chi-square test ($\chi^2$) was used to evaluate goodness-of-fit ($\chi^2_{gof}$) and independence of factors ($\chi^2_{Pearson}$).
 
@@
 
-When modelling the effect of chronological age corrected for prematurity on developmental domains, we observed a significant non-linear relationship on communication scores ($F_{smooth}$ (5.2, 224.04) = 13.43, *p* < 0.001), that reflect an overall negative marginal effect ($\beta$ = -2.36, CI~95%~[-3.47, -1.25], $t_{student}$ (224.04) = -4.2, *p* < 0.001), however, this was not true when assessing the direction of the effect in the age range between 0 to 6.8 ($\beta$ = 0.49, CI~95%~[-0.89, 1.86], $t_{student}$ (224.04) = 0.45, *p* = 0.319), and neither in the 18.4 to 48 months old group ($\beta$ = 0.45, CI~95%~[-1.32, 2.23], $t_{student}$ (224.04) = 0.42, *p* = 0.593), whereas the effect tend to be positive but non-significant.
+When modelling the effect of chronological age on developmental domains, corrected for prematurity, we observed a significant non-linear relationship on communication scores ($F_{smooth}$ (5.2, 224.04) = 13.43, *p* < 0.001), that reflect an overall negative marginal effect ($\beta$ = -2.36, CI~95%~[-3.47, -1.25], $t_{student}$ (224.04) = -4.2, *p* < 0.001), however, this was not true when assessing the direction of the effect in the age range between 0 to 6.8 ($\beta$ = 0.49, CI~95%~[-0.89, 1.86], $t_{student}$ (224.04) = 0.45, *p* = 0.319), and neither in the 18.4 to 48 months old group ($\beta$ = 0.45, CI~95%~[-1.32, 2.23], $t_{student}$ (224.04) = 0.42, *p* = 0.593), whereas the effect tend to be positive but non-significant.
 The relationship between developmental domains, corrected age and their effect derivatives can be seen in [Figure 1](#fig1).

### 2022-06-27 manuscript/manuscript.Rmd
introduction v1

@@
 
-**Title**: Neurodevelopmental features among infants with congenital hypotonia: An observational cross-sectional trial.
+**Title**: Neurodevelopmental features among infants with benign congenital hypotonia in high southern latitude: An observational cross-sectional trial.
 
@@
 Currently, a plethora of tools have been proposed to assess the developmental continuum of infants.
-In this sense, the Ages and Stages Questionnaire, Third Edition (ASQ-3) has been proposed as a global screening tool, parent/caregivers-oriented, that assesses five domains of development in children aged 0 to 5.5 years [@singh2017ages].
+In this sense, the Ages and Stages Questionnaire, Third Edition (ASQ-3) has been proposed as a global screening tool, parent/caregivers-oriented, that assesses five domains of development in children aged from 0 to 5.5 years old [@singh2017ages].
 Current evidence suggests that the ASQ-3 is an accurate, cost-effective yet parent-friendly instrument for screening, monitoring children up to pre-school age, and can help identify and exclude neurodevelopmental impairments in very preterm-born children [@kerstjens2015ages; @ballantyne2016risk; @singh2017ages].
@@
 
-However, and to the best of our knowledge, there is no robust evidence characterising the observed variation in developmental traits in infants with known hypotonia across age.
-Therefore, our main objective in this study was to describe and model the relationship between sociodemographics, prematurity and neurodevelopmental levels based on ASQ-3 scores in infants with diagnosed hypotonia.
+Accordingly, benign congenital hypotonia (BCH) is usually considered an exclusion diagnosis, and is usually made in the absence of other signs and symptoms, after every other evaluative resources have been exhausted [@leyenaar2005schematic; @gabis2021weak], however this is not true for at least one author that suggest that BCH can not be considered a diagnosis as such [@thompson2002benign].
+BCH is considered a non-progressive neuromuscular disorder that does not progress but tends to improve with time and early intervention.
+
+Nevertheless, and to the best of our knowledge, there is no robust evidence characterising the observed variation in developmental traits in infants with known BCH across age.
+Therefore, our main objective in this study was to describe and model the relationship between sociodemographics, prematurity and neurodevelopmental levels based on ASQ-3 scores in infants with diagnosed BCH.
 
@@
 
-### Instrument 1
+### Ages and Stages Questionnaire
 
@@
 #> [1] "$\\chi^2_{gof}$ (1) = 9.04, *p* = 0.003, $C_{Pearson}$ = 0.19, CI~0.95%~[0.09, 1]"
-
 ```

### 2022-06-28 manuscript/manuscript.Rmd
discussion and title update

@@
 
-**Title**: Neurodevelopmental features among infants with benign congenital hypotonia in high southern latitude: An observational cross-sectional trial.
+**Title**: Neurodevelopmental features among infants with benign congenital hypotonia in high southern latitudes: An observational cross-sectional trial.
 
@@
 
+```{r eval=FALSE}
+y_vars <- c("Communication" = "comunicacion_total"
+            , "Fine motor" = "motora_fina_total"
+            , "Gross motor" = "motora_gruesa_total"
+            , "Problem solving" = "resolucion_problemas_total"
+            , "Socio-individual" = "socio_individual_total")
+
+many_models <- sapply(y_vars, function(i) {
+  form <- as.formula(
+    paste0(i, " ~ ", 
+           "s(profesional_id, sexo_paciente, respondedor_vinculo, bs = \"re\") +",
+           "s(semanas_prematurez, k = 9)")
+  )
+  model <- gam(
+    form, 
+    data = dataset, 
+    method = "REML"
+  )
+  report_smooth(model)
+}, simplify = FALSE)
+
+rbindlist(many_models, idcol = "var")[term %like% "semanas_prematurez"]
+#>                  var                                                term                                            expr
+#>  1:    Communication                               s(semanas_prematurez)    $F_{smooth}$ (1, 228.25) = 0.13, *p* = 0.715
+#>  2:       Fine motor                               s(semanas_prematurez)       $F_{smooth}$ (1, 230.23) = 0, *p* = 0.987
+#>  3:      Gross motor                               s(semanas_prematurez)    $F_{smooth}$ (1, 228.31) = 0.85, *p* = 0.357
+#>  4:  Problem solving                               s(semanas_prematurez)       $F_{smooth}$ (1, 232) = 1.12, *p* = 0.292
+#>  5: Socio-individual                               s(semanas_prematurez)     $F_{smooth}$ (1, 231.99) = 2.3, *p* = 0.131
+```
+
 Unlike the others, socio-individual domain was not influenced by corrected age ($F_{smooth}$ (1, 231.58) = 1.16, *p* = 0.282).
+Accordingly, prematurity (measured in weeks) was not associated with any developmental domain within ASQ-3 assessment (significance for smooth: communication, *p* = 0.715; fine motor, *p* = 0.987; gross motor, *p* = 0.357; problem-solving, *p* = 0.292; personal-social, *p* = 0.131).
 
@@
 
-[...].
+Our study aimed to describe and model the relationship between sociodemographic data, prematurity and neurodevelopmental levels based on ASQ-3 scores in infants diagnosed with BCH.
+Our main findings suggest a non-linear effect of age, corrected for prematurity, with a marked decrease in scores for all neurodevelopmental traits at different age frames, even after adjusting for caregiver relationship, gender and inter-rater influence.
+However in the personal-individual domain, there was no variation observed across corrected age.
 
-[...].
+This could be associated with the described motor impairments of hypotonia in the early stages of life, which could compromise the infant's ability to explore and interact with his or her environment [@harris2008congenital; @gabis2021weak].
+A reflection of the aforementioned would be expressed in an altered development of gross motor function in the first months of life, with a consequent limitation in fine motor skills later on, which would have a subsequent negative impact on the communicative competence of infants, secondary to a reduced interaction with their environment and peers.
+The results found in our study are congruent with what other authors have discussed in relation to the role of gross and fine motor skills in language development [@gonzalez2019gross].
 
-[...].
+It is worth noting that other context-mediated social factors may also have influenced our results, mainly due to the role that other variables would also play in the neurodevelopment of our study sample, which could have an impact on many of the developmental traits assessed here, such as intrauterine growth restrictions, maternal depression, institutionalisation, exposure to social violence, maternal education and breastfeeding [@walker2011inequality].
+All together, these represent the main limitations in our study design, which need to be addressed in future research exploring the variations observed in different developmental traits in hypotonic infants.
+However, our study sheds light to an underexplored aspect of congenital myopathies, with robust statistical methods that made it possible to model the data and capture the complex relationships seen early in life.
 

### 2022-06-28 manuscript/misc/figures.R
-

@@ -24,8 +24,7 @@ y_vars <- c("Communication" = "comunicacion_total"
             , "Fine motor" = "motora_fina_total"
             , "Gross motor" = "motora_gruesa_total"
             , "Problem solving" = "resolucion_problemas_total"
-            , "Socio-individual" = "socio_individual_total"
-          )
+            , "Socio-individual" = "socio_individual_total")
 
 fig_data <- melt(
   data = dataset,

### 2022-07-01 manuscript/manuscript.Rmd
ms update
- authors and afiliations
- abstract (partial)
- methods skeleton
- statistical analysis
- dynamic parameters in results
- discussion and conclusion
- conflict of interest

@@
 
-**Authors**: Primer autor,^1^ Segundo autor,^2^ Tercer autor,^3^ Cuarto autor,^4^ Quinto autor,^5^ Sexto autor.^6^ \*
+**Authors**: Patricio Barría,^1^ Matías Castillo-Aguilar,^2,3^ ...,^4^ Cristian Núñez-Espinoza,^3,5^ \*
 
 ::: {custom-style="Afiliation"}
-^1^ Afiliación, lugar.
+^1^ Rehabilitation Center, Punta Arenas, Chile.
 
-^2^ Afiliación, lugar.
+^2^ Kinesiology Department, Magallanes University, Punta Arenas, Chile.
 
-^3^ Afiliación, lugar.
+^3^ Chilean Austral Integrative Neurophysiology Group ([NIM-ACh](https://nimach.org)), Centro Asistencial de Docencia e Investigación (CADI-UMAG), Punta Arenas Chile
 
@@
 
-^5^ Afiliación, lugar.
-
-^6^ Afiliación, lugar.
+^5^ School of Medicine, Magallanes University, Punta Arenas, Chile.
 :::
@@
 
-Autor, Afiliación, lugar. e-mail: autor\@mail.com. 
-Dirección: XXXX. Teléfono: XXXX.
+Cristian Núñez-Espinosa, School of Medicine, Magallanes University, Punta Arenas, Chile.
+Centro Asistencial de Docencia e Investigación CADI-UMAG, Chile.
+e-mail: [cristian.nunez\@umag.cl](mailto:cristian.nunez@umag.cl).
+Address: Avenida Bulnes 01855, Box 113-D.
+Phone: +56 61 2201411
 
@@
 
-**Objective**: [...]. **Material and methods**: [...]. **Results**: [...]. **Conclusion**: [...].
+**Objective**: To describe and model the relationship between sociodemographics, prematurity and neurodevelopmental levels based on the *Ages and Stages Questionnaire* scores in infants with diagnosed BCH.
+**Material and methods**: [...].
+**Results**: [...].
+**Conclusion**: After adjusting for possible confounders, we found between-subjects fluctuations in neurodevelopmental traits across age in hypotonic infants in the form of non-linear and domain-specific variations.
+Further research is warranted to determine how these findings apply in the presence of other context-mediated social factors and populations.
 
@@
 
-### Ages and Stages Questionnaire
+### Sociodemographic Data
 
@@
 
-### Instrument 2
+### Child Development
 
@@
 Restricted maximum likelihood method was used for the estimation of the smoothing parameters, and thin-plate regression splines as the smoothing basis, as they are the optimal smoother of any given basis dimension/rank [@wood2003thin].
-To describe the smooth terms by means of quasi-linear segments, we used approximative effect derivatives with 95% confidence intervals (CI~95%~).
-
-To account for any source of variability coming from subject's sex, evaluators and the type of relationship between the infants with the respondents, we incorporate them as random effects in the fitted models in the form of penalized parametric terms [@wood2016smoothing].
+In the final models, infants' sex, clinician and infants' relationship with caregivers were added as random effects in the form of penalized parametric terms to account for the variability arising from these variables in the fixed effects analysed [@wood2016smoothing].
+To describe the smooth terms by means of quasi-linear segments, we used approximative derivatives with 95% confidence intervals (CI~95%~).
 
@@
 
-```{r eval=FALSE}
+```{r}
 # Sample
@@
 
-From a total of 234 subjects with congenital hypotonia, 94 (40.2%) were females and 140 (59.8%) males ($\chi^2_{gof}$ (1) = 9.04, *p* = 0.003, $C_{Pearson}$ = 0.19, CI~0.95%~[0.09, 1]).
+From a total of `r n_sample` subjects with congenital hypotonia, `r n_sex["F", stat]` were females and `r n_sex["M", stat]` males (`r prop_sex`).
 The developmental characteristics of the sample can be seen in [Table 1](#tab1).
 
-```{r eval=FALSE}
+```{r}
+set.seed(123)
 mod <- gam(
@@
 
-report_overall(mod, "edad_corregida_meses")
+overall <- report_overall(mod, "edad_corregida_meses")
 #> [1] "$\\beta$ = -2.36, CI~95%~[-3.47, -1.25], $t_{student}$ (224.04) = -4.2, *p* < 0.001"
 
-report_slopes(mod, "edad_corregida_meses")
+slopes <- report_slopes(mod, "edad_corregida_meses")
 #>    Start  End edad_corregida_meses                                                                                 expr
@@
 
-report_smooth(mod)
+smooth <- report_smooth(mod)
 #>                                                   term                                            expr
@@
 
-When modelling the effect of chronological age on developmental domains, corrected for prematurity, we observed a significant non-linear relationship on communication scores ($F_{smooth}$ (5.2, 224.04) = 13.43, *p* < 0.001), that reflect an overall negative marginal effect ($\beta$ = -2.36, CI~95%~[-3.47, -1.25], $t_{student}$ (224.04) = -4.2, *p* < 0.001), however, this was not true when assessing the direction of the effect in the age range between 0 to 6.8 ($\beta$ = 0.49, CI~95%~[-0.89, 1.86], $t_{student}$ (224.04) = 0.45, *p* = 0.319), and neither in the 18.4 to 48 months old group ($\beta$ = 0.45, CI~95%~[-1.32, 2.23], $t_{student}$ (224.04) = 0.42, *p* = 0.593), whereas the effect tend to be positive but non-significant.
+When modelling the effect of chronological age on developmental domains, corrected for prematurity, we observed a significant non-linear relationship on communication scores (`r smooth["s(edad_corregida_meses)", expr]`), that reflect an overall negative marginal effect (`r overall`), however, this was not true when assessing the direction of the effect in the age range between 0 to 6.8 (`r slopes[1, expr]`), neither in the 18.4 to 48 months old group (`r slopes[3, expr]`), whereas the effect tend to be positive but non-significant.
 The relationship between developmental domains, corrected age and their effect derivatives can be seen in [Figure 1](#fig1).
 
-```{r eval=FALSE}
+```{r}
+set.seed(123)
 mod <- gam(
@@
 
-report_overall(mod, "edad_corregida_meses")
+overall <- report_overall(mod, "edad_corregida_meses")
 #> [1] "$\\beta$ = 1.95, CI~95%~[0.66, 3.25], $t_{student}$ (226.75) = 2.97, *p* = 0.003"
 
-report_slopes(mod, "edad_corregida_meses")
+slopes <- report_slopes(mod, "edad_corregida_meses")
 #>    Start  End edad_corregida_meses                                                                                expr
@@
 
-When analysing the motor skills domain, we found a significant non-linear effect of corrected age on gross motor scores, $F_{smooth}$ (5.24, 226.75) = 6.19, *p* < 0.001, which had an overall positive effect ($\beta$ = 1.95, CI~95%~[0.66, 3.25], $t_{student}$ (226.75) = 2.97, *p* = 0.003), however, the slope varied as a function of age, with a negative effect in the 0 to 6.8 age range ($\beta$ = -2.94, CI~95%~[-4.55, -1.34], $t_{student}$ (226.75) = -3.7, *p* = 0.004), but in the 9.7 to 15.5 interval, this relationship was inverted ($\beta$ = 1.86, CI~95%~[0.61, 3.11], $t_{student}$ (226.75) = 2.93, *p* = 0.009), although, in the 7.3 to 9.2 ($\beta$ = 0.02, CI~95%~[-1.12, 1.17]) and 16 to 48 age range ($\beta$ = -0.02, CI~95%~[-2.05, 2.01]) the slope was non-significant and virtually zero ($t_{student}$ (226.75) = 0.06, *p* = 0.45, and $t_{student}$ (226.75) = 0.07, *p* = 0.646 respectively).
+When analysing the motor skills domain, we found a significant non-linear effect of corrected age on gross motor scores, `r smooth["s(edad_corregida_meses)", expr]`, which had an overall positive effect (`r overall`), however, the slope varied as a function of age, with a negative effect in the 0 to 6.8 age range (`r slopes[1, expr]`), but in the 9.7 to 15.5 interval, this relationship was inverted (`r slopes[3, expr]`), however, in the rest of the age range the slope was non-significant and virtually zero (Age~[7.3, 9.2]~, `r slopes[2, expr]`; Age~[16, 48]~, `r slopes[4, expr]`).
 
-```{r eval=FALSE}
+```{r}
+set.seed(123)
 mod <- gam(
@@
 
-report_overall(mod, "edad_corregida_meses")
+overall <- report_overall(mod, "edad_corregida_meses")
 #> [1] "$\\beta$ = 0.04, CI~95%~[-0.45, 0.52], $t_{student}$ (226.77) = 0.14, *p* = 0.886"
 
-report_slopes(mod, "edad_corregida_meses")
+slopes <- report_slopes(mod, "edad_corregida_meses")
 #>    Start  End edad_corregida_meses                                                                                 expr
@@
 
-report_smooth(mod)
+smooth <- report_smooth(mod)
 #>                                                   term                                           expr
@@
 
-Despite the fact that a similar non-linear effect was observed when inspecting the influence of corrected age in the fine motor domain scores ($F_{smooth}$ (2.59, 226.77) = 4.2, *p* = 0.005), it was not possible to estimate a significant overall effect different from zero ($\beta$ = 0.04, CI~95%~[-0.45, 0.52], $t_{student}$ (226.77) = 0.14, *p* = 0.886), nevertheless, it was only in the 22.3 to 38.3 age range where a significant negative effect was observed ($\beta$ = -0.79, CI~95%~[-1.45, -0.12], $t_{student}$ (226.77) = -2.34, *p* = 0.022).
+Despite the fact that a similar non-linear effect was observed when inspecting the influence of corrected age in the fine motor domain scores (`r smooth["s(edad_corregida_meses)", expr]`), it was not possible to estimate a significant overall effect different from zero (`r overall`), nevertheless, it was only in the 22.3 to 38.3 age range where a significant and negative effect was observed (`r slopes[2, expr]`).
 
-```{r eval=FALSE}
+```{r}
+set.seed(123)
 mod <- gam(
@@
 
-report_overall(mod, "edad_corregida_meses")
+overall <- report_overall(mod, "edad_corregida_meses")
 #> [1] "$\\beta$ = -1.87, CI~95%~[-3.17, -0.57], $t_{student}$ (227.01) = -2.83, *p* = 0.005"
 
-report_slopes(mod, "edad_corregida_meses")
+slopes <- report_slopes(mod, "edad_corregida_meses")
 #>    Start  End edad_corregida_meses                                                                                 expr
@@
 
-Problem solving abilities were significantly influenced by corrected age ($F_{smooth}$ (5.66, 227.01) = 3.65, *p* = 0.001), with an overall negative effect ($\beta$ = -1.87, CI~95%~[-3.17, -0.57], $t_{student}$ (227.01) = -2.83, *p* = 0.005), and just like the other domains, this relationship was modified across corrected age.
-In this sense, from the 0 to 5.8 age interval, we found that for every increase in one month in corrected age, we can expect a proportional increase in 2.81 points (CI~95%~[1.18, 4.44], $t_{student}$ (227.01) = 3.49, *p* = 0.002) in the problem solving domain, while in the 9.2 to 14.1 age range this relationship is modified inversely, mainly because in this range we observed that for every increase in one month in corrected age, we could expect a decrease in 1.59 points (CI~95%~[-2.82, -0.37], $t_{student}$ (227.01) = -2.55, *p* = 0.015) in the same domain.
-Other age intervals had no significant slope different from zero (age range [6.3, 8.7], $\beta$ = 0.05, CI~95%~[-1.1, 1.2], $t_{student}$ (227.01) = 0.06, *p* = 0.395; age range [14.5, 48.0], $\beta$ = 0.03, CI~95%~[-1.99, 2.04], $t_{student}$ (227.01) = -0.06, *p* = 0.55).
+Problem solving abilities were significantly influenced by corrected age (`r smooth["s(edad_corregida_meses)", expr]`), with an overall negative effect (`r overall`), and just like the other domains, this relationship was modified across corrected age.
+In this sense, from the 0 to 5.8 age interval, we found that for every increase in one month in corrected age, we can expect a proportional increase in 2.81 points (`r slopes[1, expr]`) in the problem solving domain, while in the age range 9.2 to 14.1 the relationship changes inversely, mainly because in this age range we observe that for every one-month increase in the corrected age, a decrease of 1.59 points could be expected in the same domain (`r slopes[3, expr]`).
+The other age intervals did not have a slope that deviated significantly from zero (Age~[6.3, 8.7]~, `r slopes[2, expr]`; Age~[14.5, 48.0]~, `r slopes[4, expr]`).
 
-```{r eval=FALSE}
+```{r}
+set.seed(123)
 mod <- gam(
@@
 
-report_overall(mod, "edad_corregida_meses")
-#> [1] "$\\beta$ = -0.09, CI~95%~[-0.26, 0.08], $t_{student}$ (231.58) = -1.08, *p* = 0.282"
-
-report_slopes(mod, "edad_corregida_meses")
-#> [1] "$\\beta$ = -0.09, CI~95%~[-0.26, 0.08], $t_{student}$ (231.58) = -1.08, *p* = 0.282"
-
-report_smooth(mod)
+smooth <- report_smooth(mod)
 #>                                                   term                                            expr
@@
 
-```{r eval=FALSE}
-y_vars <- c("Communication" = "comunicacion_total"
-            , "Fine motor" = "motora_fina_total"
-            , "Gross motor" = "motora_gruesa_total"
-            , "Problem solving" = "resolucion_problemas_total"
-            , "Socio-individual" = "socio_individual_total")
+```{r}
+y_vars <- c("communication" = "comunicacion_total"
+            , "fine motor" = "motora_fina_total"
+            , "gross motor" = "motora_gruesa_total"
+            , "problem solving" = "resolucion_problemas_total"
+            , "personal-individual" = "socio_individual_total")
 
@@
 
-rbindlist(many_models, idcol = "var")[term %like% "semanas_prematurez"]
-#>                  var                                                term                                            expr
-#>  1:    Communication                               s(semanas_prematurez)    $F_{smooth}$ (1, 228.25) = 0.13, *p* = 0.715
-#>  2:       Fine motor                               s(semanas_prematurez)       $F_{smooth}$ (1, 230.23) = 0, *p* = 0.987
-#>  3:      Gross motor                               s(semanas_prematurez)    $F_{smooth}$ (1, 228.31) = 0.85, *p* = 0.357
-#>  4:  Problem solving                               s(semanas_prematurez)       $F_{smooth}$ (1, 232) = 1.12, *p* = 0.292
-#>  5: Socio-individual                               s(semanas_prematurez)     $F_{smooth}$ (1, 231.99) = 2.3, *p* = 0.131
+prematurity_smooths <- rbindlist(many_models, idcol = "var"
+          )[i = term %like% "semanas_prematurez",
+            j = paste(
+              var, 
+              regmatches(expr, regexec("\\*p\\* = 0.[[:digit:]]+", expr)), 
+              sep = ", ", 
+              collapse = "; "
+            )]
+#> [1] "communication, *p* = 0.715; fine motor, *p* = 0.987; gross motor, *p* = 0.357; problem solving, *p* = 0.292; personal-individual, *p* = 0.131"
 ```
 
-Unlike the others, socio-individual domain was not influenced by corrected age ($F_{smooth}$ (1, 231.58) = 1.16, *p* = 0.282).
-Accordingly, prematurity (measured in weeks) was not associated with any developmental domain within ASQ-3 assessment (significance for smooth: communication, *p* = 0.715; fine motor, *p* = 0.987; gross motor, *p* = 0.357; problem-solving, *p* = 0.292; personal-social, *p* = 0.131).
+Unlike the others, socio-individual domain was not influenced by corrected age (`r smooth["s(edad_corregida_meses)", expr]`).
+Accordingly, prematurity (measured in weeks) was not associated with any developmental domain within ASQ-3 assessment (significance for smooth terms: `r prematurity_smooths`).
 
@@
 Our study aimed to describe and model the relationship between sociodemographic data, prematurity and neurodevelopmental levels based on ASQ-3 scores in infants diagnosed with BCH.
-Our main findings suggest a non-linear effect of age, corrected for prematurity, with a marked decrease in scores for all neurodevelopmental traits at different age frames, even after adjusting for caregiver relationship, gender and inter-rater influence.
+Our main findings suggest a non-linear effect of age, corrected for prematurity, with a marked decrease in scores for all neurodevelopmental traits at different age frames, even after adjusting for caregiver relationship, sex and inter-rater influence.
 However in the personal-individual domain, there was no variation observed across corrected age.
@@
 All together, these represent the main limitations in our study design, which need to be addressed in future research exploring the variations observed in different developmental traits in hypotonic infants.
-However, our study sheds light to an underexplored aspect of congenital myopathies, with robust statistical methods that made it possible to model the data and capture the complex relationships seen early in life.
+However, our study sheds light to an underexplored aspect of congenital myopathies, with robust statistical methods that made it possible to model and capture the complex relationships seen early in life.
 
@@
 
-[...].
+The present study shows that the marked variations observed in neurodevelopmental traits are present across age in hypotonic infants, mainly in the form of non-linear and domain-specific variations, even after adjusting for the effect that caregiver relationship, gender and evaluators might exert.
+Moreover, we show that the observed variations in developmental domains are not solely attributable to prematurity, where age corrected for prematurity best explained the observed variability in neurodevelopment.
+Further research is warranted to determine how these findings apply when controlling for context-mediated social factors and in other populations.
 
@@
 
-[...].
+The authors declare that the research was conducted in the absence of any commercial or financial relationships that could be construed as a potential conflict of interest.
 

### 2022-07-01 manuscript/manuscript.Rmd
adding initials for domains in figure and text

@@
 
-**Objective**: To describe and model the relationship between sociodemographics, prematurity and neurodevelopmental levels based on the *Ages and Stages Questionnaire* scores in infants with diagnosed BCH.
+**Objective**: To describe and model the relationship between sociodemographics, prematurity and neurodevelopmental levels based on the *Ages and Stages Questionnaire* scores in infants diagnosed with benign congenital hypotonia (BCH).
 **Material and methods**: [...].
@@
 
-[...].
+```{r load-packages}
+source(file = "manuscript/misc/setup.R")
+```
 
-## Instruments
+```{r}
+# Sample
+n_sample <- dataset[, .N]
+#> 234
 
-### Sociodemographic Data
+# Sample by sex
+n_sex <- dataset[j = .(stat = paste0(.N, " (", round(.N*100/n_sample, 1), "%)")), 
+                 keyby = .(sex = sexo_paciente)]
+#>    sex        stat
+#> 1:   F  94 (40.2%)
+#> 2:   M 140 (59.8%)
+```
+
+A total of `r n_sample` subjects with BCH (females, `r n_sex["F", stat]`; males, `r n_sex["M", stat]`) were assessed.
+
+## Measures
+
+### The Ages and Stages Questionnaire, third edition (ASQ-3)
+
+The ASQ-3 is a parent reported initial level developmental screening instrument consisting of 21 intervals, each with 30 items in five areas: i) communication (CM), ii) gross motor (GM), iii) fine motor (FM), iv) problem-solving (CG), and v) personal-social (PS).
+The ASQ is cost-effective and widely used in the United States and other countries.
+The ASQ has been translated into several languages, and the number of international studies on its psychometric properties with diverse cultural environments is increasing.
+It has excellent psychometric properties, test-retest reliability of 92%, sensitivity of 87.4% and specificity of 95.7%.
+Its validity has been examined across different cultures and communities across the world.
+
+## Procedures
+
+### Collection of demographic data
 
@@
 
-### Child Development
-
-[...].
-
-## Procedures
-
-[...].
+### Assessment with ASQ-3
 
@@
 
-```{r load-packages}
-source(file = "manuscript/misc/setup.R")
-```
-
 ```{r}
-# Sample
-n_sample <- dataset[, .N]
-#> 234
-
-# Sample by sex
-n_sex <- dataset[j = .(stat = paste0(.N, " (", round(.N*100/n_sample, 1), "%)")), 
-                 keyby = .(sex = sexo_paciente)]
-#>    sex        stat
-#> 1:   F  94 (40.2%)
-#> 2:   M 140 (59.8%)
-
 # Prop test on sex
@@
 
-When modelling the effect of chronological age on developmental domains, corrected for prematurity, we observed a significant non-linear relationship on communication scores (`r smooth["s(edad_corregida_meses)", expr]`), that reflect an overall negative marginal effect (`r overall`), however, this was not true when assessing the direction of the effect in the age range between 0 to 6.8 (`r slopes[1, expr]`), neither in the 18.4 to 48 months old group (`r slopes[3, expr]`), whereas the effect tend to be positive but non-significant.
+When modelling the effect of chronological age on developmental domains, corrected for prematurity, we observed a significant non-linear relationship on CM scores (`r smooth["s(edad_corregida_meses)", expr]`), that reflect an overall negative marginal effect (`r overall`), however, this was not true when assessing the direction of the effect in the age range between 0 to 6.8 (`r slopes[1, expr]`), neither in the 18.4 to 48 months old group (`r slopes[3, expr]`), whereas the effect tend to be positive but non-significant.
 The relationship between developmental domains, corrected age and their effect derivatives can be seen in [Figure 1](#fig1).
@@
 
-When analysing the motor skills domain, we found a significant non-linear effect of corrected age on gross motor scores, `r smooth["s(edad_corregida_meses)", expr]`, which had an overall positive effect (`r overall`), however, the slope varied as a function of age, with a negative effect in the 0 to 6.8 age range (`r slopes[1, expr]`), but in the 9.7 to 15.5 interval, this relationship was inverted (`r slopes[3, expr]`), however, in the rest of the age range the slope was non-significant and virtually zero (Age~[7.3, 9.2]~, `r slopes[2, expr]`; Age~[16, 48]~, `r slopes[4, expr]`).
+When analysing the motor skills domain, we found a significant non-linear effect of corrected age on GM scores, `r smooth["s(edad_corregida_meses)", expr]`, which had an overall positive effect (`r overall`), however, the slope varied as a function of age, with a negative effect in the 0 to 6.8 age range (`r slopes[1, expr]`), but in the 9.7 to 15.5 interval, this relationship was inverted (`r slopes[3, expr]`), however, in the rest of the age range the slope was non-significant and virtually zero (Age~[7.3, 9.2]~, `r slopes[2, expr]`; Age~[16, 48]~, `r slopes[4, expr]`).
 
@@
 
-Despite the fact that a similar non-linear effect was observed when inspecting the influence of corrected age in the fine motor domain scores (`r smooth["s(edad_corregida_meses)", expr]`), it was not possible to estimate a significant overall effect different from zero (`r overall`), nevertheless, it was only in the 22.3 to 38.3 age range where a significant and negative effect was observed (`r slopes[2, expr]`).
+Despite the fact that a similar non-linear effect was observed when inspecting the influence of corrected age in the FM domain scores (`r smooth["s(edad_corregida_meses)", expr]`), it was not possible to estimate a significant overall effect different from zero (`r overall`), nevertheless, it was only in the 22.3 to 38.3 age range where a significant and negative effect was observed (`r slopes[2, expr]`).
 
@@
 
-Problem solving abilities were significantly influenced by corrected age (`r smooth["s(edad_corregida_meses)", expr]`), with an overall negative effect (`r overall`), and just like the other domains, this relationship was modified across corrected age.
-In this sense, from the 0 to 5.8 age interval, we found that for every increase in one month in corrected age, we can expect a proportional increase in 2.81 points (`r slopes[1, expr]`) in the problem solving domain, while in the age range 9.2 to 14.1 the relationship changes inversely, mainly because in this age range we observe that for every one-month increase in the corrected age, a decrease of 1.59 points could be expected in the same domain (`r slopes[3, expr]`).
+CG abilities were significantly influenced by corrected age (`r smooth["s(edad_corregida_meses)", expr]`), with an overall negative effect (`r overall`), and just like the other domains, this relationship was modified across corrected age.
+In this sense, from the 0 to 5.8 age interval, we found that for every increase in one month in corrected age, we can expect a proportional increase in 2.81 points (`r slopes[1, expr]`) in the CG domain, while in the age range 9.2 to 14.1 the relationship changes inversely, mainly because in this age range we observe that for every one-month increase in the corrected age, a decrease of 1.59 points could be expected in the same domain (`r slopes[3, expr]`).
 The other age intervals did not have a slope that deviated significantly from zero (Age~[6.3, 8.7]~, `r slopes[2, expr]`; Age~[14.5, 48.0]~, `r slopes[4, expr]`).
@@
 ```{r}
-y_vars <- c("communication" = "comunicacion_total"
-            , "fine motor" = "motora_fina_total"
-            , "gross motor" = "motora_gruesa_total"
-            , "problem solving" = "resolucion_problemas_total"
+y_vars <- c("CM" = "comunicacion_total"
+            , "FM" = "motora_fina_total"
+            , "GM" = "motora_gruesa_total"
+            , "CG" = "resolucion_problemas_total"
             , "personal-individual" = "socio_individual_total")
@@
             )]
-#> [1] "communication, *p* = 0.715; fine motor, *p* = 0.987; gross motor, *p* = 0.357; problem solving, *p* = 0.292; personal-individual, *p* = 0.131"
+#> [1] "CM, *p* = 0.715; FM, *p* = 0.987; GM, *p* = 0.357; CG, *p* = 0.292; personal-individual, *p* = 0.131"
 ```
 
-Unlike the others, socio-individual domain was not influenced by corrected age (`r smooth["s(edad_corregida_meses)", expr]`).
+Unlike the others, PS domain was not influenced by corrected age (`r smooth["s(edad_corregida_meses)", expr]`).
 Accordingly, prematurity (measured in weeks) was not associated with any developmental domain within ASQ-3 assessment (significance for smooth terms: `r prematurity_smooths`).
@@
 This could be associated with the described motor impairments of hypotonia in the early stages of life, which could compromise the infant's ability to explore and interact with his or her environment [@harris2008congenital; @gabis2021weak].
-A reflection of the aforementioned would be expressed in an altered development of gross motor function in the first months of life, with a consequent limitation in fine motor skills later on, which would have a subsequent negative impact on the communicative competence of infants, secondary to a reduced interaction with their environment and peers.
-The results found in our study are congruent with what other authors have discussed in relation to the role of gross and fine motor skills in language development [@gonzalez2019gross].
+A reflection of the aforementioned would be expressed in an altered development of GM function in the first months of life, with a consequent limitation in FM skills later on, which would have a subsequent negative impact on the communicative competence of infants, secondary to a reduced interaction with their environment and peers.
+The results found in our study are congruent with what other authors have discussed in relation to the role of gross and FM skills in language development [@gonzalez2019gross].
 

### 2022-07-01 manuscript/misc/figures.R

@@ -20,16 +20,20 @@ library(see)
 # Problem solving score ~ corrected age
 # Socio-individual score ~ corrected age
 
-y_vars <- c("Communication" = "comunicacion_total"
-            , "Fine motor" = "motora_fina_total"
-            , "Gross motor" = "motora_gruesa_total"
-            , "Problem solving" = "resolucion_problemas_total"
-            , "Socio-individual" = "socio_individual_total")
+## communication (CM), gross motor (GM), fine motor (FM), problem-solving (CG), and personal-social (PS)
+
+y_vars <- c("CM" = "comunicacion_total"
+            , "GM" = "motora_gruesa_total"
+            , "FM" = "motora_fina_total"
+            , "CG" = "resolucion_problemas_total"
+            , "PS" = "socio_individual_total")
 
 fig_data <- melt(
   data = dataset,
   measure.vars = y_vars
 )
+
+fig_data[, mean(value), variable]
 
 fig_data[, variable := `levels<-`(variable, names(y_vars))][]
 
@@ -50,11 +54,13 @@ fig_1a <- ggplot(fig_data, aes(x = edad_corregida_meses, y = value)) +
 
 # Figure 1b ------------------------------------------------------------------------------------
 
-models <- list("Communication" = NULL,
-               "Gross motor" = NULL,
-               "Fine motor" = NULL,
-               "Problem solving" = NULL,
-               "Socio-individual" = NULL)
+## communication (CM), gross motor (GM), fine motor (FM), problem-solving (CG), and personal-social (PS)
+
+models <- list("CM" = NULL,
+               "GM" = NULL,
+               "FM" = NULL,
+               "CG" = NULL,
+               "PS" = NULL)
 
 models[[1]] <- gam(
   comunicacion_total ~
@@ -107,7 +113,8 @@ slopes <- sapply(
 
 fig_1b_data <- rbindlist(slopes, idcol = "response")
 fig_1b_data[, Confidence := fifelse(p < 0.05, "Significant", "Not significant")
-          ][, grp := rleid(Confidence)][]
+          ][, grp := rleid(Confidence)
+          ][, response := factor(response, levels = names(models))]
 
 fig_1b <- ggplot(fig_1b_data, aes(edad_corregida_meses, Coefficient)) +
   facet_grid(response ~ ., scales = "free_y") +

### 2022-07-03 manuscript/manuscript.Rmd
figure 1 update

@@
 
+## Study design
+
+We conducted an observational, cross-sectional study, under a quantitative approach.
+
+[¿Los datos vienen de algún grupo de estudio o de cohorte en particular?]{custom-style="ErrorTok"}
+
+[¿Qué resguardos éticos se consideran? Comité de ética y otros permisos que sean pertinentes]{custom-style="ErrorTok"}
+
 ## Participants
@@
 
-A total of `r n_sample` subjects with BCH (females, `r n_sex["F", stat]`; males, `r n_sex["M", stat]`) were assessed.
+[¿Cómo fueron reclutados los sujetos? Criterios de elegibilidad, tipo de muestreo, etc.]{custom-style="ErrorTok"}
+
+[¿Cómo, cuando y quien hizo el diagnóstico de hipotonía congénita?]{custom-style="ErrorTok"}
 
@@
 
-[...].
+[¿Como, cuando y quién recolectó los datos demográficos?]{custom-style="ErrorTok"}
 
@@
 
-[...].
+[¿Cómo se evaluó el ASQ-3?
+¿Con qué protocolo?
+¿Que consideraciones se tuvieron en la evaluación?
+¿Algo más que podría ser relevante para agregar respecto a la evaluación o corrección del cuestionario?]{custom-style="ErrorTok"}
 
@@
             , "CG" = "resolucion_problemas_total"
-            , "personal-individual" = "socio_individual_total")
+            , "PS" = "socio_individual_total")
 
@@
 Relationship between corrected age (in months) and developmental domains.
-Left panel: regression lines and shaded area represent predicted values estimated from GAM models and their CI~95%~, points and error bars represent the mean and standard error at 5-month age intervals.
-Right panel: effect derivatives representing how the effect of corrected age (in months) in developmental domains changes across corrected age.
-Significant areas consider CI~95%~ that did not cross zero.
+Left panel: regression lines represent predicted values estimated from GAM models (red lines) and 100 bootstrap replicates (gray lines), points and error bars represent the mean and standard error at 5-month age intervals.
+Right panel: effect derivatives and their CI~95%~, representing how the effect of corrected age (in months) in developmental domains changes across corrected age.
+Significant areas consider CI~95%~ that do not cross zero.

### 2022-07-03 manuscript/misc/figures.R

@@ -33,22 +33,55 @@ fig_data <- melt(
   measure.vars = y_vars
 )
 
-fig_data[, mean(value), variable]
-
 fig_data[, variable := `levels<-`(variable, names(y_vars))][]
 
+# Scaled data for mean ± SE plotting
 scaled <- copy(fig_data)
 scaled[, edad_corregida_meses := ASQ3:::near(edad_corregida_meses, 5)][]
 
+# 300 evenly-spaced data points of `edad_corregida_meses`
+pred_data <- fig_data[, data.frame(
+  edad_corregida_meses = seq.default(
+    from = min(edad_corregida_meses, na.rm = TRUE),
+    to = max(edad_corregida_meses, na.rm = TRUE),
+    length.out = 300
+  )
+)]
+
+# Bootstrapping with 100 replicates
+boots <- fig_data[, {
+  boot <- vapply(1:100, function(i) { # Vectorized "looping"
+
+    # ith Model fitting
+    mod <- gam(value ~ s(edad_corregida_meses), method = "REML", data = .SD[sample(.N, replace = TRUE)])
+
+    # Estimation from bootstrapped model
+    predict(
+      object = mod,
+      newdata = pred_data
+    )
+  }, FUN.VALUE = numeric(length = nrow(pred_data)))
+
+  # Join the exposure and response
+  cbind(pred_data, boot)
+
+}, by = variable] # Grouped by variable
+
+# Melt the wide data into long data (for compatibility with ggplot)
+boots <- melt(boots, id.vars = 1:2, variable.name = "boot", value.name = "predicted")
+
+# Plot the data
 fig_1a <- ggplot(fig_data, aes(x = edad_corregida_meses, y = value)) +
   facet_grid(variable ~ ., scale = "free_y") +
   geom_count(alpha = 0.1, na.rm = TRUE) +
-  stat_summary(data = scaled, fun.data = mean_se, na.rm = TRUE) +
+  geom_line(data = boots, aes(group = boot, y = predicted), alpha = 0.1) +
   geom_smooth(method = "gam",
               formula = y ~ s(x),
               col = "red3",
               method.args = list(method = "REML"),
-              na.rm = TRUE) +
+              na.rm = TRUE,
+              se = FALSE) +
+  stat_summary(data = scaled, fun.data = mean_se, na.rm = TRUE) +
   labs(x = "Corrected age (months)", y = "Score") +
   ggpubr::theme_pubr()
 

### 2022-07-03 manuscript/misc/figures.R
figure 1 update

@@ -33,22 +33,56 @@ fig_data <- melt(
   measure.vars = y_vars
 )
 
-fig_data[, mean(value), variable]
-
 fig_data[, variable := `levels<-`(variable, names(y_vars))][]
 
+# Scaled data for mean ± SE plotting
 scaled <- copy(fig_data)
 scaled[, edad_corregida_meses := ASQ3:::near(edad_corregida_meses, 5)][]
 
+# 300 evenly-spaced data points of `edad_corregida_meses`
+pred_data <- fig_data[, data.frame(
+  edad_corregida_meses = seq.default(
+    from = min(edad_corregida_meses, na.rm = TRUE),
+    to = max(edad_corregida_meses, na.rm = TRUE),
+    length.out = 300
+  )
+)]
+
+# Bootstrapping with 100 replicates
+boots <- fig_data[, {
+  boot <- vapply(1:100, function(i) { # Vectorized "looping"
+
+    # ith Model fitting
+    mod <- gam(value ~ s(edad_corregida_meses), method = "REML", data = .SD[sample(.N, replace = TRUE)])
+
+    # Estimation from bootstrapped model
+    predict(
+      object = mod,
+      newdata = pred_data
+    )
+  }, FUN.VALUE = numeric(length = nrow(pred_data)))
+
+  # Join the exposure and response
+  cbind(pred_data, boot)
+
+}, by = variable] # Grouped by variable
+
+# Melt the wide data into long data (for compatibility with ggplot)
+boots <- melt(boots, id.vars = 1:2, variable.name = "boot", value.name = "predicted")
+
+# Plot the data
 fig_1a <- ggplot(fig_data, aes(x = edad_corregida_meses, y = value)) +
   facet_grid(variable ~ ., scale = "free_y") +
   geom_count(alpha = 0.1, na.rm = TRUE) +
-  stat_summary(data = scaled, fun.data = mean_se, na.rm = TRUE) +
+  geom_line(data = boots, aes(group = boot, y = predicted), alpha = 0.1) +
   geom_smooth(method = "gam",
               formula = y ~ s(x),
               col = "red3",
+              lwd = 0.7,
               method.args = list(method = "REML"),
-              na.rm = TRUE) +
+              na.rm = TRUE,
+              se = FALSE) +
+  stat_summary(data = scaled, fun.data = mean_se, na.rm = TRUE) +
   labs(x = "Corrected age (months)", y = "Score") +
   ggpubr::theme_pubr()
 

### 2022-07-03 manuscript/manuscript.Rmd
aesthetic changes to figures

@@
 Relationship between corrected age (in months) and developmental domains.
-Left panel: regression lines represent predicted values estimated from GAM models (red lines) and 100 bootstrap replicates (gray lines), points and error bars represent the mean and standard error at 5-month age intervals.
+Left panel: regression lines represent predicted values estimated from GAM models (bold red lines) and 200 bootstrap replicates (faded red lines), points and error bars represent the mean and standard error at 5-month age intervals.
 Right panel: effect derivatives and their CI~95%~, representing how the effect of corrected age (in months) in developmental domains changes across corrected age.

### 2022-07-03 manuscript/misc/figures.R

@@ -49,8 +49,9 @@ pred_data <- fig_data[, data.frame(
 )]
 
 # Bootstrapping with 100 replicates
+set.seed(123)
 boots <- fig_data[, {
-  boot <- vapply(1:100, function(i) { # Vectorized "looping"
+  boot <- vapply(1:200, function(i) { # Vectorized "looping"
 
     # ith Model fitting
     mod <- gam(value ~ s(edad_corregida_meses), method = "REML", data = .SD[sample(.N, replace = TRUE)])
@@ -74,14 +75,14 @@ boots <- melt(boots, id.vars = 1:2, variable.name = "boot", value.name = "predic
 fig_1a <- ggplot(fig_data, aes(x = edad_corregida_meses, y = value)) +
   facet_grid(variable ~ ., scale = "free_y") +
   geom_count(alpha = 0.1, na.rm = TRUE) +
-  geom_line(data = boots, aes(group = boot, y = predicted), alpha = 0.1) +
+  geom_line(data = boots, aes(group = boot, y = predicted), alpha = 0.02, col = "red") +
   geom_smooth(method = "gam",
               formula = y ~ s(x),
-              col = "red3",
+              col = "#ED0000FF",
               method.args = list(method = "REML"),
               na.rm = TRUE,
               se = FALSE) +
-  stat_summary(data = scaled, fun.data = mean_se, na.rm = TRUE) +
+  stat_summary(data = scaled, fun.data = mean_se, na.rm = TRUE, fatten = 3, col = "red4") +
   labs(x = "Corrected age (months)", y = "Score") +
   ggpubr::theme_pubr()
 

### 2022-07-04 manuscript/manuscript.Rmd
first complete version of manuscript

@@
 
-**Title**: Neurodevelopmental features among infants with benign congenital hypotonia in high southern latitudes: An observational cross-sectional trial.
+**Title**: Neurodevelopmental features among infants with congenital hypotonia in high southern latitudes: An observational cross-sectional study.
 
-**Authors**: Patricio Barría,^1^ Matías Castillo-Aguilar,^2,3^ ...,^4^ Cristian Núñez-Espinoza,^3,5^ \*
+**Authors**: Patricio Barría,^1^ Matías Castillo-Aguilar,^2,3^ Cristian Núñez-Espinoza,^3,4^ \*
 
@@
 
-^4^ Afiliación, lugar.
-
-^5^ School of Medicine, Magallanes University, Punta Arenas, Chile.
+^4^ School of Medicine, Magallanes University, Punta Arenas, Chile.
 :::
@@
 
-**Objective**: To describe and model the relationship between sociodemographics, prematurity and neurodevelopmental levels based on the *Ages and Stages Questionnaire* scores in infants diagnosed with benign congenital hypotonia (BCH).
+**Objective**: To describe and model the relationship between sociodemographics, prematurity and neurodevelopmental levels based on the *Ages and Stages Questionnaire* scores in infants diagnosed with congenital hypotonia (CH).
 **Material and methods**: [...].
-**Results**: [...].
+**Results**: When modelling the effect of corrected age on developmental domains we found a significant overall effect on communication skills ($\beta$ = -2.36, CI~95%~[-3.47, -1.25]), gross motor function ($\beta$ = 1.95, CI~95%~[0.66, 3.25]) and problem-solving skills ($\beta$ = -1.87, CI~95%~[-3.17, -0.57]).
+fine motor skills did not exhibit an overall linear effect ($\beta$ = 0.04, CI~95%~[-0.45, 0.52]), however, they showed a non-linear effect ($F_{smooth}$ (2.59, 226.77) = 4.2, *p* = 0.005); unlike the personal-social domain, which did not present any variation across corrected age ($F_{smooth}$ (1, 231.58) = 1.16, *p* = 0.282).
 **Conclusion**: After adjusting for possible confounders, we found between-subjects fluctuations in neurodevelopmental traits across age in hypotonic infants in the form of non-linear and domain-specific variations.
@@
 
-**Keywords**: [...].
+**Keywords**: Developmental traits, Congenital hypotonia, Ages and Stages Questionnaire (ASQ), Infants.
 
@@
 
-<!---- Antecedentes históricos ---->
 Both physical and psychological signs of early childhood development have been shown to be representative and relevant markers for the identification and monitoring of overall growth in early life [@di2016predictive] and can, therefore, be used in the screening of children at risk of developmental delay to support early referral and need further assessment to determine if they are eligible for early intervention services [@guralnick2017early; @bruder2010early].
@@
 Hypotonia has been defined both as decreased muscle tone or floppiness, involving a wide range and levels of progression [@harris2008congenital; @gabis2021weak].
-There are multiple forms of neuromuscular, metabolic and genetic conditions associated with hypotonia and it may be a sign of neurodevelopmental delay, that may predispose to cognitive impairment [@riou2009global].
+There are multiple forms of neuromuscular, metabolic and genetic conditions associated with hypotonia and it may be a sign of neurodevelopmental delay, that may predispose to cognitive impairment in some cases [@riou2009global].
 Given that hypotonia, and the hyperlaxity and motor delay associated with it, may impair the infant's capacity to engage with its surroundings, critical visual cues may be ignored, potentially leading both to an impairment of learning and cognitive development [@harris2008congenital], hence the need to explore the neurodevelopmental attributes of infants with hypotonia.
 
-Accordingly, benign congenital hypotonia (BCH) is usually considered an exclusion diagnosis, and is usually made in the absence of other signs and symptoms, after every other evaluative resources have been exhausted [@leyenaar2005schematic; @gabis2021weak], however this is not true for at least one author that suggest that BCH can not be considered a diagnosis as such [@thompson2002benign].
-BCH is considered a non-progressive neuromuscular disorder that does not progress but tends to improve with time and early intervention.
+Accordingly, benign congenital hypotonia (CH) is usually considered an exclusion diagnosis, and is usually made in the absence of other signs and symptoms, after every other evaluative resources have been exhausted [@leyenaar2005schematic; @gabis2021weak], however this is not true for at least one author that suggest that CH can not be considered a diagnosis as such [@thompson2002benign].
+CH is considered a non-progressive neuromuscular disorder that does not progress but tends to improve with time and early intervention.
 
-Nevertheless, and to the best of our knowledge, there is no robust evidence characterising the observed variation in developmental traits in infants with known BCH across age.
-Therefore, our main objective in this study was to describe and model the relationship between sociodemographics, prematurity and neurodevelopmental levels based on ASQ-3 scores in infants with diagnosed BCH.
+Nevertheless, and to the best of our knowledge, there is no robust evidence characterising the observed variation in developmental traits in infants with known CH across age.
+Therefore, our main objective in this study was to describe and model the relationship between sociodemographics, prematurity and neurodevelopmental levels based on ASQ-3 scores in infants with diagnosed CH.
 
@@
 
-[¿Los datos vienen de algún grupo de estudio o de cohorte en particular?]{custom-style="ErrorTok"}
+[¿Los datos vienen de algún grupo de estudio, cohorte o de algún proyecto en particular? Detalles de donde vino la muestra]{custom-style="ErrorTok"}
 
@@
 
-[¿Cómo fueron reclutados los sujetos? Criterios de elegibilidad, tipo de muestreo, etc.]{custom-style="ErrorTok"}
+[¿Cómo, cuando y bajo qué condiciones fueron reclutados los sujetos? Criterios de elegibilidad, tipo de muestreo, etc.]{custom-style="ErrorTok"}
 
-[¿Cómo, cuando y quien hizo el diagnóstico de hipotonía congénita?]{custom-style="ErrorTok"}
+[¿Cómo, cuando y quien hizo el diagnóstico de hipotonía congénita? ¿Fue igual para todos? ¿La información vino de parte de terceros?]{custom-style="ErrorTok"}
 
@@
 
-The ASQ-3 is a parent reported initial level developmental screening instrument consisting of 21 intervals, each with 30 items in five areas: i) communication (CM), ii) gross motor (GM), iii) fine motor (FM), iv) problem-solving (CG), and v) personal-social (PS).
-The ASQ is cost-effective and widely used in the United States and other countries.
-The ASQ has been translated into several languages, and the number of international studies on its psychometric properties with diverse cultural environments is increasing.
-It has excellent psychometric properties, test-retest reliability of 92%, sensitivity of 87.4% and specificity of 95.7%.
-Its validity has been examined across different cultures and communities across the world.
+The ASQ-3 is a parent reported initial level developmental screening instrument consisting of 21 intervals, each with 30 items in five areas: i) communication (CM), ii) gross motor (GM), iii) fine motor (FM), iv) problem-solving (CG), and v) personal-social (PS) [@squires2009ages].
+The ASQ is cost-effective and widely used in the United States and other countries [@sarmiento2011universal; @heo2008cross].
+It has been translated into several languages, and the number of international studies on its psychometric properties with diverse cultural environments is increasing [@sarmiento2011universal; @heo2008cross].
+It has shown adequate psychometric properties (75% sensitivity and 81% specificity) in Chilean term and preterm infants [@schonhaut2013validity].
 
@@
 All the statistical analyses were computed and implemented in the R programming language [@rlanguage].
-GAMs and the corresponding model estimates were computed using the *mgcv* and *modelbased* packages, each with well documented functions and methods [@wood2017generalized; @dominique2020estimation].
+GAMs and the corresponding model estimates were computed using the *mgcv* and *modelbased* packages [@wood2017generalized; @dominique2020estimation].
 Complementary R packages were used for visualization purposes [@hadley2016ggplot2; @daniel2021see].
@@
   chisq.test() |> 
-  report_chi(effsize = TRUE)
+  report_chi(effsize = FALSE)
 #> [1] "$\\chi^2_{gof}$ (1) = 9.04, *p* = 0.003, $C_{Pearson}$ = 0.19, CI~0.95%~[0.09, 1]"
@@
 
-Our study aimed to describe and model the relationship between sociodemographic data, prematurity and neurodevelopmental levels based on ASQ-3 scores in infants diagnosed with BCH.
+Our study aimed to describe and model the relationship between sociodemographic data, prematurity and neurodevelopmental levels based on ASQ-3 scores in infants diagnosed with CH.
 Our main findings suggest a non-linear effect of age, corrected for prematurity, with a marked decrease in scores for all neurodevelopmental traits at different age frames, even after adjusting for caregiver relationship, sex and inter-rater influence.
-However in the personal-individual domain, there was no variation observed across corrected age.
+However in the PS domain, there was no variation observed across corrected age.
 
-This could be associated with the described motor impairments of hypotonia in the early stages of life, which could compromise the infant's ability to explore and interact with his or her environment [@harris2008congenital; @gabis2021weak].
-A reflection of the aforementioned would be expressed in an altered development of GM function in the first months of life, with a consequent limitation in FM skills later on, which would have a subsequent negative impact on the communicative competence of infants, secondary to a reduced interaction with their environment and peers.
-The results found in our study are congruent with what other authors have discussed in relation to the role of gross and FM skills in language development [@gonzalez2019gross].
+This findings might be associated with the described motor impairments of hypotonia in the early stages of life, which compromise the infant's ability to explore and interact with his or her environment [@harris2008congenital; @gabis2021weak].
+A reflection of the aforementioned would be expressed in an altered development of GM function in the first months of life, with a consequent limitation in FM skills later on, which would have a subsequent negative impact on the communicative competence of infants, secondary to a reduced interaction with their environment and peers [@bodensteiner2008evaluation].
+In a recent systematic review [@gonzalez2019gross], differences in the predictive abilities of gross and FM skills on communication skills in infants and early childhood were reported, whereas GM skills, such as crawling and walking, favour exploration with their environment and caregivers, while FM skills, expressed through tasks such as drawing and handling utensils, could lead to improvements in language through mechanisms yet to be explored [@gonzalez2019gross].
+These milestones may be impaired in the face of poorer head and trunk control in CH, which have been shown to delay the achievement of key motor milestones in infants [@harris2008congenital; @bodensteiner2008evaluation; @gabis2021weak].
 
 It is worth noting that other context-mediated social factors may also have influenced our results, mainly due to the role that other variables would also play in the neurodevelopment of our study sample, which could have an impact on many of the developmental traits assessed here, such as intrauterine growth restrictions, maternal depression, institutionalisation, exposure to social violence, maternal education and breastfeeding [@walker2011inequality].
-All together, these represent the main limitations in our study design, which need to be addressed in future research exploring the variations observed in different developmental traits in hypotonic infants.
-However, our study sheds light to an underexplored aspect of congenital myopathies, with robust statistical methods that made it possible to model and capture the complex relationships seen early in life.
+All together, these represent the main limitations in our study design, which need to be addressed in future research, exploring the variations observed in different developmental traits in hypotonic infants.
+However, our study sheds light to an underexplored aspect of congenital myopathies, with robust statistical methods that made it possible to capture and model the complex relationships seen early in life.
 
@@
 
-The present study shows that the marked variations observed in neurodevelopmental traits are present across age in hypotonic infants, mainly in the form of non-linear and domain-specific variations, even after adjusting for the effect that caregiver relationship, gender and evaluators might exert.
+The present study shows that the marked variations observed in neurodevelopmental traits are present across age in hypotonic infants, mainly in the form of non-linear and domain-specific variations, even after adjusting for the effect that caregiver relationship, sex and evaluators might exert.
 Moreover, we show that the observed variations in developmental domains are not solely attributable to prematurity, where age corrected for prematurity best explained the observed variability in neurodevelopment.
@@
 
-# Acknowledgment
+# Data availability statement
 
-[...].
+The raw data supporting the conclusions of this article will be made available by the authors, without undue reservation.
+
+# Author Contributions
+
+All authors listed have made a substantial, direct and intellectual contribution to the work, and approved it for publication.
+
+# Funding
+
+[Añadir financiamiento si fuese aplicable]{custom-style="ErrorTok"}
 
@@
 The authors declare that the research was conducted in the absence of any commercial or financial relationships that could be construed as a potential conflict of interest.
+
+# Acknowledgment
+
+[...]
 

### 2022-07-04 manuscript/misc/figures.R

@@ -71,6 +71,8 @@ boots <- fig_data[, {
 # Melt the wide data into long data (for compatibility with ggplot)
 boots <- melt(boots, id.vars = 1:2, variable.name = "boot", value.name = "predicted")
 
+
+
 # Plot the data
 fig_1a <- ggplot(fig_data, aes(x = edad_corregida_meses, y = value)) +
   facet_grid(variable ~ ., scale = "free_y") +

### 2022-07-04 R/report_stats.R
new functions for reporting non-parametric tests (in case of need), I very much think I'll need them.

@@ -68,6 +68,149 @@ report_chi <- function(chi, effsize = TRUE, ci = 0.95) {
   return(expr)
 }
 
+#' Report Kruskal-Wallis chi-squared test
+#'
+#' This function lets you run a Kruskal-Wallis chi-squared test and report it.
+#'
+#' @param formula A formula specifying the a analysis.
+#' @param data The dataset from where to search the variables.
+#' @param effsize Logical. Whether to report effect sizes or not.
+#' @param ci Numeric. Confidence interval to effect size calculation.
+#' @param ... Other parameters passed to `kruskal.test`.
+#'
+#' @name report_kruskal
+#' @export
+
+report_kruskal <- function(formula, data, effsize = TRUE, ci = 0.95, ...) {
+
+  stopifnot(
+    "Is not an object of class \"formula\"" = inherits(formula, "formula")
+  )
+
+  if (isTRUE(effsize)) {
+    efs <- effectsize::rank_epsilon_squared(formula, data = data, ci = ci)
+    efs <- lapply(efs, round, 2)
+  }
+
+  kw <- stats::kruskal.test(formula, data, ...)
+  kw$statistic <- round(kw$statistic, 2)
+
+  kw$p.value <- format_p(kw$p.value)
+
+  kw$method <- "Kruskal-Wallis"
+
+  expr <- paste(
+    paste0("$\\chi^2_{",kw$method,"}$ (",kw$parameter,") = ",kw$statistic),
+    paste0("*p* ",kw$p.value),
+    sep = ", "
+  )
+
+  if (isTRUE(effsize)) {
+    expr <- paste(
+      expr,
+      paste0("$\\widehat{\\epsilon}^2$ = ", efs[[1]]),
+      paste0("CI~", efs$CI,"%~[", efs$CI_low,", ",efs$CI_high,"]"),
+      sep = ", "
+    )
+  }
+
+  return(expr)
+}
+
+#' Report Wilcoxon rank sum test
+#'
+#' This function lets you run a Wilcoxon rank sum test and report it.
+#'
+#' @param formula A formula specifying the a analysis.
+#' @param data The dataset from where to search the variables.
+#' @param effsize Logical. Whether to report effect sizes or not.
+#' @param ci Numeric. Confidence interval to effect size calculation.
+#' @param ... Other parameters passed to `kruskal.test`.
+#'
+#' @name report_wilcox
[…]

### 2022-07-19 manuscript/manuscript.Rmd
update ms
- add methods that were missing

@@
 
+```{r load-packages}
+source(file = "manuscript/misc/setup.R")
+```
+
 **Title**: Neurodevelopmental features among infants with congenital hypotonia in high southern latitudes: An observational cross-sectional study.
 
-**Authors**: Patricio Barría,^1^ Matías Castillo-Aguilar,^2,3^ Cristian Núñez-Espinoza,^3,4^ \*
+**Authors**: Patricio Barría^1^, Matías Castillo-Aguilar^2,3^, Katherine Harris^2,3^, Cristian Núñez-Espinosa^3,4^ \*
 
@@
 
-Nevertheless, and to the best of our knowledge, there is no robust evidence characterising the observed variation in developmental traits in infants with known CH across age.
+Nevertheless, and to the best of our knowledge, there is no robust evidence characterising the observed variation in developmental traits in infants with diagnosed CH across age.
 Therefore, our main objective in this study was to describe and model the relationship between sociodemographics, prematurity and neurodevelopmental levels based on ASQ-3 scores in infants with diagnosed CH.
@@
 
-[¿Los datos vienen de algún grupo de estudio, cohorte o de algún proyecto en particular? Detalles de donde vino la muestra]{custom-style="ErrorTok"}
-
-[¿Qué resguardos éticos se consideran? Comité de ética y otros permisos que sean pertinentes]{custom-style="ErrorTok"}
-
 ## Participants
-
-```{r load-packages}
-source(file = "manuscript/misc/setup.R")
-```
 
@@
 #> 234
-
-# Sample by sex
-n_sex <- dataset[j = .(stat = paste0(.N, " (", round(.N*100/n_sample, 1), "%)")), 
-                 keyby = .(sex = sexo_paciente)]
-#>    sex        stat
-#> 1:   F  94 (40.2%)
-#> 2:   M 140 (59.8%)
 ```
 
-[¿Cómo, cuando y bajo qué condiciones fueron reclutados los sujetos? Criterios de elegibilidad, tipo de muestreo, etc.]{custom-style="ErrorTok"}
+A total of `r n_sample` patients were enrolled as part of an intervention programme carried out by the *Cruz del Sur* Rehabilitation Centre (Punta Arenas, Chile), admitted from one month of age to 60 months, and assessed at admission, control and discharge.
+The total records of patients diagnosed with CH admitted to the institution's programme were analysed and the assessment of the subjects was conducted and guided by a nurse trained in the application of the ASQ-3.
 
-[¿Cómo, cuando y quien hizo el diagnóstico de hipotonía congénita? ¿Fue igual para todos? ¿La información vino de parte de terceros?]{custom-style="ErrorTok"}
+In this context, it is worth noting that the data, as well as the patient registry, are part of institutional strategies aimed at the diagnosis and continuous monitoring of the clinical situation of users, which are used to improve care processes and clinical decision-making.
+
+The diagnosis of CH is based on 4 sources: 1) primary care paediatrician, who refers the diagnosis to the institution, where admission to the programme is made; 2) paediatricians from private clinics, who refer the diagnosis; 3) neuropaediatricians from the clinical hospital, who refer the diagnosis to the programme; 4) physiatrist from the institution, who assigns the diagnosis if it corresponds to the semiology.
 
@@
 
-The ASQ-3 is a parent reported initial level developmental screening instrument consisting of 21 intervals, each with 30 items in five areas: i) communication (CM), ii) gross motor (GM), iii) fine motor (FM), iv) problem-solving (CG), and v) personal-social (PS) [@squires2009ages].
+The ASQ-3 is a parent-reported initial level developmental screening instrument, consisting of 21 intervals, each with 30 items in five areas: i) communication (CM), ii) gross motor (GM), iii) fine motor (FM), iv) problem-solving (CG), and v) personal-social (PS) [@squires2009ages].
 The ASQ is cost-effective and widely used in the United States and other countries [@sarmiento2011universal; @heo2008cross].
@@
 
+For the collection of research data, an authorisation consent was obtained for the use of instrumental clinical data as well as the clinical record of each patient.
+Subsequently, and in accordance with national research regulations, written authorisation was obtained from the institutional director for the use of the records and database for research purposes.
+To this end, the data were anonymised during data processing and subsequent analyses.
+
 ### Collection of demographic data
 
-[¿Como, cuando y quién recolectó los datos demográficos?]{custom-style="ErrorTok"}
+The demographic data of each patient was collected and made available throughout the study by the institutional electronic systems at the time of entry to the programme.
+The administrative registration was done by the secretaries of the user coordination unit, recording name, ID number, date of birth and diagnosis.
+In addition, age data was automatically updated by the computer system's algorithms.
 
@@
 
-[¿Cómo se evaluó el ASQ-3?
-¿Con qué protocolo?
-¿Que consideraciones se tuvieron en la evaluación?
-¿Algo más que podría ser relevante para agregar respecto a la evaluación o corrección del cuestionario?]{custom-style="ErrorTok"}
+The ASQ-3 was applied according to the protocol established by the instrument itself, with a face-to-face or telematic application being valid.
+The protocol can be found in the instrument's manuals, which were administered by the programme nurse.
+Some important considerations of the administration are that i) the main caregiver must respond, ii) in case of doubts of the caregiver about the assessed behaviour, the information is corroborated by an in situ test with the user, and iii) in relation to the correction, these procedures are standardised by the test itself.
 
@@
 ```{r}
+# Sample by sex
+n_sex <- dataset[j = .(stat = paste0(.N, " (", round(.N*100/n_sample, 1), "%)")), 
+                 keyby = .(sex = sexo_paciente)]
+#>    sex        stat
+#> 1:   F  94 (40.2%)
+#> 2:   M 140 (59.8%)
+
 # Prop test on sex
@@
   report_chi(effsize = FALSE)
-#> [1] "$\\chi^2_{gof}$ (1) = 9.04, *p* = 0.003, $C_{Pearson}$ = 0.19, CI~0.95%~[0.09, 1]"
+#> [1] "$\\chi^2_{gof}$ (1) = 9.04, *p* = 0.003"
 ```
@@
 
+No existe financiamiento extraordinario a la aplicación del instrumento.
+
 # Conflicts of interest

### 2022-07-26 manuscript/manuscript.Rmd
ms update:
- abstract update

@@
 
-**Objective**: To describe and model the relationship between sociodemographics, prematurity and neurodevelopmental levels based on the *Ages and Stages Questionnaire* scores in infants diagnosed with congenital hypotonia (CH).
-**Material and methods**: [...].
-**Results**: When modelling the effect of corrected age on developmental domains we found a significant overall effect on communication skills ($\beta$ = -2.36, CI~95%~[-3.47, -1.25]), gross motor function ($\beta$ = 1.95, CI~95%~[0.66, 3.25]) and problem-solving skills ($\beta$ = -1.87, CI~95%~[-3.17, -0.57]).
-fine motor skills did not exhibit an overall linear effect ($\beta$ = 0.04, CI~95%~[-0.45, 0.52]), however, they showed a non-linear effect ($F_{smooth}$ (2.59, 226.77) = 4.2, *p* = 0.005); unlike the personal-social domain, which did not present any variation across corrected age ($F_{smooth}$ (1, 231.58) = 1.16, *p* = 0.282).
-**Conclusion**: After adjusting for possible confounders, we found between-subjects fluctuations in neurodevelopmental traits across age in hypotonic infants in the form of non-linear and domain-specific variations.
+**Objective**: To describe and model the relationship between sociodemographics, prematurity and neurodevelopmental levels based on the *Ages and Stages Questionnaire* (ASQ-3) scores in infants diagnosed with congenital hypotonia (CH).
+**Material and methods**: A total of 234 patients diagnosed with CH were enrolled in a rehabilitation programme led by the Rehabilitation Centre of Punta Arenas (Chile).
+Neurodevelopmental status was assessed with the ASQ-3 at admission, as well as sociodemographic and obstetric data, which were obtained through the initial clinical interview.
+**Results**: When modelling the neurodevelopmental status of each domain, an overall negative effect of corrected age on communication and problem-solving skills was observed, whereas the overall effect tend to be positive in gross motor function.
+Fine motor skills did not exhibit a linear relationship with corrected age, but a non-linear effect; unlike the personal-social domain, which did not present significant variations across age corrected for prematurity.
+**Conclusion**: After adjusting for possible confounders, we found between-subjects fluctuations in neurodevelopmental traits across age in hypotonic infants.
+These fluctuations were present in the form of non-linear and domain-specific variations.
 Further research is warranted to determine how these findings apply in the presence of other context-mediated social factors and populations.
@@
 
-# Funding
-
-[Añadir financiamiento si fuese aplicable]{custom-style="ErrorTok"}
-
-No existe financiamiento extraordinario a la aplicación del instrumento.
-
 # Conflicts of interest
@@
 The authors declare that the research was conducted in the absence of any commercial or financial relationships that could be construed as a potential conflict of interest.
-
-# Acknowledgment
-
-[...]
 

### 2022-07-27 manuscript/manuscript.Rmd
adding Cristian's changes to ms

@@
 csl: "misc/apa.csl"
-link-citations: true
+link-citations: yes
 output:
@@
 
-**Authors**: Patricio Barría^1^, Matías Castillo-Aguilar^2,3^, Katherine Harris^2,3^, Cristian Núñez-Espinosa^3,4^ \*
+**Authors**: Patricio Barría^1,2^, Matías Castillo-Aguilar^3,4^, Katherine Harris^3,4^, Cristian Núñez-Espinosa^4,5,^\*
 
 ::: {custom-style="Afiliation"}
-^1^ Rehabilitation Center, Punta Arenas, Chile.
+^1^ Unidad de Investigación, Corporación de Rehabilitación Cruz del Sur, Punta Arenas, Chile.
 
-^2^ Kinesiology Department, Magallanes University, Punta Arenas, Chile.
+^2^ Brain-Machine Interface Systems Lab, Universidad Miguel Hernández de Elche, Elche, España
 
-^3^ Chilean Austral Integrative Neurophysiology Group ([NIM-ACh](https://nimach.org)), Centro Asistencial de Docencia e Investigación (CADI-UMAG), Punta Arenas Chile
+^3^ Kinesiology Department, Magallanes University, Punta Arenas, Chile.
 
-^4^ School of Medicine, Magallanes University, Punta Arenas, Chile.
+^4^ Chilean Austral Integrative Neurophysiology Group ([NIM-ACh](https://nimach.org)), Centro Asistencial de Docencia e Investigación (CADI-UMAG), Punta Arenas Chile
+
+^5^ School of Medicine, Magallanes University, Punta Arenas, Chile.
 :::
@@
 **Material and methods**: A total of 234 patients diagnosed with CH were enrolled in a rehabilitation programme led by the Rehabilitation Centre of Punta Arenas (Chile).
-Neurodevelopmental status was assessed with the ASQ-3 at admission, as well as sociodemographic and obstetric data, which were obtained through the initial clinical interview.
-**Results**: When modelling the neurodevelopmental status of each domain, an overall negative effect of corrected age on communication and problem-solving skills was observed, whereas the overall effect tend to be positive in gross motor function.
-Fine motor skills did not exhibit a linear relationship with corrected age, but a non-linear effect; unlike the personal-social domain, which did not present significant variations across age corrected for prematurity.
+Neurodevelopmental status was assessed with the ASQ-3 at admission, as well as sociodemographic and obstetric data obtained through the initial clinical interview.
+**Results**: When modelling the neurodevelopmental status of each domain, an overall negative effect of corrected age on communication and problem-solving skills was observed, whereas the overall effect tends to be positive in gross motor function.
+Fine motor skills did not exhibit a linear relationship with corrected age but a non-linear effect, unlike the personal-social domain, which did not present significant variations across age corrected for prematurity.
 **Conclusion**: After adjusting for possible confounders, we found between-subjects fluctuations in neurodevelopmental traits across age in hypotonic infants.
@@
 
-Both physical and psychological signs of early childhood development have been shown to be representative and relevant markers for the identification and monitoring of overall growth in early life [@di2016predictive] and can, therefore, be used in the screening of children at risk of developmental delay to support early referral and need further assessment to determine if they are eligible for early intervention services [@guralnick2017early; @bruder2010early].
+The physical and psychological signs of early childhood development are representative and relevant markers for the identification and monitoring of overall growth in early life [@di2016predictive].
+Therefore, both can be used in screening children at risk of developmental delay to support early referral and need further assessment to determine if they are eligible for early intervention services [@guralnick2017early; @bruder2010early].
 
-Currently, a plethora of tools have been proposed to assess the developmental continuum of infants.
-In this sense, the Ages and Stages Questionnaire, Third Edition (ASQ-3) has been proposed as a global screening tool, parent/caregivers-oriented, that assesses five domains of development in children aged from 0 to 5.5 years old [@singh2017ages].
-Current evidence suggests that the ASQ-3 is an accurate, cost-effective yet parent-friendly instrument for screening, monitoring children up to pre-school age, and can help identify and exclude neurodevelopmental impairments in very preterm-born children [@kerstjens2015ages; @ballantyne2016risk; @singh2017ages].
+Currently, many tools have been proposed to assess the developmental continuum of infants.
+In this sense, the Ages and Stages Questionnaire, Third Edition (ASQ-3) has been submitted as a global screening tool, parent/caregivers-oriented, that assesses five domains of development in children aged from 0 to 5.5 years old [@singh2017ages].
+Current evidence suggests that the ASQ-3 is an accurate, cost-effective, yet parent-friendly instrument for screening and monitoring children up to pre-school age and can help identify and exclude neurodevelopmental impairments in very preterm-born children [@kerstjens2015ages; @ballantyne2016risk; @singh2017ages].
 
-Hypotonia has been defined both as decreased muscle tone or floppiness, involving a wide range and levels of progression [@harris2008congenital; @gabis2021weak].
-There are multiple forms of neuromuscular, metabolic and genetic conditions associated with hypotonia and it may be a sign of neurodevelopmental delay, that may predispose to cognitive impairment in some cases [@riou2009global].
-Given that hypotonia, and the hyperlaxity and motor delay associated with it, may impair the infant's capacity to engage with its surroundings, critical visual cues may be ignored, potentially leading both to an impairment of learning and cognitive development [@harris2008congenital], hence the need to explore the neurodevelopmental attributes of infants with hypotonia.
+Multiple forms of neuromuscular, metabolic and genetic conditions are associated with hypotonia and may be a sign of neurodevelopmental delay, which may predispose to cognitive impairment in some cases [@riou2009global].
+Hypotonia has been defined as decreased muscle tone or floppiness, involving a wide range and levels of progression [@harris2008congenital; @gabis2021weak].
+Given that hypotonia and the hyperlaxity and motor delay associated with it may impair the infant's capacity to engage with its surroundings, critical visual cues may be ignored, potentially leading both to an impairment of learning and cognitive development [@harris2008congenital], hence the need to explore the neurodevelopmental attributes of infants with hypotonia.
 
-Accordingly, benign congenital hypotonia (CH) is usually considered an exclusion diagnosis, and is usually made in the absence of other signs and symptoms, after every other evaluative resources have been exhausted [@leyenaar2005schematic; @gabis2021weak], however this is not true for at least one author that suggest that CH can not be considered a diagnosis as such [@thompson2002benign].
-CH is considered a non-progressive neuromuscular disorder that does not progress but tends to improve with time and early intervention.
+The congenital hypotonia (CH) is usually considered an exclusion diagnosis, and is generally made in the absence of other signs and symptoms, after every other evaluative resource has been exhausted [@leyenaar2005schematic; @gabis2021weak].
+However, this is not true for at least one author that suggests that CH can not be considered a diagnosis as such [@thompson2002benign].
+CH is a non-progressive neuromuscular disorder but tends to improve with time and early intervention [@gabis2021weak].
 
-Nevertheless, and to the best of our knowledge, there is no robust evidence characterising the observed variation in developmental traits in infants with diagnosed CH across age.
+Nevertheless, to the best of our knowledge, there is no robust evidence characterising the observed variation in developmental traits in infants with diagnosed CH across ages.
 Therefore, our main objective in this study was to describe and model the relationship between sociodemographics, prematurity and neurodevelopmental levels based on ASQ-3 scores in infants with diagnosed CH.
@@
 
-We conducted an observational, cross-sectional study, under a quantitative approach.
+We conducted an observational, cross-sectional study under a quantitative approach.
 
@@
 
-A total of `r n_sample` patients were enrolled as part of an intervention programme carried out by the *Cruz del Sur* Rehabilitation Centre (Punta Arenas, Chile), admitted from one month of age to 60 months, and assessed at admission, control and discharge.
-The total records of patients diagnosed with CH admitted to the institution's programme were analysed and the assessment of the subjects was conducted and guided by a nurse trained in the application of the ASQ-3.
+A total of `r n_sample` patients were enrolled as part of an intervention programme carried out by the *Cruz del Sur* Rehabilitation Centre (Punta Arenas, Chile), admitted from one month to 60 months, and assessed at admission, control and discharge.
+The total records of patients diagnosed with CH admitted to the institution's programme were analysed, and the assessment of the subjects was conducted and guided by a nurse trained in the application of the ASQ-3.
 
@@
 
-The diagnosis of CH is based on 4 sources: 1) primary care paediatrician, who refers the diagnosis to the institution, where admission to the programme is made; 2) paediatricians from private clinics, who refer the diagnosis; 3) neuropaediatricians from the clinical hospital, who refer the diagnosis to the programme; 4) physiatrist from the institution, who assigns the diagnosis if it corresponds to the semiology.
+The diagnosis of CH is based on four sources: 1) primary care paediatrician, who refers the diagnosis to the institution where admission to the programme is made; 2) paediatricians from private clinics, who refer the diagnosis; 3) neuropaediatricians from the clinical hospital, who refer the diagnosis to the programme; 4) physiatrist from the institution, who assigns the diagnosis if it corresponds to the semiology.
 
@@
 
-The ASQ-3 is a parent-reported initial level developmental screening instrument, consisting of 21 intervals, each with 30 items in five areas: i) communication (CM), ii) gross motor (GM), iii) fine motor (FM), iv) problem-solving (CG), and v) personal-social (PS) [@squires2009ages].
+The ASQ-3 is a parent-reported initial level developmental screening instrument consisting of 21 intervals, each with 30 items in five areas: i) communication (CM), ii) gross motor (GM), iii) fine motor (FM), iv) problem-solving (CG), and v) personal-social (PS) [@squires2009ages].
 The ASQ is cost-effective and widely used in the United States and other countries [@sarmiento2011universal; @heo2008cross].
-It has been translated into several languages, and the number of international studies on its psychometric properties with diverse cultural environments is increasing [@sarmiento2011universal; @heo2008cross].
-It has shown adequate psychometric properties (75% sensitivity and 81% specificity) in Chilean term and preterm infants [@schonhaut2013validity].
+It has been translated into several languages, and international studies on its psychometric properties in diverse cultural environments are increasing [@sarmiento2011universal; @heo2008cross].
+It has shown good psychometric properties (75% sensitivity and 81% specificity) in Chilean term and preterm infants [@schonhaut2013validity].
 
@@
 
-For the collection of research data, an authorisation consent was obtained for the use of instrumental clinical data as well as the clinical record of each patient.
-Subsequently, and in accordance with national research regulations, written authorisation was obtained from the institutional director for the use of the records and database for research purposes.
+For the collection of research data, authorisation consent was obtained for the use of instrumental clinical data as well as the clinical record of each patient.
+Subsequently, and following national research regulations, written authorisation was obtained from the institutional director to use the forms and database for research purposes.
 To this end, the data were anonymised during data processing and subsequent analyses.
@@
 
-The demographic data of each patient was collected and made available throughout the study by the institutional electronic systems at the time of entry to the programme.
+Each patient's demographic data were collected and made available throughout the study by the institutional electronic systems at the time of entry to the programme.
 The administrative registration was done by the secretaries of the user coordination unit, recording name, ID number, date of birth and diagnosis.
@@
 The ASQ-3 was applied according to the protocol established by the instrument itself, with a face-to-face or telematic application being valid.
-The protocol can be found in the instrument's manuals, which were administered by the programme nurse.
-Some important considerations of the administration are that i) the main caregiver must respond, ii) in case of doubts of the caregiver about the assessed behaviour, the information is corroborated by an in situ test with the user, and iii) in relation to the correction, these procedures are standardised by the test itself.
+The protocol can be found in the instrument's manuals, which the programme nurse administered.
+Some important considerations of the administration are that i) the primary caregiver must respond, ii) in case of doubts of the caregiver about the assessed behaviour, the information is corroborated by in situ tests with the user, and iii) in relation to the correction, the test itself standardises these procedures.
 
@@
 
-Data is presented as median (*Mdn*) and interquartile range (*IQR*) for continuous variables; for categorical/discrete variables, the absolute and relative sample size was reported.
+Data are presented as median (*Mdn*) and interquartile range (*IQR*) for continuous variables; for categorical/discrete variables, the absolute and relative sample size was reported.
 
-A non-parametric approach was used since the underlying distribution of continuous measured outcomes, assessed through analytical and graphical methods, did not follow a Gaussian distribution.
+A non-parametric approach was used since the underlying distribution of measured outcomes, assessed through analytical and graphical methods, did not follow a Gaussian distribution.
 
-In order assess the differences in developmental scores between males and females, the *Wilcoxon* rank-sum test was used, meanwhile the chi-square test ($\chi^2$) was used to evaluate goodness-of-fit ($\chi^2_{gof}$) and independence of factors ($\chi^2_{Pearson}$).
+In order to assess the differences in developmental scores between males and females, the *Wilcoxon* rank-sum test was used, meanwhile the chi-square test ($\chi^2$) was used to evaluate goodness-of-fit ($\chi^2_{gof}$) and independence of factors ($\chi^2_{Pearson}$).
 
 Generalized additive models (GAM) were used to describe linear and non-linear relationships in the form of smooth terms between developmental characteristics, represented through penalized regression splines [@wood2011fast].
-Restricted maximum likelihood method was used for the estimation of the smoothing parameters, and thin-plate regression splines as the smoothing basis, as they are the optimal smoother of any given basis dimension/rank [@wood2003thin].
+The restricted maximum likelihood method was used to estimate the smoothing parameters, and thin-plate regression splines as the smoothing basis, as they are the optimal smoother of any given basis dimension/rank [@wood2003thin].
 In the final models, infants' sex, clinician and infants' relationship with caregivers were added as random effects in the form of penalized parametric terms to account for the variability arising from these variables in the fixed effects analysed [@wood2016smoothing].
-To describe the smooth terms by means of quasi-linear segments, we used approximative derivatives with 95% confidence intervals (CI~95%~).
+We used approximative derivatives with 95% confidence intervals (CI95%) to describe the smooth terms by means of quasi-linear segments.
 
-A probability of committing a type I ($\alpha$) error of less than 5% (*p* < 0.05), was considered sufficient evidence for statistical significance in hypothesis testing.
+A probability of committing a type I ($\alpha$) error of less than 5% (*p* < 0.05) was considered sufficient evidence for statistical significance in hypothesis testing.
 All the statistical analyses were computed and implemented in the R programming language [@rlanguage].
-GAMs and the corresponding model estimates were computed using the *mgcv* and *modelbased* packages [@wood2017generalized; @dominique2020estimation].
+GAMs and the corresponding model estimates were calculated using the *mgcv* and *modelbased* packages [@wood2017generalized; @dominique2020estimation].
 Complementary R packages were used for visualization purposes [@hadley2016ggplot2; @daniel2021see].
@@
 
-Despite the fact that a similar non-linear effect was observed when inspecting the influence of corrected age in the FM domain scores (`r smooth["s(edad_corregida_meses)", expr]`), it was not possible to estimate a significant overall effect different from zero (`r overall`), nevertheless, it was only in the 22.3 to 38.3 age range where a significant and negative effect was observed (`r slopes[2, expr]`).
+Although a similar non-linear effect was observed when inspecting the influence of corrected age in the FM domain scores (`r smooth["s(edad_corregida_meses)", expr]`), it was not possible to estimate a significant overall effect different from zero (`r overall`), nevertheless, it was only in the 22.3 to 38.3 age range where a significant and negative effect was observed (`r slopes[2, expr]`).
 
@@
 
-Unlike the others, PS domain was not influenced by corrected age (`r smooth["s(edad_corregida_meses)", expr]`).
+Unlike the others, the PS domain was not influenced by corrected age (`r smooth["s(edad_corregida_meses)", expr]`).
 Accordingly, prematurity (measured in weeks) was not associated with any developmental domain within ASQ-3 assessment (significance for smooth terms: `r prematurity_smooths`).
@@
 
-This findings might be associated with the described motor impairments of hypotonia in the early stages of life, which compromise the infant's ability to explore and interact with his or her environment [@harris2008congenital; @gabis2021weak].
-A reflection of the aforementioned would be expressed in an altered development of GM function in the first months of life, with a consequent limitation in FM skills later on, which would have a subsequent negative impact on the communicative competence of infants, secondary to a reduced interaction with their environment and peers [@bodensteiner2008evaluation].
+These findings might be associated with the described motor impairments of hypotonia in the early stages of life, which compromise the infant's ability to explore and interact with their environment [@harris2008congenital; @gabis2021weak].
+A reflection of those mentioned above would be expressed in altered development of GM function in the first months of life, with a consequent limitation in FM skills later on, which would have a subsequent negative impact on the communicative competence of infants, secondary to reduced interaction with their environment and peers [@bodensteiner2008evaluation].
 In a recent systematic review [@gonzalez2019gross], differences in the predictive abilities of gross and FM skills on communication skills in infants and early childhood were reported, whereas GM skills, such as crawling and walking, favour exploration with their environment and caregivers, while FM skills, expressed through tasks such as drawing and handling utensils, could lead to improvements in language through mechanisms yet to be explored [@gonzalez2019gross].
@@
 It is worth noting that other context-mediated social factors may also have influenced our results, mainly due to the role that other variables would also play in the neurodevelopment of our study sample, which could have an impact on many of the developmental traits assessed here, such as intrauterine growth restrictions, maternal depression, institutionalisation, exposure to social violence, maternal education and breastfeeding [@walker2011inequality].
-All together, these represent the main limitations in our study design, which need to be addressed in future research, exploring the variations observed in different developmental traits in hypotonic infants.
-However, our study sheds light to an underexplored aspect of congenital myopathies, with robust statistical methods that made it possible to capture and model the complex relationships seen early in life.
+Altogether, these represent the main limitations in our study design, which need to be addressed in future research, exploring the variations observed in different developmental traits in hypotonic infants.
+However, our study sheds light on an underexplored aspect of congenital myopathies, with robust statistical methods that made it possible to capture and model the complex relationships seen early in life.
 
@@
 Moreover, we show that the observed variations in developmental domains are not solely attributable to prematurity, where age corrected for prematurity best explained the observed variability in neurodevelopment.
-Further research is warranted to determine how these findings apply when controlling for context-mediated social factors and in other populations.
+Further research is warranted to determine how these findings apply when controlling for context-mediated social factors and other populations.
 
@@
 
-The raw data supporting the conclusions of this article will be made available by the authors, without undue reservation.
+The raw data supporting the conclusions of this article will be made available by the authors without undue reservation.
 
@@
 
-All authors listed have made a substantial, direct and intellectual contribution to the work, and approved it for publication.
+All authors listed have made a substantial, direct and intellectual contribution to the work and approved it for publication.
 
@@
 
-The authors declare that the research was conducted in the absence of any commercial or financial relationships that could be construed as a potential conflict of interest.
+The authors declare that the research was conducted without any commercial or financial relationships that could be construed as a potential conflict of interest.
 

### 2022-11-06 manuscript/manuscript.Rmd
new plot function and updates on manuscript

@@
 bibliography: "misc/ref.bib"
-csl: "misc/apa.csl"
+csl: "misc/vancouver-dmcn.csl"
 link-citations: yes
+notes-after-punctuation: true
 output:
@@
     reference_docx: "misc/ref.docx"
+editor_options: 
+  chunk_output_type: console
 ---
@@
 ```{r load-packages}
-source(file = "manuscript/misc/setup.R")
+source(file = "misc/setup.R")
 ```
 
-**Title**: Neurodevelopmental features among infants with congenital hypotonia in high southern latitudes: An observational cross-sectional study.
+**Title**: Neurodevelopmental features among infants with congenital hypotonia: An observational cross-sectional study.
 
-**Authors**: Patricio Barría^1,2^, Matías Castillo-Aguilar^3,4^, Katherine Harris^3,4^, Cristian Núñez-Espinosa^4,5,^\*
+**Authors**: Patricio Barría^1,2^, Matías Castillo-Aguilar^3,4^, Katherine Harris^3,4^, Kay Gittermann^1,5^, Mónica Bontes^1^, Rolando Aguilar^1,6^, Carlos A.
+Cifuentes^7,8^, Marcela Munera^9^, Cristian Núñez-Espinosa^4,10^*
 
 ::: {custom-style="Afiliation"}
-^1^ Unidad de Investigación, Corporación de Rehabilitación Cruz del Sur, Punta Arenas, Chile.
+^1^ Corporación de Rehabilitación Club de Leones Cruz del Sur, Punta Arenas, Chile.
 
-^2^ Brain-Machine Interface Systems Lab, Universidad Miguel Hernández de Elche, Elche, España
+^2^ Brain-Machine Interface Systems Lab, Universidad Miguel Hernández de Elche, Elche, España.
 
@@
 
-^4^ Chilean Austral Integrative Neurophysiology Group ([NIM-ACh](https://nimach.org)), Centro Asistencial de Docencia e Investigación (CADI-UMAG), Punta Arenas Chile
+^4^ Centro Asistencial de Docencia e Investigación (CADI-UMAG), Punta Arenas Chile.
 
-^5^ School of Medicine, Magallanes University, Punta Arenas, Chile.
+^5^ Hospital Clínico de Magallanes, Punta Arenas, Chile.
+
+^6^ Engineering Department, Magallanes University, Punta Arenas, Chile.
+
+^7^ Bristol Robotics Laboratory, University of the West of England T Block, Frenchay Campus.
+
+^8^ School of Engineering, Science and Technology, Universidad del Rosario, Bogota, Colombia.
+
+^9^ Department of Biomedical Engineering, Colombian School of Engineering Julio Garavito.
+
+^10^ School of Medicine, Magallanes University, Punta Arenas, Chile.
 :::
@@
 
-**Objective**: To describe and model the relationship between sociodemographics, prematurity and neurodevelopmental levels based on the *Ages and Stages Questionnaire* (ASQ-3) scores in infants diagnosed with congenital hypotonia (CH).
-**Material and methods**: A total of 234 patients diagnosed with CH were enrolled in a rehabilitation programme led by the Rehabilitation Centre of Punta Arenas (Chile).
+**Aim**: To describe and model the relationship between sociodemographics, prematurity and neurodevelopmental levels based on the *Ages and Stages Questionnaire* (ASQ-3) scores in infants diagnosed with congenital hypotonia (CH).
+**Method**: A total of 234 patients diagnosed with CH participated in this study (median age, 8 years; 94 (40.2%) females and 140 (59.8%) males).
 Neurodevelopmental status was assessed with the ASQ-3 at admission, as well as sociodemographic and obstetric data obtained through the initial clinical interview.
@@
 Fine motor skills did not exhibit a linear relationship with corrected age but a non-linear effect, unlike the personal-social domain, which did not present significant variations across age corrected for prematurity.
-**Conclusion**: After adjusting for possible confounders, we found between-subjects fluctuations in neurodevelopmental traits across age in hypotonic infants.
+**Interpretation**: After adjusting for possible confounders, we found between-subjects fluctuations in neurodevelopmental traits across age in hypotonic infants.
 These fluctuations were present in the form of non-linear and domain-specific variations.
-Further research is warranted to determine how these findings apply in the presence of other context-mediated social factors and populations.
+
+What this paper adds:
+
+   - Describes levels of variations observed in neurodevelopmental traits are present across age in hypotonic infants.
+   - Observed variations in developmental domains are not solely attributable to prematurity, where age corrected for prematurity best explains the observed variability in neurodevelopment.
 
@@
 
-# Introduction
+<!--- # Introduction --->
 
-The physical and psychological signs of early childhood development are representative and relevant markers for the identification and monitoring of overall growth in early life [@di2016predictive].
-Therefore, both can be used in screening children at risk of developmental delay to support early referral and need further assessment to determine if they are eligible for early intervention services [@guralnick2017early; @bruder2010early].
+Early childhood development's physical and psychological signs are representative and relevant markers for identifying and monitoring overall growth in early life [@di2016predictive].
+Therefore, both can be used in screening children at risk of developmental delay to support early referral and identify the need for further assessment to determine if they are eligible for early intervention services [@bruder2010early; @guralnick2017early].
 
-Currently, many tools have been proposed to assess the developmental continuum of infants.
-In this sense, the Ages and Stages Questionnaire, Third Edition (ASQ-3) has been submitted as a global screening tool, parent/caregivers-oriented, that assesses five domains of development in children aged from 0 to 5.5 years old [@singh2017ages].
-Current evidence suggests that the ASQ-3 is an accurate, cost-effective, yet parent-friendly instrument for screening and monitoring children up to pre-school age and can help identify and exclude neurodevelopmental impairments in very preterm-born children [@kerstjens2015ages; @ballantyne2016risk; @singh2017ages].
+Currently, many tools have been proposed to assess the continued development of infants.
+In this sense, the Ages and Stages Questionnaire, Third Edition (ASQ-3), has been presented as a global screening tool aimed at parents/caregivers, assessing five developmental domains in children aged 0 to 5.5 years [@singh2017ages].
+Current evidence suggests that the ASQ-3 is an accurate, cost-effective, and parent-friendly instrument for screening and monitoring children through pre-school age, helping to identify and eliminate neurodevelopmental deficits in children born very premature [@kerstjens2015ages; @ballantyne2016risk; @singh2017ages].
 
-Multiple forms of neuromuscular, metabolic and genetic conditions are associated with hypotonia and may be a sign of neurodevelopmental delay, which may predispose to cognitive impairment in some cases [@riou2009global].
-Hypotonia has been defined as decreased muscle tone or floppiness, involving a wide range and levels of progression [@harris2008congenital; @gabis2021weak].
-Given that hypotonia and the hyperlaxity and motor delay associated with it may impair the infant's capacity to engage with its surroundings, critical visual cues may be ignored, potentially leading both to an impairment of learning and cognitive development [@harris2008congenital], hence the need to explore the neurodevelopmental attributes of infants with hypotonia.
+Multiple neuromuscular, metabolic, and genetic conditions have been associated with hypotonia.
+Hypotonia is the decreased muscle tone or flaccidity and may represent a sign of delayed neurological development in the child, which may predispose to cognitive impairment in some cases [@riou2009global].
+Because the baby's growth-associated hypotonia, hypermobility, and motor delay can affect her ability to relate to her environment, critical visual cues are not always interpreted, leading to impaired learning and cognitive development [@harris2008congenital].
+Hypotonia implies a wide range and levels of muscle flaccidity, which should be explored in the cognitive development of the infant [@gabis2021weak; @harris2008congenital].
+There are multiple causes of hypotonia, one of them being congenital hypotonia (CH).
+Some authors suggest that CH cannot be referred to as a diagnosis [@thompson2002benign].
+Those who do consider it point it out as a diagnosis of exclusion in the investigation, considering it only when it is established in the absence of other signs and symptoms [@gabis2021weak; @leyenaar2005schematic].
+However, its research is fundamental since it is a non-progressive neuromuscular disorder, which tends to improve with time and early intervention [@gabis2021weak].
 
-The congenital hypotonia (CH) is usually considered an exclusion diagnosis, and is generally made in the absence of other signs and symptoms, after every other evaluative resource has been exhausted [@leyenaar2005schematic; @gabis2021weak].
-However, this is not true for at least one author that suggests that CH can not be considered a diagnosis as such [@thompson2002benign].
-CH is a non-progressive neuromuscular disorder but tends to improve with time and early intervention [@gabis2021weak].
-
-Nevertheless, to the best of our knowledge, there is no robust evidence characterising the observed variation in developmental traits in infants with diagnosed CH across ages.
-Therefore, our main objective in this study was to describe and model the relationship between sociodemographics, prematurity and neurodevelopmental levels based on ASQ-3 scores in infants with diagnosed CH.
+To our knowledge, there is no substantial evidence characterizing the observed variation in developmental traits in infants with diagnosed CH across ages.
+Therefore, our primary goal in this study was to describe and model the relationship between sociodemographic characteristics, prematurity, and neurodevelopmental levels as a function of ASQ-3 estimates in infants diagnosed with CH.
 
 # Material y methods
+
+## Ethics
+
+Participating subjects gave their permission, and legal guardians provide informed consent before participation.
+The Ethics Committee approved this study of the Corporación de Rehabilitación Club de Leones Cruz del Sur, Chile (code: CorporacionRCLCS010322), following the regulations established by the Declaration of Helsinki on ethical principles in human beings.
 
@@
 
-```{r}
+```{r, eval=FALSE}
 # Sample
-n_sample <- dataset[, .N]
+n_sample <- dataset[, .N]; 
+print(n_sample);
 #> 234
@@
 
-A total of `r n_sample` patients were enrolled as part of an intervention programme carried out by the *Cruz del Sur* Rehabilitation Centre (Punta Arenas, Chile), admitted from one month to 60 months, and assessed at admission, control and discharge.
-The total records of patients diagnosed with CH admitted to the institution's programme were analysed, and the assessment of the subjects was conducted and guided by a nurse trained in the application of the ASQ-3.
+A total of 234 patients were enrolled as part of an intervention program carried out by the Cruz del Sur Rehabilitation Center (Punta Arenas, Chile), admitted from one month to 60 months old, and evaluated at admission, control and discharge.
+Participating patients were diagnosed based on clinical criteria confirmed by a neurologist and a physiatrist.
+The total files of the patients diagnosed with CH, and admitted to the institution's program, were analyzed, and the evaluation of the subjects was carried out and guided by a nurse trained in the application of the ASQ-3.
 
@@
 
-The diagnosis of CH is based on four sources: 1) primary care paediatrician, who refers the diagnosis to the institution where admission to the programme is made; 2) paediatricians from private clinics, who refer the diagnosis; 3) neuropaediatricians from the clinical hospital, who refer the diagnosis to the programme; 4) physiatrist from the institution, who assigns the diagnosis if it corresponds to the semiology.
+The diagnosis of CH is based on four sources: 1) primary care pediatrician, who refers the diagnosis to the institution where admission to the program is made; 2) pediatricians from private clinics, who refer the diagnosis; 3) neuropediatricians from the clinical hospital, who refer the diagnosis to the program; 4) physiatrist from the institution, who assigns the diagnosis if it corresponds to the semiology.
 
@@
 The ASQ-3 is a parent-reported initial level developmental screening instrument consisting of 21 intervals, each with 30 items in five areas: i) communication (CM), ii) gross motor (GM), iii) fine motor (FM), iv) problem-solving (CG), and v) personal-social (PS) [@squires2009ages].
-The ASQ is cost-effective and widely used in the United States and other countries [@sarmiento2011universal; @heo2008cross].
-It has been translated into several languages, and international studies on its psychometric properties in diverse cultural environments are increasing [@sarmiento2011universal; @heo2008cross].
+The ASQ is cost-effective and widely used in the United States and other countries [@heo2008cross; @sarmiento2011universal].
+It has been translated into several languages, and international studies on its psychometric properties in diverse cultural environments are increasing [@heo2008cross; @sarmiento2011universal].
 It has shown good psychometric properties (75% sensitivity and 81% specificity) in Chilean term and preterm infants [@schonhaut2013validity].
@@
 
-For the collection of research data, authorisation consent was obtained for the use of instrumental clinical data as well as the clinical record of each patient.
-Subsequently, and following national research regulations, written authorisation was obtained from the institutional director to use the forms and database for research purposes.
-To this end, the data were anonymised during data processing and subsequent analyses.
+For the collection of research data, authorization consent was obtained for the use of instrumental clinical data as well as the clinical record of each patient.
+Subsequently, and following national research regulations, written authorization was obtained from the institutional director to use the forms and database for research purposes.
+To this end, the data were anonymized during data processing and subsequent analyses.
 
@@
 
-Each patient's demographic data were collected and made available throughout the study by the institutional electronic systems at the time of entry to the programme.
+Each patient's demographic data were collected and made available throughout the study by the institutional electronic systems at the time of entry to the program.
 The administrative registration was done by the secretaries of the user coordination unit, recording name, ID number, date of birth and diagnosis.
-In addition, age data was automatically updated by the computer system's algorithms.
+In addition, age data was automatically updated by the computer system’s algorithms.
 
@@
 The ASQ-3 was applied according to the protocol established by the instrument itself, with a face-to-face or telematic application being valid.
-The protocol can be found in the instrument's manuals, which the programme nurse administered.
-Some important considerations of the administration are that i) the primary caregiver must respond, ii) in case of doubts of the caregiver about the assessed behaviour, the information is corroborated by in situ tests with the user, and iii) in relation to the correction, the test itself standardises these procedures.
+The protocol can be found in the instrument’s manuals, which the program nurse administered.
+Some important considerations of the administration are that i) the primary caregiver must answer, ii) in case of doubts of the caregiver about the assessed behavior, the information is corroborated by in situ tests with the user, and iii) in relation to the correction, the test itself standardizes these procedures.
 
@@
 Data are presented as median (*Mdn*) and interquartile range (*IQR*) for continuous variables; for categorical/discrete variables, the absolute and relative sample size was reported.
+There was no missing data.
 
-A non-parametric approach was used since the underlying distribution of measured outcomes, assessed through analytical and graphical methods, did not follow a Gaussian distribution.
+A non-parametric approach was used since the underlying distribution of continuously measured outcomes, assessed through analytical and graphical methods, did not follow a Gaussian distribution.
 
-In order to assess the differences in developmental scores between males and females, the *Wilcoxon* rank-sum test was used, meanwhile the chi-square test ($\chi^2$) was used to evaluate goodness-of-fit ($\chi^2_{gof}$) and independence of factors ($\chi^2_{Pearson}$).
+In order to assess the differences in developmental scores between males and females, the *Wilcoxon* rank-sum test was used, meanwhile the chi-square test ($\chi^2$) was used to evaluate goodness-of-fit ($\chi^2_{gof}$).
 
@@
 The restricted maximum likelihood method was used to estimate the smoothing parameters, and thin-plate regression splines as the smoothing basis, as they are the optimal smoother of any given basis dimension/rank [@wood2003thin].
-In the final models, infants' sex, clinician and infants' relationship with caregivers were added as random effects in the form of penalized parametric terms to account for the variability arising from these variables in the fixed effects analysed [@wood2016smoothing].
-We used approximative derivatives with 95% confidence intervals (CI95%) to describe the smooth terms by means of quasi-linear segments.
+We used approximative derivatives with 95% confidence intervals (CI95%) to describe the smooth terms by means of quasi-linear segments [@dominique2020estimation].
 
-A probability of committing a type I ($\alpha$) error of less than 5% (*p* < 0.05) was considered sufficient evidence for statistical significance in hypothesis testing.
+In order to assess the influence of confounding variables, we fitted different models using the variables: i) sex of the infant, ii) clinician ID and iii) the relationship of the infant to the respondent (i.e., whether the respondent was the infant's mother, father, uncle or grandparent) as random effects in the form of penalized parametric terms [@wood2016smoothing].
+Varying random effects structures models can be seen in the *random effects*' section of the supplementary file.
+
+A probability of committing a type I error ($\alpha$) of less than 5% (*p* < 0.05) was considered sufficient evidence for statistical significance in hypothesis testing.
 All the statistical analyses were computed and implemented in the R programming language [@rlanguage].
-GAMs and the corresponding model estimates were calculated using the *mgcv* and *modelbased* packages [@wood2017generalized; @dominique2020estimation].
-Complementary R packages were used for visualization purposes [@hadley2016ggplot2; @daniel2021see].
+GAMs and the corresponding model estimates were calculated using the *mgcv* and *modelbased* packages [@dominique2020estimation; @wood2017generalized].
+Complementary R packages were used for visualization purposes [@daniel2021see; @hadley2016ggplot2].
 
@@
 
-```{r}
+```{r, eval=FALSE}
 # Sample by sex
 n_sex <- dataset[j = .(stat = paste0(.N, " (", round(.N*100/n_sample, 1), "%)")), 
-                 keyby = .(sex = sexo_paciente)]
+                 keyby = .(sex = sexo_paciente)];
+print(n_sex);
 #>    sex        stat
@@
   chisq.test() |> 
-  report_chi(effsize = FALSE)
-#> [1] "$\\chi^2_{gof}$ (1) = 9.04, *p* = 0.003"
+  report_chi(effsize = FALSE);
+cat(prop_sex);
+#> $\chi^2_{gof}$ (1) = 9.04, *p* = 0.003
 ```
 
-From a total of `r n_sample` subjects with congenital hypotonia, `r n_sex["F", stat]` were females and `r n_sex["M", stat]` males (`r prop_sex`).
+From a total of 234 subjects with congenital hypotonia, 94 (40.2%) were females and 140 (59.8%) males ($\chi^2_{gof}$ (1) = 9.04, *p* = 0.003).
 The developmental characteristics of the sample can be seen in [Table 1](#tab1).
 
-```{r}
+When modelling the effect of chronological age on developmental domains, different random effects structures were compared when assessing model estimates, these estimates can be seen in [Figure 1](#fig1).
+
+```{r, eval=FALSE}
 set.seed(123)
@@
   comunicacion_total ~ 
-    s(profesional_id, sexo_paciente, respondedor_vinculo, bs = "re") + 
+    s(profesional_id, bs = "re") + 
+    s(sexo_paciente, bs = "re") + 
+    s(respondedor_vinculo, bs = "re") + 
     s(edad_corregida_meses), 
@@
 
-overall <- report_overall(mod, "edad_corregida_meses")
-#> [1] "$\\beta$ = -2.36, CI~95%~[-3.47, -1.25], $t_{student}$ (224.04) = -4.2, *p* < 0.001"
-
-slopes <- report_slopes(mod, "edad_corregida_meses")
-#>    Start  End edad_corregida_meses                                                                                 expr
-#> 1:   0.0  6.8                  3.4    $\\beta$ = 0.49, CI~95%~[-0.89, 1.86], $t_{student}$ (224.04) = 0.45, *p* = 0.319
-#> 2:   7.3 17.9                 12.6 $\\beta$ = -2.08, CI~95%~[-3.17, -0.99], $t_{student}$ (224.04) = -3.79, *p* = 0.004
-#> 3:  18.4 48.0                 33.2    $\\beta$ = 0.45, CI~95%~[-1.32, 2.23], $t_{student}$ (224.04) = 0.42, *p* = 0.593
-
-smooth <- report_smooth(mod)
-#>                                                   term                                            expr
-#> 1:                             s(edad_corregida_meses) $F_{smooth}$ (5.2, 224.04) = 13.43, *p* < 0.001
-#> 2: s(profesional_id,sexo_paciente,respondedor_vinculo)  $F_{smooth}$ (3.76, 224.04) = 0.72, *p* = 0.02
+report_smooth(mod);
+report_overall(mod, "edad_corregida_meses") |> cat();
+report_slopes(mod, "edad_corregida_meses");
 ```
 
-When modelling the effect of chronological age on developmental domains, corrected for prematurity, we observed a significant non-linear relationship on CM scores (`r smooth["s(edad_corregida_meses)", expr]`), that reflect an overall negative marginal effect (`r overall`), however, this was not true when assessing the direction of the effect in the age range between 0 to 6.8 (`r slopes[1, expr]`), neither in the 18.4 to 48 months old group (`r slopes[3, expr]`), whereas the effect tend to be positive but non-significant.
-The relationship between developmental domains, corrected age and their effect derivatives can be seen in [Figure 1](#fig1).
+After fitting a GAM to assess the effect of corrected age on CM scores, we observed a significant non-linear relationship between them ($F_{smooth}$ (5.1, 224.74) = 13.88, *p* < 0.001), that reflect an overall negative marginal trend ($\beta$ = -2.39, CI~95%~[-3.46, -1.32], $t_{student}$ (224.74) = -4.4, *p* < 0.001), however, this was not true when assessing the direction of the effect in the age range between 0 to 6.8 ($\beta$ = 0.38, CI~95%~[-0.96, 1.72], $t_{student}$ (224.74) = 0.33, *p* = 0.351), neither in the 18.4 to 48 months old group ($\beta$ = 0.44, CI~95%~[-1.27, 2.16], $t_{student}$ (224.74) = 0.42, *p* = 0.577), whereas the effect tend to be positive but non-significant.
+The relationship between developmental domains, corrected age and their effect derivatives can be seen in [Figure 2](#fig2).
 
-```{r}
+```{r, eval=FALSE}
 set.seed(123)
@@
   motora_gruesa_total ~ 
-    s(profesional_id, sexo_paciente, respondedor_vinculo, bs = "re") + 
+    s(profesional_id, bs = "re") + 
+    s(sexo_paciente, bs = "re") + 
+    s(respondedor_vinculo, bs = "re") +
     s(edad_corregida_meses), 
@@
 
-overall <- report_overall(mod, "edad_corregida_meses")
-#> [1] "$\\beta$ = 1.95, CI~95%~[0.66, 3.25], $t_{student}$ (226.75) = 2.97, *p* = 0.003"
-
-slopes <- report_slopes(mod, "edad_corregida_meses")
-#>    Start  End edad_corregida_meses                                                                                expr
-#> 1:   0.0  6.8                  3.4 $\\beta$ = -2.94, CI~95%~[-4.55, -1.34], $t_{student}$ (226.75) = -3.7, *p* = 0.004
-#> 2:   7.3  9.2                  8.2    $\\beta$ = 0.02, CI~95%~[-1.12, 1.17], $t_{student}$ (226.75) = 0.06, *p* = 0.45
-#> 3:   9.7 15.5                 12.6    $\\beta$ = 1.86, CI~95%~[0.61, 3.11], $t_{student}$ (226.75) = 2.93, *p* = 0.009
-#> 4:  16.0 48.0                 32.0  $\\beta$ = -0.02, CI~95%~[-2.05, 2.01], $t_{student}$ (226.75) = 0.07, *p* = 0.646
-
-smooth <- report_smooth(mod)
-#>                                                   term                                            expr
-#> 1:                             s(edad_corregida_meses) $F_{smooth}$ (5.24, 226.75) = 6.19, *p* < 0.001
-#> 2: s(profesional_id,sexo_paciente,respondedor_vinculo)  $F_{smooth}$ (1.01, 226.75) = 0.1, *p* = 0.244
+report_smooth(mod)
+report_overall(mod, "edad_corregida_meses") |> cat();
+report_slopes(mod, "edad_corregida_meses")
 ```
 
-When analysing the motor skills domain, we found a significant non-linear effect of corrected age on GM scores, `r smooth["s(edad_corregida_meses)", expr]`, which had an overall positive effect (`r overall`), however, the slope varied as a function of age, with a negative effect in the 0 to 6.8 age range (`r slopes[1, expr]`), but in the 9.7 to 15.5 interval, this relationship was inverted (`r slopes[3, expr]`), however, in the rest of the age range the slope was non-significant and virtually zero (Age~[7.3, 9.2]~, `r slopes[2, expr]`; Age~[16, 48]~, `r slopes[4, expr]`).
+When analyzing the motor skills domain, we found a significant non-linear effect of corrected age on GM scores, $F_{smooth}$ (5.25, 226.82) = 5.6, *p* < 0.001, which had an overall positive effect ($\beta$ = 1.86, CI~95%~[0.57, 3.16], $t_{student}$ (226.82) = 2.83, *p* = 0.005), however, the slope varied as a function of age, with a negative effect in the 0 to 6.8 age range ($\beta$ = -3.02, CI~95%~[-4.63, -1.42], $t_{student}$ (226.82) = -3.81, *p* = 0.003), but in the 10.2 to 15.5 interval, this relationship was inverted ($\beta$ = 1.83, CI~95%~[0.57, 3.08], $t_{student}$ (226.82) = 2.87, *p* = 0.008), however, in the rest of the age range the slope was non-significant and virtually zero (Age~[7.3, 9.7]~, $\beta$ = 0.13, CI~95%~[-1.02, 1.27], $t_{student}$ (226.82) = 0.23, *p* = 0.387; Age~[16, 48]~, $\beta$ = -0.06, CI~95%~[-2.09, 1.97], $t_{student}$ (226.82) = 0.01, *p* = 0.701).
 
-```{r}
+```{r, eval=FALSE}
 set.seed(123)
@@
   motora_fina_total ~ 
-    s(profesional_id, sexo_paciente, respondedor_vinculo, bs = "re") + 
+    s(profesional_id, bs = "re") + 
+    s(sexo_paciente, bs = "re") + 
+    s(respondedor_vinculo, bs = "re") +
     s(edad_corregida_meses), 
@@
 
-overall <- report_overall(mod, "edad_corregida_meses")
-#> [1] "$\\beta$ = 0.04, CI~95%~[-0.45, 0.52], $t_{student}$ (226.77) = 0.14, *p* = 0.886"
-
-slopes <- report_slopes(mod, "edad_corregida_meses")
-#>    Start  End edad_corregida_meses                                                                                 expr
-#> 1:   0.0 21.8                 10.9    $\\beta$ = 0.03, CI~95%~[-0.54, 0.61], $t_{student}$ (226.77) = 0.04, *p* = 0.434
-#> 2:  22.3 38.3                 30.3 $\\beta$ = -0.79, CI~95%~[-1.45, -0.12], $t_{student}$ (226.77) = -2.34, *p* = 0.022
-#> 3:  38.8 48.0                 43.4  $\\beta$ = -0.88, CI~95%~[-1.97, 0.21], $t_{student}$ (226.77) = -1.62, *p* = 0.113
-
-smooth <- report_smooth(mod)
-#>                                                   term                                           expr
-#> 1:                             s(edad_corregida_meses) $F_{smooth}$ (2.59, 226.77) = 4.2, *p* = 0.005
-#> 2: s(profesional_id,sexo_paciente,respondedor_vinculo) $F_{smooth}$ (3.63, 226.77) = 0.7, *p* = 0.021
+report_smooth(mod)
+report_overall(mod, "edad_corregida_meses");
+report_slopes(mod, "edad_corregida_meses");
 ```
 
-Although a similar non-linear effect was observed when inspecting the influence of corrected age in the FM domain scores (`r smooth["s(edad_corregida_meses)", expr]`), it was not possible to estimate a significant overall effect different from zero (`r overall`), nevertheless, it was only in the 22.3 to 38.3 age range where a significant and negative effect was observed (`r slopes[2, expr]`).
+Although a similar non-linear effect was observed when inspecting the influence of corrected age in the FM domain scores ($F_{smooth}$ (2.56, 228.03) = 4.11, *p* = 0.006), it was not possible to estimate a significant overall effect different from zero ($\beta$ = 0.01, CI~95%~[-0.47, 0.49], $t_{student}$ (228.03) = 0.03, *p* = 0.977), nevertheless, it was only in the 22.3 to 37.3 age range where a significant and negative effect was observed ($\beta$ = -0.75, CI~95%~[-1.4, -0.11], $t_{student}$ (228.03) = -2.32, *p* = 0.023).
 
-```{r}
+```{r, eval=FALSE}
 set.seed(123)
@@
   resolucion_problemas_total ~ 
-    s(profesional_id, sexo_paciente, respondedor_vinculo, bs = "re") + 
+    s(profesional_id, bs = "re") + 
+    s(sexo_paciente, bs = "re") + 
+    s(respondedor_vinculo, bs = "re") +
     s(edad_corregida_meses), 
@@
 
-overall <- report_overall(mod, "edad_corregida_meses")
-#> [1] "$\\beta$ = -1.87, CI~95%~[-3.17, -0.57], $t_{student}$ (227.01) = -2.83, *p* = 0.005"
-
-slopes <- report_slopes(mod, "edad_corregida_meses")
-#>    Start  End edad_corregida_meses                                                                                 expr
-#> 1:   0.0  5.8                  2.9     $\\beta$ = 2.81, CI~95%~[1.18, 4.44], $t_{student}$ (227.01) = 3.49, *p* = 0.002
-#> 2:   6.3  8.7                  7.5      $\\beta$ = 0.05, CI~95%~[-1.1, 1.2], $t_{student}$ (227.01) = 0.06, *p* = 0.395
-#> 3:   9.2 14.1                 11.6 $\\beta$ = -1.59, CI~95%~[-2.82, -0.37], $t_{student}$ (227.01) = -2.55, *p* = 0.015
-#> 4:  14.5 48.0                 31.3    $\\beta$ = 0.03, CI~95%~[-1.99, 2.04], $t_{student}$ (227.01) = -0.06, *p* = 0.55
-
-smooth <- report_smooth(mod)
-#>                                                   term                                            expr
-#> 1:                             s(edad_corregida_meses) $F_{smooth}$ (5.66, 227.01) = 3.65, *p* = 0.001
-#> 2: s(profesional_id,sexo_paciente,respondedor_vinculo) $F_{smooth}$ (0.34, 227.01) = 0.03, *p* = 0.339
+report_smooth(mod);
+report_overall(mod, "edad_corregida_meses");
+report_slopes(mod, "edad_corregida_meses");
 ```
 
-CG abilities were significantly influenced by corrected age (`r smooth["s(edad_corregida_meses)", expr]`), with an overall negative effect (`r overall`), and just like the other domains, this relationship was modified across corrected age.
-In this sense, from the 0 to 5.8 age interval, we found that for every increase in one month in corrected age, we can expect a proportional increase in 2.81 points (`r slopes[1, expr]`) in the CG domain, while in the age range 9.2 to 14.1 the relationship changes inversely, mainly because in this age range we observe that for every one-month increase in the corrected age, a decrease of 1.59 points could be expected in the same domain (`r slopes[3, expr]`).
-The other age intervals did not have a slope that deviated significantly from zero (Age~[6.3, 8.7]~, `r slopes[2, expr]`; Age~[14.5, 48.0]~, `r slopes[4, expr]`).
+CG abilities were significantly influenced by corrected age ($F_{smooth}$ (5.69, 225.93) = 3.7, *p* = 0.001), with an overall negative effect ($\beta$ = -1.93, CI~95%~[-3.24, -0.63], $t_{student}$ (225.93) = -2.91, *p* = 0.004), and just like the other domains, this relationship was modified across the corrected age.
+In this sense, from the 0 to 5.8 age interval, we found that for every increase in one month in corrected age, we can expect a proportional increase in 2.8 points ($\beta$ = 2.8, CI~95%~[1.17, 4.43], $t_{student}$ (225.93) = 3.48, *p* = 0.002) in the CG domain, while in the age range 9.2 to 14.1 the relationship changes inversely, mainly because in this age range we observe that for every one-month increase in the corrected age, a decrease of 1.64 points could be expected in the same domain ($\beta$ = -1.64, CI~95%~[-2.86, -0.41], $t_{student}$ (225.93) = -2.62, *p* = 0.014).
+The other age intervals did not have a slope that deviated significantly from zero (Age~[6.3, 8.7]~, $\beta$ = 0.06, CI~95%~[-1.09, 1.2], $t_{student}$ (225.93) = 0.07, *p* = 0.388; Age~[14.5, 48.0]~, $\beta$ = 0.04, CI~95%~[-1.98, 2.07], $t_{student}$ (225.93) = -0.05, *p* = 0.564).
 
-```{r}
+```{r, eval=FALSE}
 set.seed(123)
@@
   socio_individual_total ~ 
-    s(profesional_id, sexo_paciente, respondedor_vinculo, bs = "re") + 
+    s(profesional_id, bs = "re") + 
+    s(sexo_paciente, bs = "re") + 
+    s(respondedor_vinculo, bs = "re") +
     s(edad_corregida_meses), 
@@
 
-smooth <- report_smooth(mod)
-#>                                                   term                                            expr
-#> 1:                             s(edad_corregida_meses)    $F_{smooth}$ (1, 231.58) = 1.16, *p* = 0.282
-#> 2: s(profesional_id,sexo_paciente,respondedor_vinculo) $F_{smooth}$ (0.42, 231.58) = 0.03, *p* = 0.338
+report_smooth(mod);
 ```
 
-```{r}
+```{r, eval=FALSE}
 y_vars <- c("CM" = "comunicacion_total"
@@
     paste0(i, " ~ ", 
-           "s(profesional_id, sexo_paciente, respondedor_vinculo, bs = \"re\") +",
+           "s(profesional_id, bs = \"re\") + 
+            s(sexo_paciente, bs = \"re\") + 
+            s(respondedor_vinculo, bs = \"re\") +",
            "s(semanas_prematurez, k = 9)")
@@
               collapse = "; "
-            )]
-#> [1] "CM, *p* = 0.715; FM, *p* = 0.987; GM, *p* = 0.357; CG, *p* = 0.292; personal-individual, *p* = 0.131"
+            )];
+cat(prematurity_smooths)
 ```
 
-Unlike the others, the PS domain was not influenced by corrected age (`r smooth["s(edad_corregida_meses)", expr]`).
-Accordingly, prematurity (measured in weeks) was not associated with any developmental domain within ASQ-3 assessment (significance for smooth terms: `r prematurity_smooths`).
+Unlike the others, the PS domain was not influenced by corrected age ($F_{smooth}$ (1, 230.4) = 1.06, *p* = 0.304).
+Accordingly, prematurity (measured in weeks) was not associated with any developmental domain within ASQ-3 assessment (significance for smooth terms: CM, *p* = 0.639; FM, *p* = 0.918; GM, *p* = 0.33; CG, *p* = 0.323; PS, *p* = 0.152).
 
@@
 Our main findings suggest a non-linear effect of age, corrected for prematurity, with a marked decrease in scores for all neurodevelopmental traits at different age frames, even after adjusting for caregiver relationship, sex and inter-rater influence.
-However in the PS domain, there was no variation observed across corrected age.
+However no variation was observed across the corrected age in the PS domain.
 
-These findings might be associated with the described motor impairments of hypotonia in the early stages of life, which compromise the infant's ability to explore and interact with their environment [@harris2008congenital; @gabis2021weak].
+These findings confirm the described motor impairments of hypotonia in the early stages of life, which compromise the infant’s ability to explore and interact with their environment [@gabis2021weak; @harris2008congenital].
 A reflection of those mentioned above would be expressed in altered development of GM function in the first months of life, with a consequent limitation in FM skills later on, which would have a subsequent negative impact on the communicative competence of infants, secondary to reduced interaction with their environment and peers [@bodensteiner2008evaluation].
-In a recent systematic review [@gonzalez2019gross], differences in the predictive abilities of gross and FM skills on communication skills in infants and early childhood were reported, whereas GM skills, such as crawling and walking, favour exploration with their environment and caregivers, while FM skills, expressed through tasks such as drawing and handling utensils, could lead to improvements in language through mechanisms yet to be explored [@gonzalez2019gross].
-These milestones may be impaired in the face of poorer head and trunk control in CH, which have been shown to delay the achievement of key motor milestones in infants [@harris2008congenital; @bodensteiner2008evaluation; @gabis2021weak].
+In a recent systematic review [@gonzalez2019gross], differences in the predictive abilities of gross and FM skills on communication skills in infants and early childhood were reported.
+Whereas GM skills, such as crawling and walking, favor exploration with their environment and caregivers, while FM skills, expressed through tasks such as drawing and handling utensils, could lead to improvements in language through mechanisms yet to be explored [@gonzalez2019gross].
+These milestones may be impaired in the face of poorer head and trunk control in CH, which have been shown to delay the achievement of crucial motor milestones in infants [@bodensteiner2008evaluation; @gabis2021weak; @harris2008congenital].
 
-It is worth noting that other context-mediated social factors may also have influenced our results, mainly due to the role that other variables would also play in the neurodevelopment of our study sample, which could have an impact on many of the developmental traits assessed here, such as intrauterine growth restrictions, maternal depression, institutionalisation, exposure to social violence, maternal education and breastfeeding [@walker2011inequality].
+It is worth noting that other context-mediated social factors may also have influenced our results, mainly due to the role that other variables would also play in the neurodevelopment of our study sample, which could have an impact on many of the developmental traits assessed here, such as intrauterine growth restrictions, maternal depression, institutionalization, exposure to social violence, maternal education and breastfeeding [@walker2011inequality].
 Altogether, these represent the main limitations in our study design, which need to be addressed in future research, exploring the variations observed in different developmental traits in hypotonic infants.
-However, our study sheds light on an underexplored aspect of congenital myopathies, with robust statistical methods that made it possible to capture and model the complex relationships seen early in life.
+However, our study sheds light on an underexplored aspect with robust statistical methods that enabled us to capture and model the complex relationships seen early in life.
 
@@
 The present study shows that the marked variations observed in neurodevelopmental traits are present across age in hypotonic infants, mainly in the form of non-linear and domain-specific variations, even after adjusting for the effect that caregiver relationship, sex and evaluators might exert.
-Moreover, we show that the observed variations in developmental domains are not solely attributable to prematurity, where age corrected for prematurity best explained the observed variability in neurodevelopment.
-Further research is warranted to determine how these findings apply when controlling for context-mediated social factors and other populations.
+Moreover, we show that the observed variations in developmental domains are not solely attributable to prematurity, where age corrected for prematurity best explains the observed variability in neurodevelopment.
+Future research may determine how these findings apply when controlling for context-mediated social factors and other clinical populations.
 
@@
 
-The authors declare that the research was conducted without any commercial or financial relationships that could be construed as a potential conflict of interest.
+The authors declare that the research was conducted without any commercial or financial relationships construed as a potential conflict of interest.
+
+# ORCID
+
+::: {custom-style="Afiliation"}
+Patricio Barría, 0000-0003-4088-9820
+
+Matías Castillo-Aguilar, 0000-0001-7291-247X
+
+Katherine Harris, 0000-0003-4271-6405
+
+Kay Gittermann, 0000-0002-8042-844X
+
+Mónica Bontes, 0000-0002-0592-7361
+
+Rolando Aguilar, 0000-0002-0104-3032
+
+Carlos Cifuentes, 0000-0002-6942-865X
+
+Marcela Munera, 0000-0001-6595-5383
+
+Cristian Núñez-Espinosa, 0000-0002-9896-7062
+:::
 
@@
 
+\{Add table\}
+
+\newpage
+
 [**Table 1**]{#tab1}.
@@
 
-[**Figure 1**]{#fig1}.
+[**Figure 1**]{#fig1}. Predicted scores on developmental domains using corrected age as an independent variable, comparing different random effects structures on the estimated score.
+
+[**Figure 2**]{#fig2}.
 Relationship between corrected age (in months) and developmental domains.

### 2022-11-06 manuscript/supplementary.Rmd
new plot function and updates on manuscript

---
title: "Supplementary file"
bibliography: "misc/ref.bib"
csl: "misc/vancouver-dmcn.csl"
link-citations: yes
notes-after-punctuation: true
output:
  word_document:
    reference_docx: "misc/ref.docx"
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = FALSE,
  include = FALSE
)
```

```{r load-packages}
source(file = "misc/setup.R")
```

**Title**: Neurodevelopmental features among infants with congenital hypotonia: An observational cross-sectional study.

**Authors**: Patricio Barría^1,2^, Matías Castillo-Aguilar^3,4^, Katherine Harris^3,4^, Kay Gittermann^1,5^, Mónica Bontes^1^, Rolando Aguilar^1,6^, Carlos A. Cifuentes^7,8^, Marcela Munera^9^, Cristian Núñez-Espinosa^4,10^*

::: {custom-style="Afiliation"}
^1^ Corporación de Rehabilitación Club de Leones Cruz del Sur, Punta Arenas, Chile.

^2^ Brain-Machine Interface Systems Lab, Universidad Miguel Hernández de Elche, Elche, España.

^3^ Kinesiology Department, Magallanes University, Punta Arenas, Chile.

^4^ Centro Asistencial de Docencia e Investigación (CADI-UMAG), Punta Arenas Chile.

^5^ Hospital Clínico de Magallanes, Punta Arenas, Chile.

^6^ Engineering Department, Magallanes University, Punta Arenas, Chile.

^7^ Bristol Robotics Laboratory, University of the West of England T Block, Frenchay Campus.

^8^ School of Engineering, Science and Technology, Universidad del Rosario, Bogota, Colombia.

^9^ Department of Biomedical Engineering, Colombian School of Engineering Julio Garavito.

^10^ School of Medicine, Magallanes University, Punta Arenas, Chile.
:::

### \*Autor de correspondencia

Cristian Núñez-Espinosa, School of Medicine, Magallanes University, Punta Arenas, Chile. Centro Asistencial de Docencia e Investigación CADI-UMAG, Chile. e-mail: [cristian.nunez\@umag.cl](mailto:cristian.nunez@umag.cl). Address: Avenida Bulnes 01855, Box 113-D. Phone: +56 61 2201411

\newpage

```{r}
params <- grep("total$", names(dataset), value = TRUE)
names(params) <- c("CM", "GM", "FM", "CG", "PS")

models <- lapply(params, gam_confounders, data = dataset, plot = FALSE);

comp <- rbindlist(
  l = lapply(models, function(x) {
    rbindlist(lapply(x, function(i) {
      j <- performance::performance(i);
      as.data.table(j)[, lapply(.SD, round, digits = 2), .SDcols = is.numeric]
    }), idcol = "Model")
  }), idcol = "Domain"
)

comp[, ` ` := ifelse(AIC == min(AIC),"a", ""), list(Domain)][]
comp[, ` ` := ifelse(BIC == min(BIC), paste0(` `,"b"), ` `), list(Domain)][]
comp[, ` ` := ifelse(R2 == max(R2), paste0(` `,"c"), ` `), list(Domain)][]
comp[, ` ` := ifelse(Sigma == min(Sigma), paste0(` `,"d"), ` `), list(Domain)][]
comp[, ` ` := ifelse(RMSE == min(RMSE), paste0(` `,"e"), ` `), list(Domain)][]
```

## Assessing Random Effects Structures

In order to asses which random effect structure would be more appropriate we fitted multiple gerenalized additive models (GAMs) using each developmental domain as the response and the corrected age (in months) as the independent fixed effect.

The following variables: i) sex of the infant, ii) clinician ID, iii) the relationship of the infant to the respondent (i.e., whether the respondent was the infant's mother, father, uncle or grandparent), were treated as random effects in the form of penalized parametric terms [@wood2016smoothing].

The models fitted for each developmental domain were the following:

- **Simple model**: No random effects.
- **Only sex**: Considering sex within the model.
- **Clinician**: Considering the clinician identification.
- **Respondent**: Considering the relationship status (mother, father, uncle or grandparent) with the infant.
- **Clinician + Respondent**: Considering both clinician ID and respondent status.
- **Sex + Respondent**: Considering the infants' sex and respondent status.
- **Clinician + Sex**: Considering the clinician ID and infants' sex.
- **Full model**: Considering all of above variables as random effects within the model.

The performance of the aforementioned models can be seen in the Table 1S.

```{r include=TRUE}
knitr::kable(comp[order(nchar(` `)), .SD, Domain])
```

[**Table 1S**]{#tab1s}. Model performance for each random effect structure grouped for each developmental domain. The letters in the last columns represent: a, model with the lowest Akaike Information Criterion (AIC); b, model with the lowest Bayesian Information Criterion (BIC); c, model with the maximum R^2^; d, model with lowest esidual standard deviation (Sigma); e, model with the lowest root mean squared error (RMSE).

Considering the previous findings, we considered most appropriate to adjust for all variables in subsequent analyses in order maintain consistency between domains and the models fitted, given that the fully adjusted models tend to reduce the RMSE, and Sigma.

# References

<div id="refs"></div>

### 2022-11-06 R/data-prep.R

@@ -177,5 +177,5 @@
 #' @seealso
 #' - \link[=dataset]{dataset}
 #'
-#' @name data-prep
+#' @name data_prep
 NULL

### 2022-11-06 R/dataset.R

@@ -43,7 +43,7 @@
 #'   questionnaire [ASQ-3](https://agesandstages.com/products-pricing/asq3/).
 #'
 #' @seealso
-#' - \link[=data-prep]{Data preparation process}
+#' - \link[=data_prep]{Data preparation process}
 #'
 #' @source CADI-UMAG, Punta Arenas, Chile.
 "dataset"

### 2022-11-06 R/plot_gam.R

@@ -17,7 +17,7 @@ near <- function(i, j) {
   return(output)
 }
 
-#' Plot GAM model of bivariate model
+#' Plot GAM model
 #'
 #' Display graphical representation of a gam model for bivariate
 #' relationships. Model specification can be set with the ellipsis
@@ -59,3 +59,102 @@ plot_gam <- function(data, x, y, xlab, ylab, round_to = 1, ...) {
                                                        " SE in ", round_to, "-unit intervals")) +
     see::theme_lucid()
 }
+
+#' Plot varying random effects structure GAM model
+#'
+#' Display graphical representation different gam models on the same
+#' response using different random effects structure and comparing
+#' the estimates.
+#'
+#' @param data Dataset from where to extract response and independent variables.
+#' @param var Character. Response variable to test on gam models with varying random effects structures.
+#' @param seed Numeric. Seed for random processes within gams.
+#' @param legend Logical. Whether to include the figure legend.
+#' @param var_name Character. The label for the y-axis. Default is NULL, so "Predicted score" is used.
+#' @param plot Logical. Whether to return the plot. If FALSE, then the models with the varying random effects structures are returned.
+#' @param ... Currently not used.
+#'
+#' @export
+
+gam_confounders <- function(data, var, seed = 1234, legend = TRUE, var_name = NULL, plot = TRUE, ...) {
+  models = list(
+    # Simple model - only smooth term
+    `Simple` = "var ~ s(edad_corregida_meses)",
+    # Testing possible confounders one by one
+    `Only sex` = "var ~ s(sexo_paciente, bs = \"re\") + s(edad_corregida_meses)",
+    `Clinician` = "var ~ s(profesional_id, bs = \"re\") + s(edad_corregida_meses)",
+    `Respondent` = "var ~ s(respondedor_vinculo, bs = \"re\") + s(edad_corregida_meses)",
+    # Testing combination of pairs
+    `Clinician + Sex` = "var ~ s(profesional_id, bs = \"re\") + s(sexo_paciente, bs = \"re\") + s(edad_corregida_meses)",
+    `Clinician + Respondent` = "var ~ s(profesional_id, bs = \"re\") + s(respondedor_vinculo, bs = \"re\") + s(edad_corregida_meses)",
[…]

### 2022-11-06 manuscript/figures/fig-1.R

@@ -1,0 +1,53 @@
+
+# Model comparisons -------------------------------------------------------
+
+## Choosing between simple or adjusted models
+
+source(file = "manuscript/misc/setup.R")
+
+params <- grep("total$", names(dataset), value = TRUE)
+
+names(params) <- c("CM", "GM", "FM", "CG", "PS")
+
+plots <- mapply(
+  FUN = gam_confounders,
+  data = list(dataset),
+  var = params,
+  var_name = names(params),
+  legend = c(T,T,T,F,F),
+  SIMPLIFY = FALSE
+)
+
+f0 <- ggpubr::ggarrange(plotlist = plots[1:3], nrow = 1, common.legend = TRUE)
+f1 <- ggpubr::ggarrange(plotlist = plots[4:5], nrow = 1)
+figure1 <- ggpubr::ggarrange(f0, f1, ncol = 1)
+
+local({
+  pdf("manuscript/figures/fig-1.pdf", width = 15, height = 10);
+  print(figure1);
+  dev.off()
+})
+
+# Comparing models
+mod_comparison <- lapply(models, function(i) {
+  out <- as.data.table(
+    x = do.call(
+    what = AIC,
+    args = unname(i)
+    )
+  )
+  out[, model := names(i)][]
+})
+
+mod_comparison <- rbindlist(mod_comparison, idcol = "Response")
+best_worst_mods <- mod_comparison[, .SD[AIC %in% range(AIC), list(AIC, model)], list(Response)]
+
+best_worst_mods[, Classification := ifelse(max(AIC) == AIC, "Worst", "Best"), Response
+                ][, tapply(paste0(round(AIC), " (", model, ")"), Classification, c, simplify = FALSE), list(Response)]
+
+# Evaluation smooth significance
+lapply(models, function(i) {
+  lapply(i, report_smooth)
+})
+
+models$CM$`Full model` |> plot()

### 2022-11-06 manuscript/figures/fig-2.R

@@ -74,7 +74,7 @@ boots <- melt(boots, id.vars = 1:2, variable.name = "boot", value.name = "predic
 
 
 # Plot the data
-fig_1a <- ggplot(fig_data, aes(x = edad_corregida_meses, y = value)) +
+fig_2a <- ggplot(fig_data, aes(x = edad_corregida_meses, y = value)) +
   facet_grid(variable ~ ., scale = "free_y") +
   geom_count(alpha = 0.1, na.rm = TRUE) +
   geom_line(data = boots, aes(group = boot, y = predicted), alpha = 0.02, col = "red") +
@@ -147,12 +147,12 @@ slopes <- sapply(
   simplify = FALSE
 )
 
-fig_1b_data <- rbindlist(slopes, idcol = "response")
-fig_1b_data[, Confidence := fifelse(p < 0.05, "Significant", "Not significant")
+fig_2b_data <- rbindlist(slopes, idcol = "response")
+fig_2b_data[, Confidence := fifelse(p < 0.05, "Significant", "Not significant")
           ][, grp := rleid(Confidence)
           ][, response := factor(response, levels = names(models))]
 
-fig_1b <- ggplot(fig_1b_data, aes(edad_corregida_meses, Coefficient)) +
+fig_2b <- ggplot(fig_2b_data, aes(edad_corregida_meses, Coefficient)) +
   facet_grid(response ~ ., scales = "free_y") +
   labs(x = "Corrected age (months)",
        y = "Effect of Corrected age",) +
@@ -162,11 +162,11 @@ fig_1b <- ggplot(fig_1b_data, aes(edad_corregida_meses, Coefficient)) +
   ggpubr::theme_pubr() +
   scale_color_manual(values = ggsci::pal_lancet()(2), aesthetics = c("fill", "col"))
 
-fig_1 <- ggpubr::ggarrange(fig_1a, fig_1b, ncol = 2)
+fig_2 <- ggpubr::ggarrange(fig_2a, fig_2b, ncol = 2)
 
 if (isTRUE(x = .save_plot)) {
-  pdf("manuscript/fig-1.pdf", width = 12, height = 10);
-  print(fig_1);
+  pdf("manuscript/figures/fig-2.pdf", width = 12, height = 10);
+  print(fig_2);
   dev.off()
 }
 

### 2022-11-06 manuscript/figures/fig-3.R

@@ -1,0 +1,11 @@
+# Preparamos el ambiente de trabajo ---------------------------------------
+
+## Script de preparación
+source("manuscript/misc/setup.R")
+
+# Preparamos los datos ----------------------------------------------------
+
+ind <- grep("interpretacion", names(dataset), value = TRUE)
+
+
+

### 2022-11-06 manuscript/misc/setup.R

@@ -16,5 +16,6 @@ dataset <- dataset[i = n_evaluacion == 1 & diagnostico %like% "hipotonia" & edad
 
 # Drop levels of unused factors
 fv <- dataset[, names(.SD), .SDcols = is.factor]
-dataset[, (fv) := lapply(.SD, droplevels), .SDcols = fv]
+dataset[, (fv) := lapply(.SD, droplevels), .SDcols = fv][]
 rm(fv)
+

### 2022-11-06 manuscript/misc/table.R

@@ -50,3 +50,6 @@ tbl1 <- modify_header(
   stat_0 = "**Overall**<br>N = {N}"
 )
 
+as_gt(tbl1) |>
+  gt::gtsave(filename = "manuscript/table2.rtf")
+

### 2022-11-06 R/plot_gam.R
add figure 3 and its code, and new function
- New function to calculate logistic GAMs providing a numeric response and a threshold
- Figure 3 and its code

@@ -158,3 +158,101 @@ utils::globalVariables(
             "Adjusted for:"),
   add = TRUE
 )
+
+
+#' Plot ordinal GAM
+#'
+#' Display graphical representation of ordinal regression gam models
+#' on the response using the random effects structure and comparing
+#' the estimates.
+#'
+#' @param data Dataset from where to extract response and independent variables.
+#' @param var Character. Response variable to test on gam models with varying random effects structures.
+#' @param seed Numeric. Seed for random processes within gams.
+#' @param legend Logical. Whether to include the figure legend.
+#' @param var_name Character. The label for the y-axis. Default is NULL, so "Predicted score" is used.
+#' @param plot Logical. Whether to return the plot. If FALSE, then the models with the varying random effects structures are returned.
+#' @param threshold Numeric. Threshold to calculate probability.
+#' @param ... Currently not used.
+#'
+#' @export
+
+gam_ordinal <- function(data, var, seed = 1234, legend = TRUE, var_name = NULL, plot = TRUE, threshold = 35, ...) {
+  model <- "as.numeric(%var% < threshold) ~ s(profesional_id, bs = \"re\") + s(sexo_paciente, bs = \"re\") + s(respondedor_vinculo, bs = \"re\") + s(edad_corregida_meses)"
+  model <- gsub("%var%", replacement = var, model)
+  model <- stats::as.formula(model)
+
+  set.seed(seed)
+  output <- mgcv::gam(model, family = stats::binomial(link = "logit"), data = data)
+
+  if (isFALSE(plot)) {
+    return(output)
+  }
+
+  testdata <- expand.grid(
+    profesional_id = c(2),
+    sexo_paciente = c("F"),
+    respondedor_vinculo = c("Madre"),
+    edad_corregida_meses = seq(0, 48, 0.05)
+  )
+
+  predicted_response <- as.data.table(x = stats::predict(output, testdata, se.fit = TRUE, type = "response"))
+  predicted_response <- cbind(predicted_response, testdata)
+
+  names(predicted_response)[1:2] <- c("Fit", "SE")
+
+  ylab <- "P(Category | Corrected Age)"
[…]

### 2022-11-06 manuscript/figures/fig-3.R

@@ -3,9 +3,48 @@
 ## Script de preparación
 source("manuscript/misc/setup.R")
 
+## Cargamos paquetes
+library(ggplot2)
+
 # Preparamos los datos ----------------------------------------------------
 
-ind <- grep("interpretacion", names(dataset), value = TRUE)
+ind <- grep("total", names(dataset), value = TRUE)
+names(ind) <- c("CM", "GM", "FM", "CG", "PS")
+
+plots <- mapply(
+  FUN = gam_ordinal,
+  data = list(dataset),
+  var = ind,
+  var_name = names(ind),
+  legend = c(T,T,F,F,F),
+  threshold = 35,
+  SIMPLIFY = FALSE
+)
+
+f0 <- ggpubr::ggarrange(plotlist = plots[1:2], nrow = 1, common.legend = TRUE)
+f1 <- ggpubr::ggarrange(plotlist = plots[3], nrow = 1)
+f2 <- ggpubr::ggarrange(plotlist = plots[4:5], nrow = 1)
+figure3 <- ggpubr::ggarrange(f0, f1, f2, ncol = 1)
+
+local({
+  pdf("manuscript/figures/fig-3.pdf", width = 10, height = 12);
+  print(figure3);
+  dev.off()
+})
+
+
+# Evaluacion de los modelos -----------------------------------------------
+
+models <- mapply(
+  FUN = gam_ordinal,
+  var = ind,
+  data = list(dataset),
+  plot = FALSE,
+  threshold = 35,
+  SIMPLIFY = FALSE,
+  USE.NAMES = TRUE
+)
 
 
 
+lapply(models, report_overall)

### 2022-11-22 manuscript/manuscript.Rmd
add figure 3 and its code, and new function
- New function to calculate logistic GAMs providing a numeric response and a threshold
- Figure 3 and its code
several updates
- add cutoff scores for delay in developmental domains
- add report functionalities for binomial family of gam regressions
- update manuscript with logistic regressions sections

@@
 Neurodevelopmental status was assessed with the ASQ-3 at admission, as well as sociodemographic and obstetric data obtained through the initial clinical interview.
-**Results**: When modelling the neurodevelopmental status of each domain, an overall negative effect of corrected age on communication and problem-solving skills was observed, whereas the overall effect tends to be positive in gross motor function.
+**Results**: When modeling the neurodevelopmental status of each domain, an overall negative effect of corrected age on communication and problem-solving skills was observed, whereas the overall effect tends to be positive in gross motor function.
 Fine motor skills did not exhibit a linear relationship with corrected age but a non-linear effect, unlike the personal-social domain, which did not present significant variations across age corrected for prematurity.
@@
 
-   - Describes levels of variations observed in neurodevelopmental traits are present across age in hypotonic infants.
-   - Observed variations in developmental domains are not solely attributable to prematurity, where age corrected for prematurity best explains the observed variability in neurodevelopment.
+   - Describes the levels of variations observed in neurodevelopmental traits across age in hypotonic infants.
+   - Provides the estimated probability of developmental delay according to the ASQ-3 across age in hypotonic infants.
+   - The variations observed in developmental domains are not solely attributable to prematurity, where age corrected for prematurity best explains the variability observed in neurodevelopment.
 
@@
 In order to assess the influence of confounding variables, we fitted different models using the variables: i) sex of the infant, ii) clinician ID and iii) the relationship of the infant to the respondent (i.e., whether the respondent was the infant's mother, father, uncle or grandparent) as random effects in the form of penalized parametric terms [@wood2016smoothing].
-Varying random effects structures models can be seen in the *random effects*' section of the supplementary file.
+Varying random effects structures models can be seen in the section *Random effects structures* of the supplementary file.
+
+Based on the technical report of the ASQ-3, we've been able to estimate the threshold of possible delay based on the total score for each developmental domain.
+In this sense, a logistic regression, within GAM framework, was used to estimate the predicted probability of having a possible delay in developmental domains given the corrected age.
 
@@
 
-When modelling the effect of chronological age on developmental domains, different random effects structures were compared when assessing model estimates, these estimates can be seen in [Figure 1](#fig1).
+When modeling the effect of chronological age on developmental domains, different random effects structures were compared when assessing model estimates, these estimates can be seen in [Figure 1](#fig1).
 
@@
 
+```{r, eval=FALSE}
+ind <- grep("interpretacion", names(dataset), value = TRUE)
+names(ind) <- c("CM", "GM", "FM", "CG", "PS")
+
+models <- mapply(
+  FUN = gam_ordinal,
+  var = ind,
+  data = list(dataset),
+  plot = FALSE,
+  SIMPLIFY = FALSE,
+  USE.NAMES = TRUE
+)
+
+lapply(models, summary)
+lapply(models, report_smooth)
+lapply(models, report_slopes, term = "edad_corregida_meses")
+```
+
+When analyzing the probability of having a developmental delay based on the scores for each domain, we found a significant non-linear effect for CM ($\chi^2_{smooth}$ (7.73) = 32.11, *p* < 0.001), whereas in this domain we observed that for every one-month increase in the corrected age in the 13.6 to 18.9 age range, the odds of a possible developmental delay increases by 60% (OR = 1.6 CI~95%~[1.16, 2.23], $t_{student}$ (224.56) = 2.89, *p* = 0.01), as well with the 28.6 to 29.6 age range, whereas the odds of delay increases by 36% in this domain (OR = 1.36 CI~95%~[1.00, 1.86], $t_{student}$ (224.56) = 1.98, *p* = 0.049).
+Similar to CM, an overall linear effect of corrected age in the GM domain was observed ($\chi^2_{smooth}$ (1) = 6.86, *p* = 0.009), whereas for every one-month increase the odds of a possible developmental delay decrease by a 3.9% (OR = 0.96, CI~95%~[0.50, 0.99], $t_{student}$ (232) = -2.62, *p* = 0.009).
+The probability associated with a possible delay according to the total score for each developmental domain in response to corrected age can be observed in [Figure 3](#fig3).
+
 # Discussion
@@
 
-\{Add table\}
+Añadir tabla
 
@@
 Significant areas consider CI~95%~ that do not cross zero.
+
+[**Figure 3**]{#fig3}.
+Predicted probability of having a possible delay according to the total score for each developmental domain in response to corrected age.
+Fitted models were estimated using logistic regression within GAM framework and adjusted for possible confounders as already described.
+The p-values of smooth terms, associated with the main effect of corrected age on developmental delay, are shown within each panel.

### 2022-11-22 R/plot_gam.R

@@ -172,13 +172,12 @@ utils::globalVariables(
 #' @param legend Logical. Whether to include the figure legend.
 #' @param var_name Character. The label for the y-axis. Default is NULL, so "Predicted score" is used.
 #' @param plot Logical. Whether to return the plot. If FALSE, then the models with the varying random effects structures are returned.
-#' @param threshold Numeric. Threshold to calculate probability.
 #' @param ... Currently not used.
 #'
 #' @export
 
-gam_ordinal <- function(data, var, seed = 1234, legend = TRUE, var_name = NULL, plot = TRUE, threshold = 35, ...) {
-  model <- "as.numeric(%var% < threshold) ~ s(profesional_id, bs = \"re\") + s(sexo_paciente, bs = \"re\") + s(respondedor_vinculo, bs = \"re\") + s(edad_corregida_meses)"
+gam_ordinal <- function(data, var, seed = 1234, legend = TRUE, var_name = NULL, plot = TRUE, ...) {
+  model <- "%var% ~ s(profesional_id, bs = \"re\") + s(sexo_paciente, bs = \"re\") + s(respondedor_vinculo, bs = \"re\") + s(edad_corregida_meses)"
   model <- gsub("%var%", replacement = var, model)
   model <- stats::as.formula(model)
 
@@ -188,6 +187,8 @@ gam_ordinal <- function(data, var, seed = 1234, legend = TRUE, var_name = NULL, 
   if (isFALSE(plot)) {
     return(output)
   }
+
+  p_val <- paste("p", format_p(x = stats::anova(output)$s.table[4L, 4L]))
 
   testdata <- expand.grid(
     profesional_id = c(2),
@@ -203,7 +204,7 @@ gam_ordinal <- function(data, var, seed = 1234, legend = TRUE, var_name = NULL, 
 
   ylab <- "P(Category | Corrected Age)"
   if (!is.null(var_name) && is.character(var_name)) {
-    ylab <- paste0("P(", paste(var_name, "<", threshold), " | Corrected Age)")
+    ylab <- paste0("P(", paste("Delay in", var_name), " | Corrected Age)")
   }
 
   ci_prob <- function(p) stats::qnorm((p+1)/2, lower.tail = TRUE)
@@ -233,6 +234,7 @@ gam_ordinal <- function(data, var, seed = 1234, legend = TRUE, var_name = NULL, 
[…]

### 2022-11-22 R/report_stats.R

@@ -324,6 +324,8 @@ report_smooth <- function(model, ...) {
     "Is not a GAM model" = inherits(model, "gam")
   )
 
+  is_binomial <- stats::family(model)$family == "binomial"
+
   mod_summary <- summary(model)
 
   tbl <- data.table::as.data.table(
@@ -331,17 +333,31 @@ report_smooth <- function(model, ...) {
     keep.rownames = TRUE
   )
 
-  tbl$df_error  <- round(x = stats::df.residual(model), digits = 2)
   tbl$`p-value` <- format_p(tbl$`p-value`)
   tbl$edf       <- round(tbl$edf, 2)
-  tbl$F         <- round(tbl$F, 2)
+  if (is_binomial) {
+    tbl$Ref.df      <- round(x = tbl$Ref.df, digits = 2)
+    tbl$Chi.sq  <- round(tbl$Chi.sq, 2)
+  } else {
+    tbl$df      <- round(x = stats::df.residual(model), digits = 2)
+    tbl$F       <- round(tbl$F, 2)
+  }
 
-  expr <- paste(
-    paste0("$F_{smooth}$ (", tbl$edf, ", ", tbl$df_error, ") = ", tbl$F),
-    paste0("*p* ", tbl$`p-value`),
-    sep = ", ",
-    recycle0 = TRUE
-  )
+  if (is_binomial) {
+    expr <- paste(
+      paste0("$\\chi^2_{smooth}$ (", tbl$Ref.df, ") = ", tbl$Chi.sq),
+      paste0("*p* ", tbl$`p-value`),
+      sep = ", ",
+      recycle0 = TRUE
+    )
+  } else {
+    expr <- paste(
+      paste0("$F_{smooth}$ (", tbl$edf, ", ", tbl$df, ") = ", tbl$F),
+      paste0("*p* ", tbl$`p-value`),
+      sep = ", ",
+      recycle0 = TRUE
+    )
+  }
 
   if (length(tbl$rn) > 1) {
     expr <- data.table::data.table(
@@ -354,3 +370,4 @@ report_smooth <- function(model, ...) {
   return(expr)
 
 }
+

### 2022-11-22 data-raw/clean_dataset.R

@@ -172,6 +172,46 @@ library(data.table)
   setcolorder(dataset, neworder = readLines("data-raw/helpers/colorder"))
   setkey(dataset, paciente_id, fecha_evaluacion)
 
+
+# add cutoff values -------------------------------------------------------
+
+  # Use cutoffs based on technical report
+  cutoffs <- fread("data-raw/helpers/asq_cutoffs.csv")
+
+  # We find the rightmost age interval from the data
+  cutoffs <- cutoffs[, .(
+    edad_corregida = dataset$edad_corregida_meses,
+    edad = edad[findInterval(dataset$edad_corregida_meses, edad, left.open = TRUE) + 1],
+    cutoff = cutoff[findInterval(dataset$edad_corregida_meses, edad, left.open = TRUE) + 1]
+  ), domain]
+
+  # Create an id for reshaping the data to wide format
+  cutoffs[, id := seq_len(.N), domain][]
+
+  # Reshape the data to wide format
+  cutoffs <- dcast(cutoffs, id + edad_corregida ~ domain, value.var = "cutoff")[, -c(1L, 2L)]
+
+  # Bind colum-wise the cutoffs and the dataset
+  dataset <- cbind(dataset, cutoffs)
+
+  # Apply the cutoffs to their respective developmental domain
+  dataset[, `:=`(
+    comunicacion_interpretacion = ifelse(comunicacion_total < comunicacion_cutoff, "Possible Delay", "As expected"),
+    motora_gruesa_interpretacion = ifelse(motora_gruesa_total < motora_gruesa_cutoff, "Possible Delay", "As expected"),
+    motora_fina_interpretacion = ifelse(motora_fina_total < motora_fina_cutoff, "Possible Delay", "As expected"),
+    resolucion_problemas_interpretacion = ifelse(resolucion_problemas_total < resolucion_problemas_cutoff, "Possible Delay", "As expected"),
+    socio_individual_interpretacion = ifelse(socio_individual_total < socio_individual_cutoff, "Possible Delay", "As expected")
+  )]
+
+  # Find interpretation columns
+  ind <- grep("interpretacion", names(dataset), value = TRUE)
+
+  # And then transform them into factors
+  dataset[, (ind) := lapply(.SD, factor), .SDcols = ind]
+
+  # Remove temporal files
+  rm(cutoffs, ind)
+
 # Export dataset ------------------------------------------------------------------------------
 
   use_data(dataset, overwrite = TRUE)

### 2022-11-22 manuscript/figures/fig-3.R

@@ -8,21 +8,21 @@ library(ggplot2)
 
 # Preparamos los datos ----------------------------------------------------
 
-ind <- grep("total", names(dataset), value = TRUE)
+ind <- grep("interpretacion", names(dataset), value = TRUE)
 names(ind) <- c("CM", "GM", "FM", "CG", "PS")
+
 
 plots <- mapply(
   FUN = gam_ordinal,
   data = list(dataset),
   var = ind,
   var_name = names(ind),
-  legend = c(T,T,F,F,F),
-  threshold = 35,
+  legend = c(F,T,T,F,F),
   SIMPLIFY = FALSE
 )
 
-f0 <- ggpubr::ggarrange(plotlist = plots[1:2], nrow = 1, common.legend = TRUE)
-f1 <- ggpubr::ggarrange(plotlist = plots[3], nrow = 1)
+f0 <- ggpubr::ggarrange(plotlist = plots[2:3], nrow = 1, common.legend = TRUE)
+f1 <- ggpubr::ggarrange(plotlist = plots[1], nrow = 1)
 f2 <- ggpubr::ggarrange(plotlist = plots[4:5], nrow = 1)
 figure3 <- ggpubr::ggarrange(f0, f1, f2, ncol = 1)
 
@@ -40,11 +40,8 @@ models <- mapply(
   var = ind,
   data = list(dataset),
   plot = FALSE,
-  threshold = 35,
   SIMPLIFY = FALSE,
   USE.NAMES = TRUE
 )
 
-
-
-lapply(models, report_overall)
+lapply(models, summary)

### 2022-11-22 manuscript/misc/setup.R

@@ -5,6 +5,10 @@ library(data.table)
 # For GAMS
 library(mgcv)
 
+# For tables
+library(gt)
+library(gtsummary)
+
 # Load the data
 data("dataset", package = "ASQ3")
 

### 2022-11-22 manuscript/tables/table.R

@@ -41,15 +41,18 @@ tbl1 <- add_overall(x = tbl1)
 tbl1 <- add_p(
   x = tbl1,
   test = everything() ~ "wilcox.test",
-  pvalue_fun = function(x) style_pvalue(x, digits = 3)
+  pvalue_fun = function(x) round(x, digits = 3)
 )
 
-tbl1 <- modify_header(
-  x = tbl1,
-  all_stat_cols(stat_0 = FALSE) ~ "**{level}**<br>N = {n}",
-  stat_0 = "**Overall**<br>N = {N}"
-)
+# tbl1 <- modify_header(
+#   x = tbl1,
+#   all_stat_cols(stat_0 = FALSE) ~ "**{level}**<br>N = {n}",
+#   stat_0 = "**Overall**<br>N = {N}"
+# )
 
-as_gt(tbl1) |>
-  gt::gtsave(filename = "manuscript/table2.rtf")
+tbl1 <- as_gt(tbl1)
 
+
+
+saveRDS(tbl1, file = "manuscript/tables/table-1.RDS")
+gt::gtsave(tbl1, filename = "manuscript/tables/table-1.rtf")

### 2022-12-14 manuscript/CICYPE/presentation.qmd
presentacion para el CICYPE

---
title: "Descripción y caracterización del neurodesarrollo de niños con hipotonía congénita"
title-slide-attributes:
  data-background-image: resources/title-background.jpeg
  data-background-opacity: "0.4"
  data-background-color: "black"
subtitle: "de la región de Magallanes y la Antártica Chilena"
author:
  - Matías Castillo-Aguilar ^1,2^
institute: 
  - ^1^ Departamento de Kinesiología, Universidad de Magallanes, Punta Arenas, Chile.
  - ^2^ Neurofisiología Integrativa Molecular Austral Chilena (NIM-ACh), Punta Arenas, Chile.
format: 
  revealjs:
    auto-play-media: true
    theme: simple
    toc: true
    toc-depth: 1
    toc-title: Contenidos
    transition: slide
    background-transition: slide
    controls-layout: bottom-right
    default-image-extension: svg
    view-distance: 3
editor: visual
footer: "Matías Castillo-Aguilar"
css:
  - css/title-slide.css
  - css/logo.css 
  - css/media.css
  - css/styles.css
logo: logo.png
bibliography: references.bib
editor_options: 
  chunk_output_type: console
---

# Conceptualización

Entendiendo la problemática

## {background-color="black" background-video="resources/boy_vid.mp4" background-video-loop="true" background-video-muted="true" background-opacity=0.7 .smaller}

. . .

:::{.absolute top=25% width=35%}
Los signos físicos y psicológicos del desarrollo en la primera infancia son marcadores representativos y pertinentes para identificar y controlar el crecimiento general en los primeros años de vida [@di2016predictive].
:::

## {background-color="black" background-video="resources/boy_vid_2.mp4" background-video-loop="true" background-video-muted="true" background-opacity=0.5 .smaller}

. . .

:::{.absolute top=12% left=33% width=35%}
Múltiples afecciones neuromusculares, metabólicas y genéticas se han asociado a un neurodesarrollo variable en infantes.

En donde la disminución del tono muscular puede representar un signo de retraso en el desarrollo neurológico del niño, que puede predisponer a una peor función cognitiva en algunos casos [@riou2009global].
:::

## Lo motor y lo cognitivo van de la mano {.smaller}

![](resources/writing_kid.mp4){#writing_kid .absolute right=6% top=20% width=40% loop=true}
![](resources/girl_playing.mp4){#girl_playing .absolute left=0 top=60% width=40% loop=true}

<script>
var video = document.getElementById("writing_kid");
var video2 = document.getElementById("girl_playing");
video.controls = false;
video2.controls = false;
</script>

:::{.absolute .fragment .fade-in top=20% left=0 width=45%}
Existe evidencia que sugiere que las habilidades motoras gruesas y finas podrían tener un papel para predecir el normal desarrollo de habilidades comunicativas [@gonzalez2019gross]. 
:::

:::{.absolute .fragment .fade-in top=65% right=6% width=45%}
Esto pone sobre la mesa, la necesidad de tener estrategias de monitorización para favorecer la [derivación y atención temprana]{.fragment .highlight-red}.
:::

## Hipotonía, un signo pasado por alto {.smaller}

![](resources/baby.mp4){#baby .absolute right=6% top=27% width=50% loop=true}

:::: {.fragment .fade-in-then-out .absolute top=30% width=40%}
La hipotonía tiende a generar alteraciones que dificultan el desarrollo psicomotor adecuado, pudiendo generar retrasos en futuras capacidades neuromotrices [@gabis2021weak].
::::

:::: {.fragment .fade-in .absolute top=28% width=40%}
Existen múltiples causas de hipotonía, pero cuando esta se encuentra presente [en ausencia de otros signos o síntomas]{.fragment .highlight-red} entonces hablamos hipotonía congénita (HC) [@leyenaar2005schematic].
::::

<script>
var video = document.getElementById("baby");
video.controls = false;
</script>

## {background-color="black" background-video="resources/group_discussion.mp4" background-video-loop="true" background-video-muted="true" background-opacity=0.5 background-transition="fade"}

::: {.fragment .fade-in-then-out .absolute left=25% top=20% width=50%}
Investigar el desarrollo de la HC es fundamental, dado que es un desorden neuromuscular no progresivo que tiende a mejorar [con la intervención temprana]{style="font-weight: 1000; color: red;"}.
:::

. . .

:::{.r-fit-text .absolute top=25% style="font-weight: 1000; text-align: center;"}
[NO HAY]{.fragment} [SUFICIENTE]{.fragment}

[EVIDENCIA]{.fragment}
:::

# Hipótesis y Objetivos

## Hipótesis

:::{.absolute left=20% width=60% top=30%}
Los niños con hipotonía congénita presentan
características diferentes en su neurodesarrollo de acuerdo a su edad.
:::

## Objetivos

:::{.fragment .fade-out .absolute left=10% top=35% width=80%}
### General
Describir y caracterizar el neurodesarrollo de niños con HC de la
región de Magallanes y la Antártica Chilena.
:::

::::{.fragment .fade-in .absolute left=10% top=25% width=80%}
### Específicos
:::{style="font-size: 28px;"}
- Describir y caracterizar la relación entre el [neurodesarrollo y la prematuridad]{.fragment .highlight-red} en
niños diagnosticados de HC.
- Describir y caracterizar la relación entre el [neurodesarrollo y la edad]{.fragment .highlight-red} en niños
diagnosticados de HC.
- Describir y caracterizar [la influencia de la edad sobre el riesgo de presentar un
retardo en el neurodesarrollo]{.fragment .highlight-red} en niños diagnosticados de HC.
:::
::::

# Métodos

¿Cómo abordamos este problema?

## Diseño del estudio

![](resources/methods_design.jpeg){.absolute left=8% width=35% top=20%}

::: {.absolute width=40% top=25% right=10%}
- Realizamos un estudio observacional.

- Corte transversal.

- Paradigma cuantitativo.
:::

## Participantes

![](resources/participants.mp4){.absolute #participants left=0 width=50% top=30% loop=true}

::: {.fragment .fade-out .absolute width=45% right=0 top=29% style="font-size: 32px;"}
Un total de 234 usuarios con diagnostico de HC fueron incluídos como parte de un programa de intervención llevado a cabo por el Centro de Rehabilitación Cruz del Sur (CRCS) de Punta Arenas, Chile.
:::

::: {.fragment .fade-in-then-out .absolute width=45% right=0 top=25% style="font-size: 30px;"}
El diagnóstico de HC se basó en 4 fuentes:

1. Pediatra de atención primaria.
2. Pediatras de clínicas privadas.
3. Neuropediatra del Hospital Clínico.
4. Fisiatra de la institución.
:::

::: {.fragment .fade-in .absolute width=45% right=0 top=36% style="font-size: 34px;"}
Una vez hecho el diagnóstico, este se refirió al programa de la institución para su posterior ingreso.
:::

<script>
var video = document.getElementById("participants");
video.controls = false;
</script>

## Resguardos éticos {.smaller}

![](resources/ethics.mp4){.absolute #ethics right=0 width=50% top=30% loop=true}

::: {.absolute width=45% top=30%}
- Los cuidadores legales otorgaron su consentimiento informado antes de la participación a este estudio.

- El Comité de Ética del CRCS aprobó este estudio siguiendo las regulaciones establecidas por la Declaración de Helsinki.
:::

<script>
var video = document.getElementById("ethics");
video.controls = false;
</script>

## Mediciones

![](resources/measures.mp4){.absolute #measures width=50% top=30% loop=true}

::: {.fragment .fade-in-then-out .absolute width=45% right=0 top=22% style="font-size: 32px;"}
### The Ages and Stages Questionnaire (ASQ-3)

- Instrumento de cribado del desarrollo de nivel inicial informado por los padres.
- Consta de 21 intervalos, cada uno con 30 ítems en cinco áreas.
:::

:::: {.fragment .fade-in-then-out .absolute width=45% right=0 top=23%}
#### Dominios del desarrollo:
:::{style="font-size: 32px;"}
- Comunicación (CM)
- Habilidades motoras gruesas (GM)
- Habilidades motoras finas (FM)
- Resolución de problemas (CG)
- Personal-social (PS)
:::
::::

::: {.fragment .fade-in-then-out .absolute width=45% right=0 top=28% style="font-size: 28px;"}
Es costo efectivo y ampliamente usado internacionalmente [@heo2008cross; @sarmiento2011universal].

Validado en infantes de término y pretérmino chilenos (75% sensibilidad and 81% especificidad) [@schonhaut2013validity].
:::

<script>
var video = document.getElementById("measures");
video.controls = false;
</script>

## Procedimientos {background-color="black" background-video="resources/procedures.mp4" background-video-loop="true" background-video-muted="true" background-opacity=0.5 background-transition="fade"}

:::: {.fragment .fade-out fragment-index=3}
::: {.fragment .fade-in-then-semi-out .absolute width=45% left=0 top=25% style="font-size: 32px;" fragment-index=1}
- Los datos sociodemográficos se obtuvieron de la base de datos electrónica de la institución.
:::
::::

::: {.fragment .fade-in-then-out .absolute width=45% left=0 top=45% style="font-size: 32px;" fragment-index=2}
- El ASQ-3 lo aplicó la enfermera del programa de acuerdo al protocolo establecido por el propio instrumento.
:::

::: {.fragment .fade-in .absolute width=45% left=0 top=22% style="font-size: 32px;" fragment-index=4}
Algunas consideraciones importantes de la administración son que:

1. Debe responder el [cuidador principal]{.fragment .highlight-red}.
2. En caso de dudas del cuidador sobre el comportamiento evaluado, la información se corrobora mediante pruebas *in-situ* con el usuario.
:::

# Análisis estadístico

```{r load-packages}
#| include: false
#| message: false

source("prep-workspace.R")
```

## Software estadístico

![](resources/rstudio.png){.absolute width=40% left=10% top=30%}

![](resources/rstudio-desktop-screen.png){.absolute width=40% right=10% top=12%}

## Análisis exploratorio

::::{.fragment .fade-in-then-out .absolute width=80%% left=10% right=10%}
![](resources/density.svg)

::: {.absolute bottom=0% style="font-size: 20px; color: gray;"}
*Gráficos de densidad univariada. Las líneas rojas representan la media, y las azules la mediana.*
:::
::::

```{r}
#| echo: true
#| classes: ".fragment .absolute top=20% left=15%"
# En R, podemos evaluarlo así:
dataset[, list(
  CM = shapiro_test(comunicacion_total),
  GM = shapiro_test(motora_gruesa_total),
  FM = shapiro_test(motora_fina_total),
  CG = shapiro_test(resolucion_problemas_total),
  PS = shapiro_test(socio_individual_total)
)]
```

::: {.fragment .absolute top=65% width=60% left=15% style="font-size:30px;"}
De acuerdo a lo anterior, no podemos asumir la normalidad de las variables.
:::

## Estadísticos {.smaller}

::: {.fragment .fade-in-then-semi-out .absolute top=20% width=45%}
### Descriptivos

- Variables numéricas: mediana y rango intercuartílico (estimadores robustos de centralidad y dispersión).
- Variables discretas: frecuencia absoluta y relativa.
:::

::: {.fragment .fade-in-then-out .absolute top=20% width=45% right=7%}
### Comparaciones

- La prueba de suma de rangos de Wilcoxon para comparar puntajes del ASQ-3 entre hombres y mujeres.
- La prueba chi-cuadrado ( $\chi^2$) se usó para comparar la bondad de ajuste.
:::

## Modelado con GAMs {.smaller}

::: {.fragment .fade-out fragment-index=5}
![](resources/gams_vs_lm.svg){.fragment .fade-in .absolute right=0 top=10% fragment-index=1}
:::

:::: {.fragment .fade-out fragment-index=4}
::: {.fragment .fade-in-then-semi-out .absolute top=20% width=40% fragment-index=2}
- Empleamos modelos aditivos generalizados (GAMs) para describir el comportamiento del neurodesarrollo en respuesta de nuestros parámetros de interés.
:::
::::

::: {.fragment .fade-in-then-out .absolute top=55% width=40% fragment-index=3}
- Permite modelar relaciones no lineales de alta dimensionalidad, pudiendo describir efectos complejos.
:::

:::: {.fragment .fade-out fragment-index=9}

![](resources/simpson_paradox.png){.fragment .fade-in .absolute left=20% top=20% width=60% fragment-index=6}

::: {.fragment .fade-in .absolute top=25% right=0 width=20% fragment-index=7 style="font-size: 20px; color: gray;"}
*Paradoja de simpson*
:::

::: {.fragment .fade-in .absolute top=55% left=27% width=50% fragment-index=8}
Sumado a lo anterior, también podemos controlar por la influencia de otras variables para obtener estimaciones más robustas a factores de confusión.
:::
::::

::: {.fragment .fade-in .absolute top=30% left=25% width=50% fragment-index=10, style="font-size:30px;"}
Para describir las estimaciones de los modelos usaremos derivadas aproximadas de primer orden con intervalos de confianza al 95%, con el fin de poder describir los efectos observados en términos de segmentos quasi-lineales [@dominique2020estimation].
:::

## Efectos aleatorios

:::: {.fragment .fade-out fragment-index=3}
::: {.fragment .fade-in .absolute left=0 width=45% top=30% style="font-size:30px;" fragment-index=1}
Para evaluar la influencia de las variables de confusión, ajustamos diferentes modelos con efectos aleatorios en forma de términos paramétricos penalizados. [@wood2016smoothing].
:::

::: {.fragment .fade-in .absolute right=0 width=45% top=30% style="font-size:30px;" fragment-index=2}
Se usaron las siguientes variables:

1. Sexo del lactante.
2. ID del clínico. 
3. Relación del lactante con el encuestado.
:::
::::

![](resources/confounders.svg){.fragment .fade-in .absolute width=100% left=5% fragment-index=4}

# Resultados

¿Qué fué lo que se encontró?

## Características de la muestra

:::{.absolute top=12% left=10% width=80% style="font-size:16px;"}
```{r}
readRDS("resources/table-1.RDS")
```
:::

## {background-color="white" background-video="resources/in_results.mp4" background-video-loop="true" background-video-muted="true" background-opacity=0.5 background-transition="fade"}

:::{style="font-weight: 1000; text-align: center; font-size: 80px;"}
[NEURODESARROLLO]{.fragment top=35% left=0 .absolute}

[Y PREMATURIDAD]{.fragment top=50% left=0 .absolute}
:::

---

:::{.absolute top=20% left=0 width=65% style="font-size: 30px;"}
```{r}
readRDS("resources/table-2.RDS")
```
:::

::: {.absolute top=25% right=0% width=30% style="font-size: 20px; color: gray;"}
*Tabla 1. Significancia de los términos no-lineales para la prematurez sobre las variables del desarrollo. Se presentan los modelos ajustados por las variables de confusión previamente descritas.*
:::

## {background-color="white" background-video="resources/in_results_2.mp4" background-video-loop="true" background-video-muted="true" background-opacity=0.5 background-transition="fade"}

:::{style="font-weight: 1000; text-align: center; font-size: 80px;"}
[NEURODESARROLLO]{.fragment top=35% left=0 .absolute}

[Y LA EDAD]{.fragment top=50% left=0 .absolute}
:::

---

::::{.fragment .fade-in-then-out}
:::{.absolute top=20% left=0 width=75% style="font-size: 30px;"}
```{r}
readRDS("resources/table-3.RDS")
```
:::

::: {.absolute top=25% right=5% width=20% style="font-size: 20px; color: gray;"}
*Tabla 2. Significancia de los términos no-lineales para la edad corregida (meses) sobre las variables del desarrollo. Se presentan los modelos ajustados por las variables de confusión previamente descritas.*
:::
::::

::::{.fragment .fade-in-then-out}
![](resources/results_main_CM.svg){.absolute left=5% width=90% top=0}

:::{.absolute top=70% left=10% width=80% style="font-size: 22px;"}
```{r}
readRDS("resources/table-main-CM.RDS")
```
:::
::::

::::{.fragment .fade-in-then-out}
![](resources/results_main_GM.svg){.absolute left=5% width=90% top=0}

:::{.absolute top=70% left=10% width=80% style="font-size: 22px;"}
```{r}
readRDS("resources/table-main-GM.RDS")
```
:::
::::

::::{.fragment .fade-in-then-out}
![](resources/results_main_FM.svg){.absolute left=5% width=90% top=0}

:::{.absolute top=70% left=10% width=80% style="font-size: 22px;"}
```{r}
readRDS("resources/table-main-FM.RDS")
```
:::
::::

::::{.fragment .fade-in-then-out}
![](resources/results_main_CG.svg){.absolute left=5% width=90% top=0}

:::{.absolute top=70% left=10% width=80% style="font-size: 22px;"}
```{r}
readRDS("resources/table-main-CG.RDS")
```
:::
::::

::::{.fragment .fade-in-then-out}
![](resources/results_main_PS.svg){.absolute left=5% width=90% top=10%}
::::

## {background-color="white" background-video="resources/in_results_2.mp4" background-video-loop="true" background-video-muted="true" background-opacity=0.5 background-transition="fade"}

:::{style="font-weight: 1000; font-size: 80px;"}
[EDAD]{.fragment top=20% left=0 .absolute}

[Y RETRASO DEL DESARROLLO]{.fragment top=35% left=0 .absolute}
:::

---

::::{.fragment .fade-in-then-out}
:::{.absolute top=20% left=0 width=80% style="font-size: 30px;"}
```{r}
readRDS("resources/table-risk-smooth.RDS")
```
:::

::: {.absolute top=25% right=5% width=20% style="font-size: 20px; color: gray;"}
*Tabla 7. Significancia de los términos no-lineales logísticos para la edad corregida (meses) sobre las variables del desarrollo. Se presentan los modelos ajustados por las variables de confusión previamente descritas.*
:::
::::

---

::::{.fragment .fade-in-then-out}
![](resources/results_risk_CM.svg){.absolute left=5% width=90% top=0}

:::{.absolute top=70% left=10% width=80% style="font-size: 22px;"}
```{r}
readRDS("resources/table-risk-CM.RDS")
```
:::
::::

::::{.fragment .fade-in-then-out}
![](resources/results_risk_GM.svg){.absolute left=5% width=90% top=0}

:::{.absolute top=70% left=10% width=80% style="font-size: 22px;"}
```{r}
readRDS("resources/table-risk-GM.RDS")
```
:::
::::

::::{.fragment .fade-in-then-out}
![](resources/results_risk_FM.svg){.absolute left=5% width=90% top=0}

:::{.absolute top=70% left=10% width=80% style="font-size: 22px;"}
```{r}
readRDS("resources/table-risk-FM.RDS")
```
:::
::::

::::{.fragment .fade-in-then-out}
![](resources/results_risk_CG.svg){.absolute left=5% width=90% top=0}

:::{.absolute top=70% left=10% width=80% style="font-size: 22px;"}
```{r}
readRDS("resources/table-risk-CG.RDS")
```
:::
::::

::::{.fragment .fade-in-then-out}
![](resources/results_risk_PS.svg){.absolute left=5% width=90% top=0}

:::{.absolute top=70% left=10% width=80% style="font-size: 22px;"}
```{r}
readRDS("resources/table-risk-PS.RDS")
```
:::
::::

# Conclusión

## {background-color="black" background-video="resources/conclusion.mp4" background-video-loop="true" background-video-muted="true" background-opacity=0.5 background-transition="fade"}

:::: {.fragment .fade-out fragment-index=3}
::: {.fragment .fade-in-then-semi-out .absolute width=50% left=0 top=5% style="font-size: 32px;" fragment-index=1}
- Se observan marcadas variaciones en los rasgos del neurodesarrollo a lo largo de la edad en los infantes hipotónicos, principalmente en forma de variaciones no lineales y de dominio específico.
:::
::::

::: {.fragment .fade-in-then-out .absolute width=50% left=0 top=45% style="font-size: 30px;" fragment-index=2}
- Mostramos que las variaciones observadas en los dominios del desarrollo no son atribuibles únicamente a la prematuridad, donde la edad corregida explica mejor la variabilidad observada en el neurodesarrollo.
:::

::: {.fragment .fade-in .absolute width=45% left=0 top=22% style="font-size: 32px;" fragment-index=4}
Futuras investigaciones podrían determinar cómo se aplican estos resultados cuando se controlan los factores sociales mediados por el contexto, así como otras poblaciones clínicas.
:::

# Referencias

## {.smaller}

### 2022-12-14 R/plot_gam.R

@@ -160,9 +160,9 @@ utils::globalVariables(
 )
 
 
-#' Plot ordinal GAM
+#' Plot logistic GAM
 #'
-#' Display graphical representation of ordinal regression gam models
+#' Display graphical representation of logistic regression gam models
 #' on the response using the random effects structure and comparing
 #' the estimates.
 #'
@@ -176,19 +176,20 @@ utils::globalVariables(
 #'
 #' @export
 
-gam_ordinal <- function(data, var, seed = 1234, legend = TRUE, var_name = NULL, plot = TRUE, ...) {
+gam_binomial <- function(data, var, seed = 1234, legend = TRUE, var_name = NULL, plot = TRUE, ...) {
   model <- "%var% ~ s(profesional_id, bs = \"re\") + s(sexo_paciente, bs = \"re\") + s(respondedor_vinculo, bs = \"re\") + s(edad_corregida_meses)"
   model <- gsub("%var%", replacement = var, model)
   model <- stats::as.formula(model)
 
   set.seed(seed)
-  output <- mgcv::gam(model, family = stats::binomial(link = "logit"), data = data)
+  output <- mgcv::gam(model, family = stats::binomial(link = "logit"), data = data,
+                      method = "REML")
 
   if (isFALSE(plot)) {
     return(output)
   }
 
-  p_val <- paste("p", format_p(x = stats::anova(output)$s.table[4L, 4L]))
+  p_val <- paste("p", format_p(x = summary(output)$s.table[4L, 4L]))
 
   testdata <- expand.grid(
     profesional_id = c(2),

### 2022-12-14 R/report_stats.R

@@ -285,9 +285,19 @@ report_slopes <- function(model, term, ci = 0.95, length = 100, k = 1, ...) {
 
   slopes$df_error <- round(slopes$df_error, 2)
   slopes$t <- round(slopes$t, 2)
+  b <- "$\\beta$ = "
+
+  is_binomial <- stats::family(model)$family == "binomial"
+  if(is_binomial) {
+    message("transforming coefficients from logodds to OR")
+    slopes$Coefficient <- exp(slopes$Coefficient)
+    slopes$CI_low      <- exp(slopes$CI_low)
+    slopes$CI_high     <- exp(slopes$CI_high)
+    b <- "OR = "
+  }
 
   expr <- paste(
-    paste0("$\\beta$ = ", round(slopes$Coefficient, 2)),
+    paste0(b, round(slopes$Coefficient, 2)),
     paste0("CI~",slopes$CI*100,"%~[", round(slopes$CI_low, 2), ", ", round(slopes$CI_high, 2), "]"),
     paste0("$t_{student}$ (", slopes$df_error, ") = ", slopes$t),
     paste0("*p* ", format_p(slopes$p)),

### 2022-12-14 manuscript/CICYPE/functions.R

@@ -1,0 +1,100 @@
+
+# Versión en español de las funciones en el paquete ASQ3
+
+gam_confounders_es <- function(data, var, seed = 1234, legend = TRUE, var_name = NULL, plot = TRUE, ...) {
+  models = list(
+    # Simple model - only smooth term
+    `Simple` = "var ~ s(edad_corregida_meses)",
+    # Testing possible confounders one by one
+    `Solo sexo` = "var ~ s(sexo_paciente, bs = \"re\") + s(edad_corregida_meses)",
+    `Clínico` = "var ~ s(profesional_id, bs = \"re\") + s(edad_corregida_meses)",
+    `Vínculo` = "var ~ s(respondedor_vinculo, bs = \"re\") + s(edad_corregida_meses)",
+    # Testing combination of pairs
+    `Clínico + Sexo` = "var ~ s(profesional_id, bs = \"re\") + s(sexo_paciente, bs = \"re\") + s(edad_corregida_meses)",
+    `Clínico + Vínculo` = "var ~ s(profesional_id, bs = \"re\") + s(respondedor_vinculo, bs = \"re\") + s(edad_corregida_meses)",
+    `Sexo + Vínculo` = "var ~ s(sexo_paciente, bs = \"re\") + s(respondedor_vinculo, bs = \"re\") + s(edad_corregida_meses)",
+    # And full model
+    `Ajuste completo` = "var ~ s(profesional_id, bs = \"re\") + s(sexo_paciente, bs = \"re\") + s(respondedor_vinculo, bs = \"re\") + s(edad_corregida_meses)"
+  )
+
+  models <- lapply(models, stats::as.formula)
+  models <- lapply(models, `[[<-`, 2, as.name(var))
+
+  set.seed(seed)
+  output <- lapply(models, mgcv::gam, data = data, method = "REML")
+
+  if (isFALSE(plot)) {
+    return(output)
+  }
+
+  testdata <- expand.grid(
+    profesional_id = c(2),
+    sexo_paciente = c("F", "M"),
+    respondedor_vinculo = c("Madre","Padre","Abuelo/a","Tio/a"),
+    edad_corregida_meses = seq(0, 48, 0.05)
+  )
+
+  testdata <- data.table::as.data.table(testdata)
+
+  predicted_response <- lapply(
+    output,
+    stats::predict,
+    type = "response",
+    newdata = testdata
+  )
+
+  predicted_response <- data.table::as.data.table(predicted_response)
[…]

### 2022-12-14 manuscript/CICYPE/prep-workspace.R

@@ -1,0 +1,32 @@
+# Cargamos paquetes
+
+## Paquete de datos
+library(ASQ3)
+
+## Modelado con GAMs
+library(mgcv)
+
+## Manipulación general
+library(data.table)
+
+## Tablas
+library(kableExtra)
+library(gt)
+library(gtsummary)
+
+## Graficos
+library(ggplot2)
+library(ggpubr)
+
+# Estilo por defecto para gráficos
+theme_set(theme_classic())
+
+source(here::here("manuscript/CICYPE/functions.R"))
+
+# Preparamos los datos
+dataset <- dataset[i = n_evaluacion == 1 & diagnostico %like% "hipotonia" & edad_cronologica_meses <= 60]
+
+# Drop levels of unused factors
+fv <- dataset[, names(.SD), .SDcols = is.factor]
+dataset[, (fv) := lapply(.SD, droplevels), .SDcols = fv]
+rm(fv)

### 2022-12-14 manuscript/CICYPE/resources/code/confounders.R

@@ -1,0 +1,22 @@
+source("manuscript/CICYPE/prep-workspace.R")
+
+params <- grep("total$", names(dataset), value = TRUE)
+
+names(params) <- c("CM", "GM", "FM", "CG", "PS")
+
+plots <- mapply(
+  FUN = gam_confounders_es,
+  data = list(dataset),
+  var = params,
+  var_name = names(params),
+  legend = c(T,T,T,F,F),
+  SIMPLIFY = FALSE
+)
+
+f0 <- ggpubr::ggarrange(plotlist = plots[1:3], nrow = 1, common.legend = TRUE)
+f1 <- ggpubr::ggarrange(plotlist = plots[4:5], nrow = 1)
+fig <- ggpubr::ggarrange(f0, f1, ncol = 1)
+
+svglite::svglite(filename = "manuscript/CICYPE/resources/confounders.svg", width = 11.25, height = 6)
+print(fig)
+dev.off()

### 2022-12-14 manuscript/CICYPE/resources/code/density.R

@@ -1,0 +1,13 @@
+source("manuscript/CICYPE/prep-workspace.R")
+
+fig <- ggarrange(
+  density_plot(comunicacion_total, "CM"),
+  density_plot(motora_gruesa_total, "GM"),
+  density_plot(motora_fina_total, "FM"),
+  density_plot(resolucion_problemas_total, "CG"),
+  density_plot(socio_individual_total, "PS")
+)
+
+svglite::svglite(filename = "manuscript/CICYPE/resources/density.svg", width = 8, height = 5)
+print(fig)
+dev.off()

### 2022-12-14 manuscript/CICYPE/resources/code/gams-vs-lm.R

@@ -1,0 +1,18 @@
+library(ggplot2)
+
+fig <- ggplot(MASS::mcycle, aes(times, accel)) +
+  geom_point() +
+  geom_smooth(method = "gam", se = TRUE, aes(col = "GAM")) +
+  geom_smooth(method = "lm", se = F, aes(col = "LM")) +
+  labs(col = "Modelos",
+       title = "Datos simulados de un accidente en motocicleta",
+       subtitle = "Medidas de aceleración de la cabeza",
+       caption = "Fuente: Silverman, B. W. (1985).",
+       x = "Tiempo", y = "Aceleración") +
+  scale_color_manual(values = c(GAM = "blue2", LM = "red3")) +
+  theme_classic()
+
+svglite::svglite(filename = "manuscript/CICYPE/resources/gams_vs_lm.svg", width = 6, height = 6)
+print(fig)
+dev.off()
+

### 2022-12-14 manuscript/CICYPE/resources/code/main-effect.R

@@ -1,0 +1,178 @@
+
+# Load packages and scripts -------------------------------------------------------------------
+
+# Setup script
+source("manuscript/CICYPE/prep-workspace.R")
+
+# R Packages
+library(ggplot2)
+library(see)
+
+# Figure 1a ------------------------------------------------------------------------------------
+
+# Effect of corrected age on developmental skills (many smooths)
+# Communication score ~ corrected age
+# Gross motor score ~ corrected age
+# Fine motor score ~ corrected age
+# Problem solving score ~ corrected age
+# Socio-individual score ~ corrected age
+
+## communication (CM), gross motor (GM), fine motor (FM), problem-solving (CG), and personal-social (PS)
+
+y_vars <- c("CM" = "comunicacion_total"
+            , "GM" = "motora_gruesa_total"
+            , "FM" = "motora_fina_total"
+            , "CG" = "resolucion_problemas_total"
+            , "PS" = "socio_individual_total")
+
+fig_a_data <- melt(
+  data = dataset,
+  measure.vars = y_vars
+)
+
+fig_a_data[, variable := `levels<-`(variable, names(y_vars))][]
+
+# Scaled data for mean ± SE plotting
+scaled <- copy(fig_a_data)
+scaled[, edad_corregida_meses := ASQ3:::near(edad_corregida_meses, 5)][]
+
+# 300 evenly-spaced data points of `edad_corregida_meses`
+pred_data <- fig_a_data[, data.frame(
+  edad_corregida_meses = seq.default(
+    from = min(edad_corregida_meses, na.rm = TRUE),
+    to = max(edad_corregida_meses, na.rm = TRUE),
+    length.out = 300
+  )
+)]
+
+# Bootstrapping with 100 replicates
+set.seed(123)
+boots <- fig_a_data[, {
+  boot <- vapply(1:200, function(i) { # Vectorized "looping"
+
+    # ith Model fitting
+    mod <- gam(value ~ s(edad_corregida_meses), method = "REML", data = .SD[sample(.N, replace = TRUE)])
+
+    # Estimation from bootstrapped model
+    predict(
+      object = mod,
+      newdata = pred_data
+    )
+  }, FUN.VALUE = numeric(length = nrow(pred_data)))
+
+  # Join the exposure and response
+  cbind(pred_data, boot)
+
+}, by = variable] # Grouped by variable
+
[…]

### 2022-12-14 manuscript/CICYPE/resources/code/risk-effects.R

@@ -1,0 +1,105 @@
+# Preparamos el ambiente de trabajo ---------------------------------------
+
+## Script de preparación
+source("manuscript/CICYPE/prep-workspace.R")
+
+
+# Preparamos los datos ----------------------------------------------------
+
+ind <- grep("interpretacion", names(dataset), value = TRUE)
+names(ind) <- c("CM", "GM", "FM", "CG", "PS")
+
+plot_risk_effects <- function(var, var_name = var, save = FALSE) {
+  form <- as.formula(
+    paste0(ind[[var]], " ~ ",
+           "s(profesional_id, bs = \"re\") +
+            s(sexo_paciente, bs = \"re\") +
+            s(respondedor_vinculo, bs = \"re\") +",
+           "s(edad_corregida_meses)")
+  )
+
+  set.seed(1234)
+  output <- mgcv::gam(form, family = stats::binomial(link = "logit"), data = dataset,
+                      method = "REML")
+
+  format_p <- function(x) {
+    x <- data.table::fifelse(
+      test = x < 0.001,
+      yes = "< 0.001",
+      no = paste("=", round(x, 3))
+    )
+
+    return(x)
+  }
+
+  p_val <- paste("p", format_p(x = stats::anova(output)$s.table[4L, 4L]))
+
+  testdata <- expand.grid(
+    profesional_id = c(2),
+    sexo_paciente = c("F"),
+    respondedor_vinculo = c("Madre"),
+    edad_corregida_meses = seq(0, 48, 0.05)
+  )
+
+  predicted_response <- as.data.table(x = stats::predict(output, testdata, se.fit = TRUE, type = "response"))
+  predicted_response <- cbind(predicted_response, testdata)
+
+  names(predicted_response)[1:2] <- c("Fit", "SE")
+
+  ylab <- "P(Categoría | Edad corregida)"
+  if (!is.null(var_name) && is.character(var_name)) {
+    ylab <- paste0("P(", paste("Retraso en", var_name), " | Edad corregida)")
+  }
+
+  ci_prob <- function(p) stats::qnorm((p+1)/2, lower.tail = TRUE)
+
+  plot <- ggplot2::ggplot(predicted_response, ggplot2::aes(edad_corregida_meses, Fit)) +
[…]

### 2022-12-14 manuscript/CICYPE/resources/code/table-1.R

@@ -1,0 +1,51 @@
+
+# Load packages and scripts -------------------------------------------------------------------
+
+source("manuscript/CICYPE/prep-workspace.R")
+
+# Table 1 -------------------------------------------------------------------------------------
+
+tbl_data <- copy(dataset)[, sexo_paciente := `levels<-`(sexo_paciente, c("Mujer", "Hombre"))]
+
+gtsummary::theme_gtsummary_language("es")
+
+tbl1 <- tbl_summary(tbl_data,
+                    by = "sexo_paciente",
+                    include = c(
+                      "semanas_prematurez",
+                      "edad_cronologica_meses",
+                      "edad_corregida_meses",
+                      "comunicacion_total",
+                      "motora_gruesa_total",
+                      "motora_fina_total",
+                      "resolucion_problemas_total",
+                      "socio_individual_total",
+                    ),
+                    label = list(
+                      semanas_prematurez ~ "Prematuridad (semanas)",
+                      edad_cronologica_meses ~ "Edad cronológica",
+                      edad_corregida_meses ~ "Edad corregida",
+                      comunicacion_total ~ "Puntaje CM",
+                      motora_gruesa_total ~ "Puntaje GM",
+                      motora_fina_total ~ "Puntaje FM",
+                      resolucion_problemas_total ~ "Puntaje CG",
+                      socio_individual_total ~ "Puntaje PS"
+                    ), digits = everything() ~ 1)
+
+tbl1 <- add_overall(x = tbl1)
+
+tbl1 <- add_p(
+  x = tbl1,
+  test = everything() ~ "wilcox.test",
+  pvalue_fun = function(x) round(x, digits = 3)
+)
+
+# tbl1 <- modify_header(
+#   x = tbl1,
+#   all_stat_cols(stat_0 = FALSE) ~ "**{level}**<br>N = {n}",
+#   stat_0 = "**Overall**<br>N = {N}"
+# )
+
+tbl1 <- as_gt(tbl1)
+
+saveRDS(tbl1, file = "manuscript/CICYPE/resources/table-1.RDS")

### 2022-12-14 manuscript/CICYPE/resources/code/table-2.R

@@ -1,0 +1,34 @@
+# Load packages and scripts -------------------------------------------------------------------
+
+source("manuscript/CICYPE/prep-workspace.R")
+
+y_vars <- c("CM" = "comunicacion_total"
+            , "FM" = "motora_fina_total"
+            , "GM" = "motora_gruesa_total"
+            , "CG" = "resolucion_problemas_total"
+            , "PS" = "socio_individual_total")
+
+many_models <- sapply(y_vars, function(i) {
+  form <- as.formula(
+    paste0(i, " ~ ",
+           "s(profesional_id, bs = \"re\") +
+            s(sexo_paciente, bs = \"re\") +
+            s(respondedor_vinculo, bs = \"re\") +",
+           "s(semanas_prematurez, k = 9)")
+  )
+  model <- gam(
+    form,
+    data = dataset,
+    method = "REML"
+  )
+  report_smooth(model)
+}, simplify = FALSE)
+
+tbl_data <- rbindlist(many_models, idcol = "Variable")
+
+tbl_data <- tbl_data[term %like% "semanas_prematurez", list(Variable, Estadístico = expr)]
+
+tbl <- knitr::kable(tbl_data, format = "simple")
+
+
+saveRDS(tbl, file = "manuscript/CICYPE/resources/table-2.RDS")

### 2022-12-14 manuscript/CICYPE/resources/code/table-3.R

@@ -1,0 +1,34 @@
+# Load packages and scripts -------------------------------------------------------------------
+
+source("manuscript/CICYPE/prep-workspace.R")
+
+y_vars <- c("CM" = "comunicacion_total"
+            , "FM" = "motora_fina_total"
+            , "GM" = "motora_gruesa_total"
+            , "CG" = "resolucion_problemas_total"
+            , "PS" = "socio_individual_total")
+
+many_models <- sapply(y_vars, function(i) {
+  form <- as.formula(
+    paste0(i, " ~ ",
+           "s(profesional_id, bs = \"re\") +
+            s(sexo_paciente, bs = \"re\") +
+            s(respondedor_vinculo, bs = \"re\") +",
+           "s(edad_corregida_meses)")
+  )
+  model <- gam(
+    form,
+    data = dataset,
+    method = "REML"
+  )
+  report_smooth(model)
+}, simplify = FALSE)
+
+tbl_data <- rbindlist(many_models, idcol = "Variable")
+
+tbl_data <- tbl_data[term %like% "edad_corregida_meses", list(Variable, Estadístico = expr)]
+
+tbl <- knitr::kable(tbl_data, format = "pipe")
+
+
+saveRDS(tbl, file = "manuscript/CICYPE/resources/table-3.RDS")

### 2022-12-14 manuscript/CICYPE/resources/code/table-main-effects.R

@@ -1,0 +1,39 @@
+# Load packages and scripts -------------------------------------------------------------------
+
+
+# Cargamos paquetes -------------------------------------------------------
+
+source("manuscript/CICYPE/prep-workspace.R")
+
+
+# Ajustamos modelos -------------------------------------------------------
+
+table_main_effects <- function(var) {
+  y_vars <- c("CM" = "comunicacion_total"
+              , "GM" = "motora_gruesa_total"
+              , "FM" = "motora_fina_total"
+              , "CG" = "resolucion_problemas_total")
+  form <- as.formula(
+    paste0(y_vars[[var]], " ~ ",
+           "s(profesional_id, bs = \"re\") +
+            s(sexo_paciente, bs = \"re\") +
+            s(respondedor_vinculo, bs = \"re\") +",
+           "s(edad_corregida_meses)")
+  )
+  set.seed(123)
+  model <- gam(
+    form,
+    data = dataset,
+    method = "REML"
+  )
+  slopes <- report_slopes(model, "edad_corregida_meses")
+  slopes <- slopes[i = as.numeric(substrRight(expr, 5)) < 0.05,
+                   j = list(Inicio = Start, Termino = End, Estadístico = expr)]
+  tbl <- knitr::kable(slopes, format = "pipe")
+  return(tbl)
+}
+
+for (i in names(y_vars)) {
+  saveRDS(table_main_effects(i),
+          file = paste0("manuscript/CICYPE/resources/table-main-",i,".RDS"))
+}

### 2022-12-14 manuscript/CICYPE/resources/code/table-risk-effects.R

@@ -1,0 +1,66 @@
+# Cargamos paquetes -------------------------------------------------------
+
+source("manuscript/CICYPE/prep-workspace.R")
+
+# Términos Smooth -------------------------------------------------------
+
+y_vars <- grep("interpretacion", names(dataset), value = TRUE)
+names(y_vars) <- c("CM", "GM", "FM", "CG", "PS")
+
+many_models <- sapply(y_vars, function(i) {
+  form <- as.formula(
+    paste0(i, " ~ ",
+           "s(profesional_id, bs = \"re\") +
+            s(sexo_paciente, bs = \"re\") +
+            s(respondedor_vinculo, bs = \"re\") +",
+           "s(edad_corregida_meses)")
+  )
+  model <- gam(
+    form,
+    data = dataset,
+    family = stats::binomial(link = "logit"),
+    method = "REML"
+  )
+  report_smooth(model)
+}, simplify = FALSE)
+
+tbl_data <- rbindlist(many_models, idcol = "Variable")
+tbl_data <- tbl_data[term %like% "edad_corregida_meses", list(Variable, Estadístico = expr)]
+
+tbl <- knitr::kable(tbl_data, format = "pipe")
+saveRDS(tbl, file = "manuscript/CICYPE/resources/table-risk-smooth.RDS")
+
+
+# Slopes ------------------------------------------------------------------
+
+table_risk_effects <- function(var) {
+  ind <- grep("interpretacion", names(dataset), value = TRUE)
+  names(ind) <- c("CM", "GM", "FM", "CG", "PS")
+  form <- as.formula(
+    paste0(ind[[var]], " ~ ",
+           "s(profesional_id, bs = \"re\") +
+            s(sexo_paciente, bs = \"re\") +
+            s(respondedor_vinculo, bs = \"re\") +",
+           "s(edad_corregida_meses)")
+  )
+
+  set.seed(1234)
+  model <- mgcv::gam(form, family = stats::binomial(link = "logit"), data = dataset)
+
+  slopes <- report_slopes(model, "edad_corregida_meses")
+  if(is.atomic(slopes)) {
+    slopes <- data.table(Estadístico = slopes)
+  } else{
+    slopes <- slopes[i = as.numeric(substrRight(expr, 5)) < 0.05,
+                     j = list(Inicio = Start, Termino = End, Estadístico = expr)]
+  }
+
[…]

### 2022-12-14 manuscript/figures/fig-3.R

@@ -13,7 +13,7 @@ names(ind) <- c("CM", "GM", "FM", "CG", "PS")
 
 
 plots <- mapply(
-  FUN = gam_ordinal,
+  FUN = gam_binomial,
   data = list(dataset),
   var = ind,
   var_name = names(ind),
@@ -36,7 +36,7 @@ local({
 # Evaluacion de los modelos -----------------------------------------------
 
 models <- mapply(
-  FUN = gam_ordinal,
+  FUN = gam_binomial,
   var = ind,
   data = list(dataset),
   plot = FALSE,

### 2022-12-14 manuscript/CICYPE/presentation.qmd
minor updates on CICYPE presentation

@@
 
-## Modelado con GAMs {.smaller}
-
-::: {.fragment .fade-out fragment-index=5}
-![](resources/gams_vs_lm.svg){.fragment .fade-in .absolute right=0 top=10% fragment-index=1}
-:::
+## Modelos Aditivos Generalizados {.smaller}
 
 :::: {.fragment .fade-out fragment-index=4}
+![](resources/gams_vs_lm.svg){.fragment .fade-in .absolute right=5% top=15% width=50% fragment-index=1}
+
 ::: {.fragment .fade-in-then-semi-out .absolute top=20% width=40% fragment-index=2}
@@
 
-:::: {.fragment .fade-out fragment-index=9}
+::::: {.fragment .fade-out fragment-index=6}
 
-![](resources/simpson_paradox.png){.fragment .fade-in .absolute left=20% top=20% width=60% fragment-index=6}
+:::: {.fragment .fade-in fragment-index=5}
+![](resources/simpson_paradox.png){.absolute left=20% top=20% width=60%}
 
-::: {.fragment .fade-in .absolute top=25% right=0 width=20% fragment-index=7 style="font-size: 20px; color: gray;"}
+::: {.absolute top=25% right=0 width=20% style="font-size: 20px; color: gray;"}
 *Paradoja de simpson*
@@
 
-::: {.fragment .fade-in .absolute top=55% left=27% width=50% fragment-index=8}
+::: {.absolute top=55% left=27% width=50%}
 Sumado a lo anterior, también podemos controlar por la influencia de otras variables para obtener estimaciones más robustas a factores de confusión.
 :::
+
 ::::
 
-::: {.fragment .fade-in .absolute top=30% left=25% width=50% fragment-index=10, style="font-size:30px;"}
+:::::
+
+::: {.fragment .fade-in .absolute top=30% left=25% width=50% fragment-index=7 style="font-size:30px;"}
 Para describir las estimaciones de los modelos usaremos derivadas aproximadas de primer orden con intervalos de confianza al 95%, con el fin de poder describir los efectos observados en términos de segmentos quasi-lineales [@dominique2020estimation].
@@
 ::::{.fragment .fade-in-then-out}
-![](resources/results_risk_CM.svg){.absolute left=5% width=90% top=0}
+![](resources/results_risk_CM.svg){.absolute left=5% width=80% top=0}
 
-:::{.absolute top=70% left=10% width=80% style="font-size: 22px;"}
+:::{.absolute top=70% left=10% width=75% style="font-size: 22px;"}
 ```{r}
@@
 ::::{.fragment .fade-in-then-out}
-![](resources/results_risk_GM.svg){.absolute left=5% width=90% top=0}
+![](resources/results_risk_GM.svg){.absolute left=5% width=80% top=0}
 
-:::{.absolute top=70% left=10% width=80% style="font-size: 22px;"}
+:::{.absolute top=70% left=10% width=75% style="font-size: 22px;"}
 ```{r}
@@
 ::::{.fragment .fade-in-then-out}
-![](resources/results_risk_FM.svg){.absolute left=5% width=90% top=0}
-
-:::{.absolute top=70% left=10% width=80% style="font-size: 22px;"}
-```{r}
-readRDS("resources/table-risk-FM.RDS")
-```
-:::
-::::
-
-::::{.fragment .fade-in-then-out}
-![](resources/results_risk_CG.svg){.absolute left=5% width=90% top=0}
-
-:::{.absolute top=70% left=10% width=80% style="font-size: 22px;"}
-```{r}
-readRDS("resources/table-risk-CG.RDS")
-```
-:::
-::::
-
-::::{.fragment .fade-in-then-out}
-![](resources/results_risk_PS.svg){.absolute left=5% width=90% top=0}
-
-:::{.absolute top=70% left=10% width=80% style="font-size: 22px;"}
-```{r}
-readRDS("resources/table-risk-PS.RDS")
-```
-:::
+![](resources/results_risk_NS.svg){.absolute left=5% width=90% top=10%}
 ::::

### 2022-12-14 manuscript/CICYPE/resources/code/risk-effects.R

@@ -90,7 +90,7 @@ plot_risk_effects <- function(var, var_name = var, save = FALSE) {
 
   if (save) {
     path <- gsub("%var%", var, "manuscript/CICYPE/resources/results_risk_%var%.svg", fixed = TRUE)
-    svglite::svglite(filename = path, height = 4, width = 8, always_valid = TRUE)
+    svglite::svglite(filename = path, height = 4, width = 7, always_valid = TRUE)
     print(plot)
     dev.off()
   } else {
@@ -98,8 +98,19 @@ plot_risk_effects <- function(var, var_name = var, save = FALSE) {
   }
 }
 
-for (i in names(ind)) {
+for (i in names(ind[1:2])) {
   plot_risk_effects(i, save = TRUE)
 }
 
+fig <- ggpubr::ggarrange(
+  plotlist = lapply(names(ind[3:5]), plot_risk_effects),
+  nrow = 1,
+  ncol = 3,
+  common.legend = TRUE
+)
+svglite::svglite(filename = "manuscript/CICYPE/resources/results_risk_NS.svg",
+                 height = 5, width = 10,
+                 always_valid = TRUE)
+print(fig)
+dev.off()
 

### 2022-12-15 manuscript/CICYPE/resources/code/risk-effects.R
update presentation based on CN suggestions

@@ -98,19 +98,7 @@ plot_risk_effects <- function(var, var_name = var, save = FALSE) {
   }
 }
 
-for (i in names(ind[1:2])) {
+for (i in names(ind)) {
   plot_risk_effects(i, save = TRUE)
 }
-
-fig <- ggpubr::ggarrange(
-  plotlist = lapply(names(ind[3:5]), plot_risk_effects),
-  nrow = 1,
-  ncol = 3,
-  common.legend = TRUE
-)
-svglite::svglite(filename = "manuscript/CICYPE/resources/results_risk_NS.svg",
-                 height = 5, width = 10,
-                 always_valid = TRUE)
-print(fig)
-dev.off()
 

### 2022-12-15 manuscript/CICYPE/presentation.qmd
update presentation based on CN suggestions

@@
 author:
-  - Matías Castillo-Aguilar ^1,2^
-institute: 
-  - ^1^ Departamento de Kinesiología, Universidad de Magallanes, Punta Arenas, Chile.
-  - ^2^ Neurofisiología Integrativa Molecular Austral Chilena (NIM-ACh), Punta Arenas, Chile.
+  - 'Patricio Barría, [Matías Castillo-Aguilar]{style="color:yellow;"}, Katherine Harris, Kay Gittermann, Mónica Bontes, Rolando Aguilar, Carlos A. Cifuentes, Marcela Munera, Cristian Núñez-Espinosa.'
 format: 
@@
     theme: simple
-    toc: true
-    toc-depth: 1
-    toc-title: Contenidos
     transition: slide
-    background-transition: slide
+    background-transition: fade
     controls-layout: bottom-right
@@
     view-distance: 3
-editor: visual
 footer: "Matías Castillo-Aguilar"
@@
 
-## {background-color="black" background-video="resources/group_discussion.mp4" background-video-loop="true" background-video-muted="true" background-opacity=0.5 background-transition="fade"}
+## {background-color="black" background-video="resources/end_intro.mp4" background-video-loop="true" background-video-muted="true" background-opacity=0.5 background-transition="fade"}
 
@@
 ### Específicos
+
 :::{style="font-size: 28px;"}
-- Describir y caracterizar la relación entre el [neurodesarrollo y la prematuridad]{.fragment .highlight-red} en
-niños diagnosticados de HC.
-- Describir y caracterizar la relación entre el [neurodesarrollo y la edad]{.fragment .highlight-red} en niños
-diagnosticados de HC.
-- Describir y caracterizar [la influencia de la edad sobre el riesgo de presentar un
-retardo en el neurodesarrollo]{.fragment .highlight-red} en niños diagnosticados de HC.
+Describir y caracterizar...
+
+- la relación entre el [neurodesarrollo y la prematuridad]{.fragment .highlight-red},
+- la relación entre el [neurodesarrollo y la edad]{.fragment .highlight-red},
+- [la influencia de la edad sobre el riesgo de presentar un
+retardo en el neurodesarrollo]{.fragment .highlight-red},
+
+...en niños diagnosticados de HC.
 :::
@@
 
-::: {.fragment .fade-in .absolute width=45% right=0 top=36% style="font-size: 34px;"}
-Una vez hecho el diagnóstico, este se refirió al programa de la institución para su posterior ingreso.
-:::
-
 <script>
@@
 
-## Resguardos éticos {.smaller}
-
-![](resources/ethics.mp4){.absolute #ethics right=0 width=50% top=30% loop=true}
-
-::: {.absolute width=45% top=30%}
-- Los cuidadores legales otorgaron su consentimiento informado antes de la participación a este estudio.
-
-- El Comité de Ética del CRCS aprobó este estudio siguiendo las regulaciones establecidas por la Declaración de Helsinki.
-:::
-
-<script>
-var video = document.getElementById("ethics");
-video.controls = false;
-</script>
-
-## Mediciones
+## The Ages and Stages Questionnaire
 
@@
 
-::: {.fragment .fade-in-then-out .absolute width=45% right=0 top=22% style="font-size: 32px;"}
-### The Ages and Stages Questionnaire (ASQ-3)
-
+::: {.fragment .fade-in-then-out .absolute width=45% right=0 top=32% style="font-size: 32px;"}
 - Instrumento de cribado del desarrollo de nivel inicial informado por los padres.
-- Consta de 21 intervalos, cada uno con 30 ítems en cinco áreas.
+- Consta de cinco áreas del desarrollo evaluadas.
 :::
@@
 ::::
-
-::: {.fragment .fade-in-then-out .absolute width=45% right=0 top=28% style="font-size: 28px;"}
-Es costo efectivo y ampliamente usado internacionalmente [@heo2008cross; @sarmiento2011universal].
-
-Validado en infantes de término y pretérmino chilenos (75% sensibilidad and 81% especificidad) [@schonhaut2013validity].
-:::
 
@@
 
-::: {.fragment .fade-in .absolute width=45% left=0 top=22% style="font-size: 32px;" fragment-index=4}
-Algunas consideraciones importantes de la administración son que:
+# Resultados
 
-1. Debe responder el [cuidador principal]{.fragment .highlight-red}.
-2. En caso de dudas del cuidador sobre el comportamiento evaluado, la información se corrobora mediante pruebas *in-situ* con el usuario.
+¿Qué fué lo que se encontró?
+
+## Características de la muestra
+
+:::{.absolute top=12% left=10% width=80% style="font-size:16px;"}
+```{r}
+readRDS("resources/table-1.RDS")
+```
 :::
 
-# Análisis estadístico
+## {background-color="white" background-video="resources/in_results_2.mp4" background-video-loop="true" background-video-muted="true" background-opacity=0.5 background-transition="fade"}
+
+:::{style="font-weight: 1000; text-align: center; font-size: 80px;"}
+[NEURODESARROLLO]{.fragment top=35% left=0 .absolute}
+
+[Y LA EDAD]{.fragment top=50% left=0 .absolute}
+:::
+
+## Comunicación
+
+![](resources/results_main_CM.svg){.absolute left=5% width=85% top=10%}
+
+:::{.absolute top=80% left=10% width=80% style="font-size: 22px;"}
+```{r}
+readRDS("resources/table-main-CM.RDS")
+```
+:::
+
+## Habilidades motoras gruesas
+
+![](resources/results_main_GM.svg){.absolute left=5% width=85% top=10%}
+
+:::{.absolute top=80% left=10% width=80% style="font-size: 22px;"}
+```{r}
+readRDS("resources/table-main-GM.RDS")
+```
+:::
+
+## Habilidades motoras finas
+
+![](resources/results_main_FM.svg){.absolute left=5% width=85% top=10%}
+
+:::{.absolute top=80% left=10% width=80% style="font-size: 22px;"}
+```{r}
+readRDS("resources/table-main-FM.RDS")
+```
+:::
+
+## Resolución de problemas
+
+![](resources/results_main_CG.svg){.absolute left=5% width=85% top=10%}
+
+:::{.absolute top=80% left=10% width=80% style="font-size: 22px;"}
+```{r}
+readRDS("resources/table-main-CG.RDS")
+```
+:::
+
+## Personal-social
+
+![](resources/results_main_PS.svg){.absolute left=5% width=85% top=10%}
+
+:::{.absolute top=80% left=10% width=80% style="font-size: 22px;"}
+```{r}
+readRDS("resources/table-main-PS.RDS")
+```
+:::
+
+## {background-color="white" background-video="resources/in_results_2.mp4" background-video-loop="true" background-video-muted="true" background-opacity=0.5 background-transition="fade"}
+
+[EDAD]{.absolute .fragment top=35% left=0 style="font-weight: 1000; font-size: 80px;"}
+
+[Y RETRASO DEL DESARROLLO]{.absolute .fragment top=50% left=0 style="font-weight: 1000; font-size: 70px;"}
+
+## Comunicación
+
+![](resources/results_risk_CM.svg){.absolute left=10% width=80% top=10%}
+
+:::{.absolute top=80% left=15% width=75% style="font-size: 22px;"}
+```{r}
+readRDS("resources/table-risk-CM.RDS")
+```
+:::
+
+## Habilidades motoras gruesas
+
+![](resources/results_risk_GM.svg){.absolute left=10% width=80% top=10%}
+
+:::{.absolute top=80% left=15% width=75% style="font-size: 22px;"}
+```{r}
+readRDS("resources/table-risk-GM.RDS")
+```
+:::
+
+## Habilidades motoras finas
+
+![](resources/results_risk_FM.svg){.absolute left=10% width=80% top=10%}
+
+:::{.absolute top=80% left=15% width=75% style="font-size: 22px;"}
+```{r}
+readRDS("resources/table-risk-FM.RDS")
+```
+:::
+
+## Resolución de problemas
+
+![](resources/results_risk_CG.svg){.absolute left=10% width=80% top=10%}
+
+:::{.absolute top=80% left=15% width=75% style="font-size: 22px;"}
+```{r}
+readRDS("resources/table-risk-CG.RDS")
+```
+:::
+
+## Personal-social
+
+![](resources/results_risk_PS.svg){.absolute left=10% width=80% top=10%}
+
+:::{.absolute top=80% left=15% width=75% style="font-size: 22px;"}
+```{r}
+readRDS("resources/table-risk-PS.RDS")
+```
+:::
+
+# Conclusión
+
+## {background-color="black" background-video="resources/conclusion.mp4" background-video-loop="true" background-video-muted="true" background-opacity=0.5 background-transition="fade"}
+
+:::: {.fragment .fade-out fragment-index=3}
+::: {.fragment .fade-in-then-semi-out .absolute width=40% left=0 top=20% style="font-size: 28px;" fragment-index=1}
+- Se observan marcadas variaciones no lineales en los rasgos del neurodesarrollo en respuesta de la edad.
+:::
+::::
+
+::: {.fragment .fade-in-then-out .absolute width=40% left=0 top=45% style="font-size: 28px;" fragment-index=2}
+- Las variaciones observadas en el neurodesarrollo no son atribuibles únicamente a la prematuridad.
+:::
+
+::: {.fragment .fade-in .absolute width=30% left=0 top=22% style="font-size: 28px;" fragment-index=4}
+Futuras investigaciones podrían determinar cómo se aplican estos resultados cuando se controlan los factores sociales mediados por el contexto, así como otras poblaciones clínicas.
+:::
+
+## {background-color="black" background-image="resources/title-background.jpeg" background-opacity=0.5 background-transition="fade"}
+
+::::{.absolute top=21%}
+:::{style="text-align: center;"}
+<h1 class="title">Descripción y caracterización del neurodesarrollo de niños con hipotonía congénita</h1>
+:::
+:::{.subtitle style="text-align: center;"}
+de la región de Magallanes y la Antártica Chilena
+:::
+::::
+
+:::{.quarto-title-author-name .absolute top=50% style="text-align:center;"}
+Patricio Barría, [Matías Castillo-Aguilar]{style="color:yellow;"}, Katherine Harris, Kay Gittermann, Mónica Bontes, Rolando Aguilar, Carlos A. Cifuentes, Marcela Munera, Cristian Núñez-Espinosa. 
+:::
+
+## Referencias {.smaller}
+
+<div id="refs" class="references csl-bib-body hanging-indent" role="doc-bibliography"></div>
+
+# Anexos
+
+## Resguardos éticos {.smaller}
+
+![](resources/ethics.mp4){.absolute #ethics right=0 width=50% top=30% loop=true}
+
+::: {.absolute width=45% top=30%}
+- Los cuidadores legales otorgaron su consentimiento informado antes de la participación a este estudio.
+
+- El Comité de Ética del CRCS aprobó este estudio siguiendo las regulaciones establecidas por la Declaración de Helsinki.
+:::
+
+<script>
+var video = document.getElementById("ethics");
+video.controls = false;
+</script>
+
+## Análisis exploratorio
 
@@
 ```
-
-## Software estadístico
-
-![](resources/rstudio.png){.absolute width=40% left=10% top=30%}
-
-![](resources/rstudio-desktop-screen.png){.absolute width=40% right=10% top=12%}
-
-## Análisis exploratorio
 
@@
 
-# Resultados
-
-¿Qué fué lo que se encontró?
-
-## Características de la muestra
-
-:::{.absolute top=12% left=10% width=80% style="font-size:16px;"}
-```{r}
-readRDS("resources/table-1.RDS")
-```
-:::
-
-## {background-color="white" background-video="resources/in_results.mp4" background-video-loop="true" background-video-muted="true" background-opacity=0.5 background-transition="fade"}
-
-:::{style="font-weight: 1000; text-align: center; font-size: 80px;"}
-[NEURODESARROLLO]{.fragment top=35% left=0 .absolute}
-
-[Y PREMATURIDAD]{.fragment top=50% left=0 .absolute}
-:::
-
----
+## Prematurez y neurodesarrollo
 
@@
 ::: {.absolute top=25% right=0% width=30% style="font-size: 20px; color: gray;"}
-*Tabla 1. Significancia de los términos no-lineales para la prematurez sobre las variables del desarrollo. Se presentan los modelos ajustados por las variables de confusión previamente descritas.*
+*Tabla 1S. Significancia de los términos no-lineales para la prematurez sobre las variables del desarrollo. Se presentan los modelos ajustados por las variables de confusión previamente descritas.*
 :::
 
-## {background-color="white" background-video="resources/in_results_2.mp4" background-video-loop="true" background-video-muted="true" background-opacity=0.5 background-transition="fade"}
+## Neurodesarrollo y edad corregida
 
-:::{style="font-weight: 1000; text-align: center; font-size: 80px;"}
-[NEURODESARROLLO]{.fragment top=35% left=0 .absolute}
-
-[Y LA EDAD]{.fragment top=50% left=0 .absolute}
-:::
-
----
-
-::::{.fragment .fade-in-then-out}
 :::{.absolute top=20% left=0 width=75% style="font-size: 30px;"}
@@
 ::: {.absolute top=25% right=5% width=20% style="font-size: 20px; color: gray;"}
-*Tabla 2.
+*Tabla 2S.
 Significancia de los términos no-lineales para la edad corregida (meses) sobre las variables del desarrollo.
@@
 :::
-::::
 
-::::{.fragment .fade-in-then-out}
-![](resources/results_main_CM.svg){.absolute left=5% width=90% top=0}
+## Retardo del desarrollo y edad corregida
 
-:::{.absolute top=70% left=10% width=80% style="font-size: 22px;"}
-```{r}
-readRDS("resources/table-main-CM.RDS")
-```
-:::
-::::
-
-::::{.fragment .fade-in-then-out}
-![](resources/results_main_GM.svg){.absolute left=5% width=90% top=0}
-
-:::{.absolute top=70% left=10% width=80% style="font-size: 22px;"}
-```{r}
-readRDS("resources/table-main-GM.RDS")
-```
-:::
-::::
-
-::::{.fragment .fade-in-then-out}
-![](resources/results_main_FM.svg){.absolute left=5% width=90% top=0}
-
-:::{.absolute top=70% left=10% width=80% style="font-size: 22px;"}
-```{r}
-readRDS("resources/table-main-FM.RDS")
-```
-:::
-::::
-
-::::{.fragment .fade-in-then-out}
-![](resources/results_main_CG.svg){.absolute left=5% width=90% top=0}
-
-:::{.absolute top=70% left=10% width=80% style="font-size: 22px;"}
-```{r}
-readRDS("resources/table-main-CG.RDS")
-```
-:::
-::::
-
-::::{.fragment .fade-in-then-out}
-![](resources/results_main_PS.svg){.absolute left=5% width=90% top=10%}
-::::
-
-## {background-color="white" background-video="resources/in_results_2.mp4" background-video-loop="true" background-video-muted="true" background-opacity=0.5 background-transition="fade"}
-
-:::{style="font-weight: 1000; font-size: 80px;"}
-[EDAD]{.fragment top=20% left=0 .absolute}
-
-[Y RETRASO DEL DESARROLLO]{.fragment top=35% left=0 .absolute}
-:::
-
----
-
-::::{.fragment .fade-in-then-out}
 :::{.absolute top=20% left=0 width=80% style="font-size: 30px;"}
@@
 ::: {.absolute top=25% right=5% width=20% style="font-size: 20px; color: gray;"}
-*Tabla 7.
+*Tabla 3S.
 Significancia de los términos no-lineales logísticos para la edad corregida (meses) sobre las variables del desarrollo.
@@
 :::
-::::
-
----
-
-::::{.fragment .fade-in-then-out}
-![](resources/results_risk_CM.svg){.absolute left=5% width=80% top=0}
-
-:::{.absolute top=70% left=10% width=75% style="font-size: 22px;"}
-```{r}
-readRDS("resources/table-risk-CM.RDS")
-```
-:::
-::::
-
-::::{.fragment .fade-in-then-out}
-![](resources/results_risk_GM.svg){.absolute left=5% width=80% top=0}
-
-:::{.absolute top=70% left=10% width=75% style="font-size: 22px;"}
-```{r}
-readRDS("resources/table-risk-GM.RDS")
-```
-:::
-::::
-
-::::{.fragment .fade-in-then-out}
-![](resources/results_risk_NS.svg){.absolute left=5% width=90% top=10%}
-::::
-
-# Conclusión
-
-## {background-color="black" background-video="resources/conclusion.mp4" background-video-loop="true" background-video-muted="true" background-opacity=0.5 background-transition="fade"}
-
-:::: {.fragment .fade-out fragment-index=3}
-::: {.fragment .fade-in-then-semi-out .absolute width=50% left=0 top=5% style="font-size: 32px;" fragment-index=1}
-- Se observan marcadas variaciones en los rasgos del neurodesarrollo a lo largo de la edad en los infantes hipotónicos, principalmente en forma de variaciones no lineales y de dominio específico.
-:::
-::::
-
-::: {.fragment .fade-in-then-out .absolute width=50% left=0 top=45% style="font-size: 30px;" fragment-index=2}
-- Mostramos que las variaciones observadas en los dominios del desarrollo no son atribuibles únicamente a la prematuridad, donde la edad corregida explica mejor la variabilidad observada en el neurodesarrollo.
-:::
-
-::: {.fragment .fade-in .absolute width=45% left=0 top=22% style="font-size: 32px;" fragment-index=4}
-Futuras investigaciones podrían determinar cómo se aplican estos resultados cuando se controlan los factores sociales mediados por el contexto, así como otras poblaciones clínicas.
-:::
-
-# Referencias
-
-## {.smaller}

### 2022-12-15 manuscript/CICYPE/resources/code/table-main-effects.R

@@ -12,7 +12,8 @@ table_main_effects <- function(var) {
   y_vars <- c("CM" = "comunicacion_total"
               , "GM" = "motora_gruesa_total"
               , "FM" = "motora_fina_total"
-              , "CG" = "resolucion_problemas_total")
+              , "CG" = "resolucion_problemas_total"
+              , "PS" = "socio_individual_total")
   form <- as.formula(
     paste0(y_vars[[var]], " ~ ",
            "s(profesional_id, bs = \"re\") +
@@ -27,8 +28,14 @@ table_main_effects <- function(var) {
     method = "REML"
   )
   slopes <- report_slopes(model, "edad_corregida_meses")
-  slopes <- slopes[i = as.numeric(substrRight(expr, 5)) < 0.05,
-                   j = list(Inicio = Start, Termino = End, Estadístico = expr)]
+  if(is.atomic(slopes)) {
+    slopes <- data.table(Estadístico = slopes)
+  } else{
+    slopes <- slopes[i = as.numeric(substrRight(expr, 5)) < 0.05,
+                     j = list(Inicio = Start, Termino = End, Estadístico = expr)]
+  }
+
+  slopes[, exp(as.numeric(gsub(",", "", substr(Estadístico, 10, 16))))]
   tbl <- knitr::kable(slopes, format = "pipe")
   return(tbl)
 }
@@ -37,3 +44,4 @@ for (i in names(y_vars)) {
   saveRDS(table_main_effects(i),
           file = paste0("manuscript/CICYPE/resources/table-main-",i,".RDS"))
 }
+

### 2023-04-23 manuscript/captions.Rmd
update on paper files, going for the third try!

--- 2022-11-06 manuscript/manuscript.Rmd
@@
 title: ""
-bibliography: "misc/ref.bib"
-csl: "misc/vancouver-dmcn.csl"
-link-citations: yes
-notes-after-punctuation: true
 output:
@@
     reference_docx: "misc/ref.docx"
-editor_options: 
-  chunk_output_type: console
 ---
-
-```{r setup, include=FALSE}
-knitr::opts_chunk$set(
-  echo = FALSE,
-  include = FALSE
-)
-```
-
-```{r load-packages}
-source(file = "misc/setup.R")
-```
-
-**Title**: Neurodevelopmental features among infants with congenital hypotonia: An observational cross-sectional study.
-
-**Authors**: Patricio Barría^1,2^, Matías Castillo-Aguilar^3,4^, Katherine Harris^3,4^, Kay Gittermann^1,5^, Mónica Bontes^1^, Rolando Aguilar^1,6^, Carlos A.
-Cifuentes^7,8^, Marcela Munera^9^, Cristian Núñez-Espinosa^4,10^*
-
-::: {custom-style="Afiliation"}
-^1^ Corporación de Rehabilitación Club de Leones Cruz del Sur, Punta Arenas, Chile.
-
-^2^ Brain-Machine Interface Systems Lab, Universidad Miguel Hernández de Elche, Elche, España.
-
-^3^ Kinesiology Department, Magallanes University, Punta Arenas, Chile.
-
-^4^ Centro Asistencial de Docencia e Investigación (CADI-UMAG), Punta Arenas Chile.
-
-^5^ Hospital Clínico de Magallanes, Punta Arenas, Chile.
-
-^6^ Engineering Department, Magallanes University, Punta Arenas, Chile.
-
-^7^ Bristol Robotics Laboratory, University of the West of England T Block, Frenchay Campus.
-
-^8^ School of Engineering, Science and Technology, Universidad del Rosario, Bogota, Colombia.
-
-^9^ Department of Biomedical Engineering, Colombian School of Engineering Julio Garavito.
-
-^10^ School of Medicine, Magallanes University, Punta Arenas, Chile.
-:::
-
-### \*Autor de correspondencia
-
-Cristian Núñez-Espinosa, School of Medicine, Magallanes University, Punta Arenas, Chile.
-Centro Asistencial de Docencia e Investigación CADI-UMAG, Chile.
-e-mail: [cristian.nunez\@umag.cl](mailto:cristian.nunez@umag.cl).
-Address: Avenida Bulnes 01855, Box 113-D.
-Phone: +56 61 2201411
-
-\newpage
-
-## Abstract
-
-**Aim**: To describe and model the relationship between sociodemographics, prematurity and neurodevelopmental levels based on the *Ages and Stages Questionnaire* (ASQ-3) scores in infants diagnosed with congenital hypotonia (CH).
-**Method**: A total of 234 patients diagnosed with CH participated in this study (median age, 8 years; 94 (40.2%) females and 140 (59.8%) males).
-Neurodevelopmental status was assessed with the ASQ-3 at admission, as well as sociodemographic and obstetric data obtained through the initial clinical interview.
-**Results**: When modelling the neurodevelopmental status of each domain, an overall negative effect of corrected age on communication and problem-solving skills was observed, whereas the overall effect tends to be positive in gross motor function.
-Fine motor skills did not exhibit a linear relationship with corrected age but a non-linear effect, unlike the personal-social domain, which did not present significant variations across age corrected for prematurity.
-**Interpretation**: After adjusting for possible confounders, we found between-subjects fluctuations in neurodevelopmental traits across age in hypotonic infants.
-These fluctuations were present in the form of non-linear and domain-specific variations.
-
-What this paper adds:
-
-   - Describes levels of variations observed in neurodevelopmental traits are present across age in hypotonic infants.
-   - Observed variations in developmental domains are not solely attributable to prematurity, where age corrected for prematurity best explains the observed variability in neurodevelopment.
-
-**Keywords**: Developmental traits, Congenital hypotonia, Ages and Stages Questionnaire (ASQ), Infants.
-
-\newpage
-
-<!--- # Introduction --->
-
-Early childhood development's physical and psychological signs are representative and relevant markers for identifying and monitoring overall growth in early life [@di2016predictive].
-Therefore, both can be used in screening children at risk of developmental delay to support early referral and identify the need for further assessment to determine if they are eligible for early intervention services [@bruder2010early; @guralnick2017early].
-
-Currently, many tools have been proposed to assess the continued development of infants.
-In this sense, the Ages and Stages Questionnaire, Third Edition (ASQ-3), has been presented as a global screening tool aimed at parents/caregivers, assessing five developmental domains in children aged 0 to 5.5 years [@singh2017ages].
-Current evidence suggests that the ASQ-3 is an accurate, cost-effective, and parent-friendly instrument for screening and monitoring children through pre-school age, helping to identify and eliminate neurodevelopmental deficits in children born very premature [@kerstjens2015ages; @ballantyne2016risk; @singh2017ages].
-
-Multiple neuromuscular, metabolic, and genetic conditions have been associated with hypotonia.
-Hypotonia is the decreased muscle tone or flaccidity and may represent a sign of delayed neurological development in the child, which may predispose to cognitive impairment in some cases [@riou2009global].
-Because the baby's growth-associated hypotonia, hypermobility, and motor delay can affect her ability to relate to her environment, critical visual cues are not always interpreted, leading to impaired learning and cognitive development [@harris2008congenital].
-Hypotonia implies a wide range and levels of muscle flaccidity, which should be explored in the cognitive development of the infant [@gabis2021weak; @harris2008congenital].
-There are multiple causes of hypotonia, one of them being congenital hypotonia (CH).
-Some authors suggest that CH cannot be referred to as a diagnosis [@thompson2002benign].
-Those who do consider it point it out as a diagnosis of exclusion in the investigation, considering it only when it is established in the absence of other signs and symptoms [@gabis2021weak; @leyenaar2005schematic].
-However, its research is fundamental since it is a non-progressive neuromuscular disorder, which tends to improve with time and early intervention [@gabis2021weak].
-
-To our knowledge, there is no substantial evidence characterizing the observed variation in developmental traits in infants with diagnosed CH across ages.
-Therefore, our primary goal in this study was to describe and model the relationship between sociodemographic characteristics, prematurity, and neurodevelopmental levels as a function of ASQ-3 estimates in infants diagnosed with CH.
-
-# Material y methods
-
-## Ethics
-
-Participating subjects gave their permission, and legal guardians provide informed consent before participation.
-The Ethics Committee approved this study of the Corporación de Rehabilitación Club de Leones Cruz del Sur, Chile (code: CorporacionRCLCS010322), following the regulations established by the Declaration of Helsinki on ethical principles in human beings.
-
-## Study design
-
-We conducted an observational, cross-sectional study under a quantitative approach.
-
-## Participants
-
-```{r, eval=FALSE}
-# Sample
-n_sample <- dataset[, .N]; 
-print(n_sample);
-#> 234
-```
-
-A total of 234 patients were enrolled as part of an intervention program carried out by the Cruz del Sur Rehabilitation Center (Punta Arenas, Chile), admitted from one month to 60 months old, and evaluated at admission, control and discharge.
-Participating patients were diagnosed based on clinical criteria confirmed by a neurologist and a physiatrist.
-The total files of the patients diagnosed with CH, and admitted to the institution's program, were analyzed, and the evaluation of the subjects was carried out and guided by a nurse trained in the application of the ASQ-3.
-
-In this context, it is worth noting that the data, as well as the patient registry, are part of institutional strategies aimed at the diagnosis and continuous monitoring of the clinical situation of users, which are used to improve care processes and clinical decision-making.
-
-The diagnosis of CH is based on four sources: 1) primary care pediatrician, who refers the diagnosis to the institution where admission to the program is made; 2) pediatricians from private clinics, who refer the diagnosis; 3) neuropediatricians from the clinical hospital, who refer the diagnosis to the program; 4) physiatrist from the institution, who assigns the diagnosis if it corresponds to the semiology.
-
-## Measures
-
-### The Ages and Stages Questionnaire, third edition (ASQ-3)
-
-The ASQ-3 is a parent-reported initial level developmental screening instrument consisting of 21 intervals, each with 30 items in five areas: i) communication (CM), ii) gross motor (GM), iii) fine motor (FM), iv) problem-solving (CG), and v) personal-social (PS) [@squires2009ages].
-The ASQ is cost-effective and widely used in the United States and other countries [@heo2008cross; @sarmiento2011universal].
-It has been translated into several languages, and international studies on its psychometric properties in diverse cultural environments are increasing [@heo2008cross; @sarmiento2011universal].
-It has shown good psychometric properties (75% sensitivity and 81% specificity) in Chilean term and preterm infants [@schonhaut2013validity].
-
-## Procedures
-
-For the collection of research data, authorization consent was obtained for the use of instrumental clinical data as well as the clinical record of each patient.
-Subsequently, and following national research regulations, written authorization was obtained from the institutional director to use the forms and database for research purposes.
-To this end, the data were anonymized during data processing and subsequent analyses.
-
-### Collection of demographic data
-
-Each patient's demographic data were collected and made available throughout the study by the institutional electronic systems at the time of entry to the program.
-The administrative registration was done by the secretaries of the user coordination unit, recording name, ID number, date of birth and diagnosis.
-In addition, age data was automatically updated by the computer system’s algorithms.
-
-### Assessment with ASQ-3
-
-The ASQ-3 was applied according to the protocol established by the instrument itself, with a face-to-face or telematic application being valid.
-The protocol can be found in the instrument’s manuals, which the program nurse administered.
-Some important considerations of the administration are that i) the primary caregiver must answer, ii) in case of doubts of the caregiver about the assessed behavior, the information is corroborated by in situ tests with the user, and iii) in relation to the correction, the test itself standardizes these procedures.
-
-## Statistical analysis
-
-Data are presented as median (*Mdn*) and interquartile range (*IQR*) for continuous variables; for categorical/discrete variables, the absolute and relative sample size was reported.
-There was no missing data.
-
-A non-parametric approach was used since the underlying distribution of continuously measured outcomes, assessed through analytical and graphical methods, did not follow a Gaussian distribution.
-
-In order to assess the differences in developmental scores between males and females, the *Wilcoxon* rank-sum test was used, meanwhile the chi-square test ($\chi^2$) was used to evaluate goodness-of-fit ($\chi^2_{gof}$).
-
-Generalized additive models (GAM) were used to describe linear and non-linear relationships in the form of smooth terms between developmental characteristics, represented through penalized regression splines [@wood2011fast].
-The restricted maximum likelihood method was used to estimate the smoothing parameters, and thin-plate regression splines as the smoothing basis, as they are the optimal smoother of any given basis dimension/rank [@wood2003thin].
-We used approximative derivatives with 95% confidence intervals (CI95%) to describe the smooth terms by means of quasi-linear segments [@dominique2020estimation].
-
-In order to assess the influence of confounding variables, we fitted different models using the variables: i) sex of the infant, ii) clinician ID and iii) the relationship of the infant to the respondent (i.e., whether the respondent was the infant's mother, father, uncle or grandparent) as random effects in the form of penalized parametric terms [@wood2016smoothing].
-Varying random effects structures models can be seen in the *random effects*' section of the supplementary file.
-
-A probability of committing a type I error ($\alpha$) of less than 5% (*p* < 0.05) was considered sufficient evidence for statistical significance in hypothesis testing.
-All the statistical analyses were computed and implemented in the R programming language [@rlanguage].
-GAMs and the corresponding model estimates were calculated using the *mgcv* and *modelbased* packages [@dominique2020estimation; @wood2017generalized].
-Complementary R packages were used for visualization purposes [@daniel2021see; @hadley2016ggplot2].
-
-# Results
-
-```{r, eval=FALSE}
-# Sample by sex
-n_sex <- dataset[j = .(stat = paste0(.N, " (", round(.N*100/n_sample, 1), "%)")), 
-                 keyby = .(sex = sexo_paciente)];
-print(n_sex);
-#>    sex        stat
-#> 1:   F  94 (40.2%)
-#> 2:   M 140 (59.8%)
-
-# Prop test on sex
-prop_sex <- dataset[, table(sexo_paciente)] |> 
-  chisq.test() |> 
-  report_chi(effsize = FALSE);
-cat(prop_sex);
-#> $\chi^2_{gof}$ (1) = 9.04, *p* = 0.003
-```
-
-From a total of 234 subjects with congenital hypotonia, 94 (40.2%) were females and 140 (59.8%) males ($\chi^2_{gof}$ (1) = 9.04, *p* = 0.003).
-The developmental characteristics of the sample can be seen in [Table 1](#tab1).
-
-When modelling the effect of chronological age on developmental domains, different random effects structures were compared when assessing model estimates, these estimates can be seen in [Figure 1](#fig1).
-
-```{r, eval=FALSE}
-set.seed(123)
-mod <- gam(
-  comunicacion_total ~ 
-    s(profesional_id, bs = "re") + 
-    s(sexo_paciente, bs = "re") + 
-    s(respondedor_vinculo, bs = "re") + 
-    s(edad_corregida_meses), 
-  data = dataset, 
-  method = "REML"
-)
-
-report_smooth(mod);
-report_overall(mod, "edad_corregida_meses") |> cat();
-report_slopes(mod, "edad_corregida_meses");
-```
-
-After fitting a GAM to assess the effect of corrected age on CM scores, we observed a significant non-linear relationship between them ($F_{smooth}$ (5.1, 224.74) = 13.88, *p* < 0.001), that reflect an overall negative marginal trend ($\beta$ = -2.39, CI~95%~[-3.46, -1.32], $t_{student}$ (224.74) = -4.4, *p* < 0.001), however, this was not true when assessing the direction of the effect in the age range between 0 to 6.8 ($\beta$ = 0.38, CI~95%~[-0.96, 1.72], $t_{student}$ (224.74) = 0.33, *p* = 0.351), neither in the 18.4 to 48 months old group ($\beta$ = 0.44, CI~95%~[-1.27, 2.16], $t_{student}$ (224.74) = 0.42, *p* = 0.577), whereas the effect tend to be positive but non-significant.
-The relationship between developmental domains, corrected age and their effect derivatives can be seen in [Figure 2](#fig2).
-
-```{r, eval=FALSE}
-set.seed(123)
-mod <- gam(
-  motora_gruesa_total ~ 
-    s(profesional_id, bs = "re") + 
-    s(sexo_paciente, bs = "re") + 
-    s(respondedor_vinculo, bs = "re") +
-    s(edad_corregida_meses), 
-  data = dataset, 
-  method = "REML"
-)
-
-report_smooth(mod)
-report_overall(mod, "edad_corregida_meses") |> cat();
-report_slopes(mod, "edad_corregida_meses")
-```
-
-When analyzing the motor skills domain, we found a significant non-linear effect of corrected age on GM scores, $F_{smooth}$ (5.25, 226.82) = 5.6, *p* < 0.001, which had an overall positive effect ($\beta$ = 1.86, CI~95%~[0.57, 3.16], $t_{student}$ (226.82) = 2.83, *p* = 0.005), however, the slope varied as a function of age, with a negative effect in the 0 to 6.8 age range ($\beta$ = -3.02, CI~95%~[-4.63, -1.42], $t_{student}$ (226.82) = -3.81, *p* = 0.003), but in the 10.2 to 15.5 interval, this relationship was inverted ($\beta$ = 1.83, CI~95%~[0.57, 3.08], $t_{student}$ (226.82) = 2.87, *p* = 0.008), however, in the rest of the age range the slope was non-significant and virtually zero (Age~[7.3, 9.7]~, $\beta$ = 0.13, CI~95%~[-1.02, 1.27], $t_{student}$ (226.82) = 0.23, *p* = 0.387; Age~[16, 48]~, $\beta$ = -0.06, CI~95%~[-2.09, 1.97], $t_{student}$ (226.82) = 0.01, *p* = 0.701).
-
-```{r, eval=FALSE}
-set.seed(123)
-mod <- gam(
-  motora_fina_total ~ 
-    s(profesional_id, bs = "re") + 
-    s(sexo_paciente, bs = "re") + 
-    s(respondedor_vinculo, bs = "re") +
-    s(edad_corregida_meses), 
-  data = dataset, 
-  method = "REML"
-)
-
-report_smooth(mod)
-report_overall(mod, "edad_corregida_meses");
-report_slopes(mod, "edad_corregida_meses");
-```
-
-Although a similar non-linear effect was observed when inspecting the influence of corrected age in the FM domain scores ($F_{smooth}$ (2.56, 228.03) = 4.11, *p* = 0.006), it was not possible to estimate a significant overall effect different from zero ($\beta$ = 0.01, CI~95%~[-0.47, 0.49], $t_{student}$ (228.03) = 0.03, *p* = 0.977), nevertheless, it was only in the 22.3 to 37.3 age range where a significant and negative effect was observed ($\beta$ = -0.75, CI~95%~[-1.4, -0.11], $t_{student}$ (228.03) = -2.32, *p* = 0.023).
-
-```{r, eval=FALSE}
-set.seed(123)
-mod <- gam(
-  resolucion_problemas_total ~ 
-    s(profesional_id, bs = "re") + 
-    s(sexo_paciente, bs = "re") + 
-    s(respondedor_vinculo, bs = "re") +
-    s(edad_corregida_meses), 
-  data = dataset, 
-  method = "REML"
-)
-
-report_smooth(mod);
-report_overall(mod, "edad_corregida_meses");
-report_slopes(mod, "edad_corregida_meses");
-```
-
-CG abilities were significantly influenced by corrected age ($F_{smooth}$ (5.69, 225.93) = 3.7, *p* = 0.001), with an overall negative effect ($\beta$ = -1.93, CI~95%~[-3.24, -0.63], $t_{student}$ (225.93) = -2.91, *p* = 0.004), and just like the other domains, this relationship was modified across the corrected age.
-In this sense, from the 0 to 5.8 age interval, we found that for every increase in one month in corrected age, we can expect a proportional increase in 2.8 points ($\beta$ = 2.8, CI~95%~[1.17, 4.43], $t_{student}$ (225.93) = 3.48, *p* = 0.002) in the CG domain, while in the age range 9.2 to 14.1 the relationship changes inversely, mainly because in this age range we observe that for every one-month increase in the corrected age, a decrease of 1.64 points could be expected in the same domain ($\beta$ = -1.64, CI~95%~[-2.86, -0.41], $t_{student}$ (225.93) = -2.62, *p* = 0.014).
-The other age intervals did not have a slope that deviated significantly from zero (Age~[6.3, 8.7]~, $\beta$ = 0.06, CI~95%~[-1.09, 1.2], $t_{student}$ (225.93) = 0.07, *p* = 0.388; Age~[14.5, 48.0]~, $\beta$ = 0.04, CI~95%~[-1.98, 2.07], $t_{student}$ (225.93) = -0.05, *p* = 0.564).
-
-```{r, eval=FALSE}
-set.seed(123)
-mod <- gam(
-  socio_individual_total ~ 
-    s(profesional_id, bs = "re") + 
-    s(sexo_paciente, bs = "re") + 
-    s(respondedor_vinculo, bs = "re") +
-    s(edad_corregida_meses), 
-  data = dataset, 
-  method = "REML"
-)
-
-report_smooth(mod);
-```
-
-```{r, eval=FALSE}
-y_vars <- c("CM" = "comunicacion_total"
-            , "FM" = "motora_fina_total"
-            , "GM" = "motora_gruesa_total"
-            , "CG" = "resolucion_problemas_total"
-            , "PS" = "socio_individual_total")
-
-many_models <- sapply(y_vars, function(i) {
-  form <- as.formula(
-    paste0(i, " ~ ", 
-           "s(profesional_id, bs = \"re\") + 
-            s(sexo_paciente, bs = \"re\") + 
-            s(respondedor_vinculo, bs = \"re\") +",
-           "s(semanas_prematurez, k = 9)")
-  )
-  model <- gam(
-    form, 
-    data = dataset, 
-    method = "REML"
-  )
-  report_smooth(model)
-}, simplify = FALSE)
-
-prematurity_smooths <- rbindlist(many_models, idcol = "var"
-          )[i = term %like% "semanas_prematurez",
-            j = paste(
-              var, 
-              regmatches(expr, regexec("\\*p\\* = 0.[[:digit:]]+", expr)), 
-              sep = ", ", 
-              collapse = "; "
-            )];
-cat(prematurity_smooths)
-```
-
-Unlike the others, the PS domain was not influenced by corrected age ($F_{smooth}$ (1, 230.4) = 1.06, *p* = 0.304).
-Accordingly, prematurity (measured in weeks) was not associated with any developmental domain within ASQ-3 assessment (significance for smooth terms: CM, *p* = 0.639; FM, *p* = 0.918; GM, *p* = 0.33; CG, *p* = 0.323; PS, *p* = 0.152).
-
-# Discussion
-
-Our study aimed to describe and model the relationship between sociodemographic data, prematurity and neurodevelopmental levels based on ASQ-3 scores in infants diagnosed with CH.
-Our main findings suggest a non-linear effect of age, corrected for prematurity, with a marked decrease in scores for all neurodevelopmental traits at different age frames, even after adjusting for caregiver relationship, sex and inter-rater influence.
-However no variation was observed across the corrected age in the PS domain.
-
-These findings confirm the described motor impairments of hypotonia in the early stages of life, which compromise the infant’s ability to explore and interact with their environment [@gabis2021weak; @harris2008congenital].
-A reflection of those mentioned above would be expressed in altered development of GM function in the first months of life, with a consequent limitation in FM skills later on, which would have a subsequent negative impact on the communicative competence of infants, secondary to reduced interaction with their environment and peers [@bodensteiner2008evaluation].
-In a recent systematic review [@gonzalez2019gross], differences in the predictive abilities of gross and FM skills on communication skills in infants and early childhood were reported.
-Whereas GM skills, such as crawling and walking, favor exploration with their environment and caregivers, while FM skills, expressed through tasks such as drawing and handling utensils, could lead to improvements in language through mechanisms yet to be explored [@gonzalez2019gross].
-These milestones may be impaired in the face of poorer head and trunk control in CH, which have been shown to delay the achievement of crucial motor milestones in infants [@bodensteiner2008evaluation; @gabis2021weak; @harris2008congenital].
-
-It is worth noting that other context-mediated social factors may also have influenced our results, mainly due to the role that other variables would also play in the neurodevelopment of our study sample, which could have an impact on many of the developmental traits assessed here, such as intrauterine growth restrictions, maternal depression, institutionalization, exposure to social violence, maternal education and breastfeeding [@walker2011inequality].
-Altogether, these represent the main limitations in our study design, which need to be addressed in future research, exploring the variations observed in different developmental traits in hypotonic infants.
-However, our study sheds light on an underexplored aspect with robust statistical methods that enabled us to capture and model the complex relationships seen early in life.
-
-# Conclusion
-
-The present study shows that the marked variations observed in neurodevelopmental traits are present across age in hypotonic infants, mainly in the form of non-linear and domain-specific variations, even after adjusting for the effect that caregiver relationship, sex and evaluators might exert.
-Moreover, we show that the observed variations in developmental domains are not solely attributable to prematurity, where age corrected for prematurity best explains the observed variability in neurodevelopment.
-Future research may determine how these findings apply when controlling for context-mediated social factors and other clinical populations.
-
-# Data availability statement
-
-The raw data supporting the conclusions of this article will be made available by the authors without undue reservation.
-
-# Author Contributions
-
-All authors listed have made a substantial, direct and intellectual contribution to the work and approved it for publication.
-
-# Conflicts of interest
-
-The authors declare that the research was conducted without any commercial or financial relationships construed as a potential conflict of interest.
-
-# ORCID
-
-::: {custom-style="Afiliation"}
-Patricio Barría, 0000-0003-4088-9820
-
-Matías Castillo-Aguilar, 0000-0001-7291-247X
-
-Katherine Harris, 0000-0003-4271-6405
-
-Kay Gittermann, 0000-0002-8042-844X
-
-Mónica Bontes, 0000-0002-0592-7361
-
-Rolando Aguilar, 0000-0002-0104-3032
-
-Carlos Cifuentes, 0000-0002-6942-865X
-
-Marcela Munera, 0000-0001-6595-5383
-
-Cristian Núñez-Espinosa, 0000-0002-9896-7062
-:::
-
-# References
-
-<div id="refs"></div>
-
-\newpage
-
-\{Add table\}
-
-\newpage
 

### 2023-04-23 manuscript/manuscript.Rmd
presentacion para el CICYPE
update on paper files, going for the third try!

@@
 
-**Aim**: To describe and model the relationship between sociodemographics, prematurity and neurodevelopmental levels based on the *Ages and Stages Questionnaire* (ASQ-3) scores in infants diagnosed with congenital hypotonia (CH).
-**Method**: A total of 234 patients diagnosed with CH participated in this study (median age, 8 years; 94 (40.2%) females and 140 (59.8%) males).
+**Objective**: To describe and model the relationship between sociodemographics, prematurity and neurodevelopmental levels based on the *Ages and Stages Questionnaire* (ASQ-3) scores in infants diagnosed with congenital hypotonia (CH).
+**Method**: A total of 234 patients diagnosed with CH participated in this study (median age, 8 years; 93 (40.6%) females and 136 (59.4%) males).
 Neurodevelopmental status was assessed with the ASQ-3 at admission, as well as sociodemographic and obstetric data obtained through the initial clinical interview.
@@
 Fine motor skills did not exhibit a linear relationship with corrected age but a non-linear effect, unlike the personal-social domain, which did not present significant variations across age corrected for prematurity.
-**Interpretation**: After adjusting for possible confounders, we found between-subjects fluctuations in neurodevelopmental traits across age in hypotonic infants.
+**Conclusion**: After adjusting for possible confounders, we found between-subjects fluctuations in neurodevelopmental traits across age in hypotonic infants.
 These fluctuations were present in the form of non-linear and domain-specific variations.
@@
 In this sense, the Ages and Stages Questionnaire, Third Edition (ASQ-3), has been presented as a global screening tool aimed at parents/caregivers, assessing five developmental domains in children aged 0 to 5.5 years [@singh2017ages].
-Current evidence suggests that the ASQ-3 is an accurate, cost-effective, and parent-friendly instrument for screening and monitoring children through pre-school age, helping to identify and eliminate neurodevelopmental deficits in children born very premature [@kerstjens2015ages; @ballantyne2016risk; @singh2017ages].
+Current evidence suggests that the ASQ-3 is an accurate, cost-effective, and parent-friendly instrument for screening and monitoring children through pre-school age, helping to identify and prevent neurodevelopmental deficits in children born very premature [@kerstjens2015ages; @ballantyne2016risk; @singh2017ages].
 
@@
 Some authors suggest that CH cannot be referred to as a diagnosis [@thompson2002benign].
-Those who do consider it point it out as a diagnosis of exclusion in the investigation, considering it only when it is established in the absence of other signs and symptoms [@gabis2021weak; @leyenaar2005schematic].
-However, its research is fundamental since it is a non-progressive neuromuscular disorder, which tends to improve with time and early intervention [@gabis2021weak].
+Those who do consider it point it out as a diagnosis of exclusion, considering it only when it is established in the absence of other signs and symptoms [@gabis2021weak; @leyenaar2005schematic].
+However, its research is fundamental since it is a non-progressive neuromuscular disorder, which tends to improve with early intervention [@gabis2021weak].
 
-To our knowledge, there is no substantial evidence characterizing the observed variation in developmental traits in infants with diagnosed CH across ages.
-Therefore, our primary goal in this study was to describe and model the relationship between sociodemographic characteristics, prematurity, and neurodevelopmental levels as a function of ASQ-3 estimates in infants diagnosed with CH.
+To our knowledge, there is no substantial evidence characterizing the variations in developmental traits in infants with diagnosed CH across ages.
+Therefore, our primary goal in this study was to describe and model the relationship between prematurity, and neurodevelopmental levels assessed through ASQ-3 as a function of age in infants diagnosed with CH.
 
@@
 print(n_sample);
-#> 234
+#> 229
 ```
 
-A total of 234 patients were enrolled as part of an intervention program carried out by the Cruz del Sur Rehabilitation Center (Punta Arenas, Chile), admitted from one month to 60 months old, and evaluated at admission, control and discharge.
+A total of 229 patients were enrolled as part of an intervention program carried out by the Cruz del Sur Rehabilitation Center (Punta Arenas, Chile), admitted from one month to 35 months old, and evaluated at admission, control and discharge.
 Participating patients were diagnosed based on clinical criteria confirmed by a neurologist and a physiatrist.
@@
 
-In order to assess the influence of confounding variables, we fitted different models using the variables: i) sex of the infant, ii) clinician ID and iii) the relationship of the infant to the respondent (i.e., whether the respondent was the infant's mother, father, uncle or grandparent) as random effects in the form of penalized parametric terms [@wood2016smoothing].
-Varying random effects structures models can be seen in the section *Random effects structures* of the supplementary file.
+In order to assess the influence of confounding variables, we fitted different models using the categorical variables: i) sex of the infant, ii) clinician ID and iii) the relationship of the infant to the respondent (i.e., whether the respondent was the infant's mother, father, uncle or grandparent) as random effects in the form of penalized parametric terms [@wood2016smoothing].
+Varying random effects structures and model performance metrics can be seen in the supplementary file.
 
 Based on the technical report of the ASQ-3, we've been able to estimate the threshold of possible delay based on the total score for each developmental domain.
-In this sense, a logistic regression, within GAM framework, was used to estimate the predicted probability of having a possible delay in developmental domains given the corrected age.
+Within this framework, a logistic regression was used to estimate the predicted probability of a developmental delay given the corrected age for each domain.
 
-A probability of committing a type I error ($\alpha$) of less than 5% (*p* < 0.05) was considered sufficient evidence for statistical significance in hypothesis testing.
+A probability of committing a type I error of less than 5% was considered sufficient evidence for null hypothesis significance testing.
 All the statistical analyses were computed and implemented in the R programming language [@rlanguage].
@@
 #>    sex        stat
-#> 1:   F  94 (40.2%)
-#> 2:   M 140 (59.8%)
+#> 1:   F  93 (40.6%)
+#> 2:   M 136 (59.4%)
 
 # Prop test on sex
+
 prop_sex <- dataset[, table(sexo_paciente)] |> 
@@
 cat(prop_sex);
-#> $\chi^2_{gof}$ (1) = 9.04, *p* = 0.003
+#> $\chi^2_{gof}$ (1) = 8.07, *p* = 0.004
 ```
 
-From a total of 234 subjects with congenital hypotonia, 94 (40.2%) were females and 140 (59.8%) males ($\chi^2_{gof}$ (1) = 9.04, *p* = 0.003).
+From a total of 229 subjects with congenital hypotonia, 93 (40.6%) were females and 136 (59.4%) males ($\chi^2_{gof}$ (1) = 8.07, *p* = 0.004).
 The developmental characteristics of the sample can be seen in [Table 1](#tab1).
+
+## Neurodevelopment as a function of corrected age
 
@@
 
-After fitting a GAM to assess the effect of corrected age on CM scores, we observed a significant non-linear relationship between them ($F_{smooth}$ (5.1, 224.74) = 13.88, *p* < 0.001), that reflect an overall negative marginal trend ($\beta$ = -2.39, CI~95%~[-3.46, -1.32], $t_{student}$ (224.74) = -4.4, *p* < 0.001), however, this was not true when assessing the direction of the effect in the age range between 0 to 6.8 ($\beta$ = 0.38, CI~95%~[-0.96, 1.72], $t_{student}$ (224.74) = 0.33, *p* = 0.351), neither in the 18.4 to 48 months old group ($\beta$ = 0.44, CI~95%~[-1.27, 2.16], $t_{student}$ (224.74) = 0.42, *p* = 0.577), whereas the effect tend to be positive but non-significant.
+### Communication
+
+After fitting a GAM to assess the effect of corrected age on CM scores, we observed a significant non-linear relationship between them ($F_{smooth}$ (4.81, 219.93) = 15.39, *p* < 0.001), that reflect an overall negative marginal trend ($\beta$ = -2.06, CI~95%~[-3.32, -0.8], $t_{student}$ (219.93) = -3.23, *p* = 0.001), however, this was not true when assessing the direction of the effect in the age range between 0 to 6.7 ($\beta$ = 0.57, CI~95%~[-1.01, 2.16], $t_{student}$ (219.93) = 0.43, *p* = 0.288), neither in the 17.3 to 35 months old group ($\beta$ = 0.2, CI~95%~[-1.7, 2.11], $t_{student}$ (219.93) = 0.16, *p* = 0.677), whereas the effect tend to be positive but non-significant.
 The relationship between developmental domains, corrected age and their effect derivatives can be seen in [Figure 2](#fig2).
@@
 
-When analyzing the motor skills domain, we found a significant non-linear effect of corrected age on GM scores, $F_{smooth}$ (5.25, 226.82) = 5.6, *p* < 0.001, which had an overall positive effect ($\beta$ = 1.86, CI~95%~[0.57, 3.16], $t_{student}$ (226.82) = 2.83, *p* = 0.005), however, the slope varied as a function of age, with a negative effect in the 0 to 6.8 age range ($\beta$ = -3.02, CI~95%~[-4.63, -1.42], $t_{student}$ (226.82) = -3.81, *p* = 0.003), but in the 10.2 to 15.5 interval, this relationship was inverted ($\beta$ = 1.83, CI~95%~[0.57, 3.08], $t_{student}$ (226.82) = 2.87, *p* = 0.008), however, in the rest of the age range the slope was non-significant and virtually zero (Age~[7.3, 9.7]~, $\beta$ = 0.13, CI~95%~[-1.02, 1.27], $t_{student}$ (226.82) = 0.23, *p* = 0.387; Age~[16, 48]~, $\beta$ = -0.06, CI~95%~[-2.09, 1.97], $t_{student}$ (226.82) = 0.01, *p* = 0.701).
+### Gross motor
+
+When analyzing the motor skills domain, we found a significant non-linear effect of corrected age on GM scores, $F_{smooth}$ (4.75, 222.3) = 6.31, *p* < 0.001, which had an overall positive effect ($\beta$ = 1.77, CI~95%~[0.31, 3.24], $t_{student}$ (222.3) = 2.38, *p* = 0.018), however, the slope varied as a function of age, with a negative effect in the 0 to 6.7 age range ($\beta$ = -3.14, CI~95%~[-4.99, -1.3], $t_{student}$ (222.3) = -3.5, *p* = 0.005), but in the 10.3 to 15.2 interval, this relationship was inverted ($\beta$ = 1.9, CI~95%~[0.42, 3.38], $t_{student}$ (222.3) = 2.53, *p* = 0.015), however, in the rest of the age range the slope was non-significant and virtually zero (Age~[7.1, 9.9]~, $\beta$ = 0.27, CI~95%~[-1.11, 1.66], $t_{student}$ (222.3) = 0.32, *p* = 0.369; Age~[15.6, 35]~, $\beta$ = 0.24, CI~95%~[-1.92, 2.4], $t_{student}$ (222.3) = 0.29, *p* = 0.763).
 
@@
 
-Although a similar non-linear effect was observed when inspecting the influence of corrected age in the FM domain scores ($F_{smooth}$ (2.56, 228.03) = 4.11, *p* = 0.006), it was not possible to estimate a significant overall effect different from zero ($\beta$ = 0.01, CI~95%~[-0.47, 0.49], $t_{student}$ (228.03) = 0.03, *p* = 0.977), nevertheless, it was only in the 22.3 to 37.3 age range where a significant and negative effect was observed ($\beta$ = -0.75, CI~95%~[-1.4, -0.11], $t_{student}$ (228.03) = -2.32, *p* = 0.023).
+### Fine motor
+
+Although a similar non-linear effect wasn't observed when inspecting the influence of corrected age in the FM domain scores on the global test statistics ($F_{smooth}$ (2.56, 228.03) = 4.11, *p* = 0.006), and neither was possible to estimate a significant overall effect different from zero ($\beta$ = 0.01, CI~95%~[-0.47, 0.49], $t_{student}$ (228.03) = 0.03, *p* = 0.977), nevertheless, it was only in the 23 to 28.3 age range where a significant and negative effect was observed ($\beta$ = -0.8, CI~95%~[-1.55, -0.04], $t_{student}$ (223.23) = -2.09, *p* = 0.038).
 
@@
 
-CG abilities were significantly influenced by corrected age ($F_{smooth}$ (5.69, 225.93) = 3.7, *p* = 0.001), with an overall negative effect ($\beta$ = -1.93, CI~95%~[-3.24, -0.63], $t_{student}$ (225.93) = -2.91, *p* = 0.004), and just like the other domains, this relationship was modified across the corrected age.
-In this sense, from the 0 to 5.8 age interval, we found that for every increase in one month in corrected age, we can expect a proportional increase in 2.8 points ($\beta$ = 2.8, CI~95%~[1.17, 4.43], $t_{student}$ (225.93) = 3.48, *p* = 0.002) in the CG domain, while in the age range 9.2 to 14.1 the relationship changes inversely, mainly because in this age range we observe that for every one-month increase in the corrected age, a decrease of 1.64 points could be expected in the same domain ($\beta$ = -1.64, CI~95%~[-2.86, -0.41], $t_{student}$ (225.93) = -2.62, *p* = 0.014).
-The other age intervals did not have a slope that deviated significantly from zero (Age~[6.3, 8.7]~, $\beta$ = 0.06, CI~95%~[-1.09, 1.2], $t_{student}$ (225.93) = 0.07, *p* = 0.388; Age~[14.5, 48.0]~, $\beta$ = 0.04, CI~95%~[-1.98, 2.07], $t_{student}$ (225.93) = -0.05, *p* = 0.564).
+### Problem solving
+
+CG abilities were significantly influenced by corrected age ($F_{smooth}$ (5.28, 220.66) = 4.18, *p* < 0.001), with an overall negative effect ($\beta$ = -1.99, CI~95%~[-3.5, -0.48], $t_{student}$ (220.66) = -2.6, *p* = 0.01), and just like the other domains, this relationship was modified across the corrected age.
+In this sense, from the 0 to 5.7 age interval, we found that for every increase in one month in corrected age, we can expect a proportional increase in 2.94 points ($\beta$ = 2.94, CI~95%~[1, 4.88], $t_{student}$ (220.66) = 3.11, *p* = 0.007) in the CG domain, while in the age range 9.5 to 13.4 the relationship changes inversely, mainly because in this age range we observe that for every one-month increase in the corrected age, a decrease of 1.93 points could be expected in the same domain ($\beta$ = -1.93, CI~95%~[-3.44, -0.41], $t_{student}$ (220.66) = -2.51, *p* = 0.017).
+The other age intervals did not have a slope that deviated significantly from zero (Age~[6, 9.2]~, $\beta$ = -0.08, CI~95%~[-1.45, 1.29], $t_{student}$ (220.66) = -0.08, *p* = 0.358; Age~[13.8, 35]~, $\beta$ = -0.16, CI~95%~[-2.34, 2.01], $t_{student}$ (220.66) = -0.18, *p* = 0.507).
 
@@
 report_smooth(mod);
+report_slopes(mod, term = "edad_corregida_meses");
 ```
@@
 
-Unlike the others, the PS domain was not influenced by corrected age ($F_{smooth}$ (1, 230.4) = 1.06, *p* = 0.304).
-Accordingly, prematurity (measured in weeks) was not associated with any developmental domain within ASQ-3 assessment (significance for smooth terms: CM, *p* = 0.639; FM, *p* = 0.918; GM, *p* = 0.33; CG, *p* = 0.323; PS, *p* = 0.152).
+### Personal-social
+
+Unlike the others, the PS domain was not influenced by corrected age ($F_{smooth}$ (1, 225.36) = 0.99, *p* = 0.321), given that the marginal effect across age was virtually zero ($\beta$ = -0.09, CI~95%~[-0.28, 0.09]).
+
+## Prematurity
+
+Accordingly, prematurity (measured in weeks) was not associated with any non-linear variations in neurodevelopmental domains within ASQ-3 assessment (significance for smooth terms: CM, *p* = 0.594; FM, *p* = 0.901; GM, *p* = 0.342; CG, *p* = 0.371; PS, *p* = 0.172).
 
@@
 models <- mapply(
-  FUN = gam_ordinal,
+  FUN = gam_binomial,
   var = ind,
-  data = list(dataset),
+  data = list(dataset = dataset),
   plot = FALSE,
@@
 
-When analyzing the probability of having a developmental delay based on the scores for each domain, we found a significant non-linear effect for CM ($\chi^2_{smooth}$ (7.73) = 32.11, *p* < 0.001), whereas in this domain we observed that for every one-month increase in the corrected age in the 13.6 to 18.9 age range, the odds of a possible developmental delay increases by 60% (OR = 1.6 CI~95%~[1.16, 2.23], $t_{student}$ (224.56) = 2.89, *p* = 0.01), as well with the 28.6 to 29.6 age range, whereas the odds of delay increases by 36% in this domain (OR = 1.36 CI~95%~[1.00, 1.86], $t_{student}$ (224.56) = 1.98, *p* = 0.049).
-Similar to CM, an overall linear effect of corrected age in the GM domain was observed ($\chi^2_{smooth}$ (1) = 6.86, *p* = 0.009), whereas for every one-month increase the odds of a possible developmental delay decrease by a 3.9% (OR = 0.96, CI~95%~[0.50, 0.99], $t_{student}$ (232) = -2.62, *p* = 0.009).
+## Risk of developmental delay and corrected age
+
+When analyzing the probability of having a developmental delay based on the scores for each domain, we found a significant non-linear effect for CM ($\chi^2_{smooth}$ (2.83) = 27.71, *p* < 0.001), whereas in this domain we observed that for every one-month increase in the corrected age in the 9.9 to 31.1 age range, the odds of a possible developmental delay increases by 20% (OR = 1.2, CI~95%~[1.05, 1.37], $t_{student}$ (223.84) = 2.83, *p* = 0.01).
+Similar to CM, an overall linear effect of corrected age in the GM domain was observed ($\chi^2_{smooth}$ (1) = 8.88, *p* = 0.003), whereas for every one-month increase the odds of a possible developmental delay decrease by a 5% (OR = 0.95, CI~95%~[0.92, 0.98], $t_{student}$ (227) = -2.98, *p* = 0.003).
 The probability associated with a possible delay according to the total score for each developmental domain in response to corrected age can be observed in [Figure 3](#fig3).
@@
 
-Our study aimed to describe and model the relationship between sociodemographic data, prematurity and neurodevelopmental levels based on ASQ-3 scores in infants diagnosed with CH.
-Our main findings suggest a non-linear effect of age, corrected for prematurity, with a marked decrease in scores for all neurodevelopmental traits at different age frames, even after adjusting for caregiver relationship, sex and inter-rater influence.
-However no variation was observed across the corrected age in the PS domain.
+The objective of our investigation was to elucidate and conceptualize the association between prematurity and neurodevelopmental functioning, evaluated through the Ages and Stages Questionnaires-Third Edition (ASQ-3), as a function of age in infants diagnosed with congenital heart disease (CH).
+Our principal discoveries indicate a curvilinear influence of age, which was adjusted for prematurity, resulting in a significant reduction of essential neurodevelopmental characteristics at distinct age intervals, while accounting for the impact of caregiver rapport, gender, and inter-rater bias.
+However, no fluctuation was detected in the PS domain across the adjusted age.
 
 These findings confirm the described motor impairments of hypotonia in the early stages of life, which compromise the infant’s ability to explore and interact with their environment [@gabis2021weak; @harris2008congenital].
-A reflection of those mentioned above would be expressed in altered development of GM function in the first months of life, with a consequent limitation in FM skills later on, which would have a subsequent negative impact on the communicative competence of infants, secondary to reduced interaction with their environment and peers [@bodensteiner2008evaluation].
+The limitations in GM function during the early stages of life are expected to result in impaired FM skills during subsequent developmental stages.
+Consequently, these deficiencies may negatively impact the communicative capabilities of infants, as their interaction and engagement with their environment and peers would be reduced.
+[@bodensteiner2008evaluation].
 In a recent systematic review [@gonzalez2019gross], differences in the predictive abilities of gross and FM skills on communication skills in infants and early childhood were reported.
@@
 <div id="refs"></div>
-
-\newpage
-
-Añadir tabla
-
-\newpage
-
-[**Table 1**]{#tab1}.
-Overall baseline and developmental characteristics of the sample and grouped by sex.
-^1^ Data is presented as sample size, and *Mdn* (*IQR*); ^2^ p-values are computed from the *Wilcoxon* rank-sum test.
-
-[**Figure 1**]{#fig1}. Predicted scores on developmental domains using corrected age as an independent variable, comparing different random effects structures on the estimated score.
-
-[**Figure 2**]{#fig2}.
-Relationship between corrected age (in months) and developmental domains.
-Left panel: regression lines represent predicted values estimated from GAM models (bold red lines) and 200 bootstrap replicates (faded red lines), points and error bars represent the mean and standard error at 5-month age intervals.
-Right panel: effect derivatives and their CI~95%~, representing how the effect of corrected age (in months) in developmental domains changes across corrected age.
-Significant areas consider CI~95%~ that do not cross zero.
-
-[**Figure 3**]{#fig3}.
-Predicted probability of having a possible delay according to the total score for each developmental domain in response to corrected age.
-Fitted models were estimated using logistic regression within GAM framework and adjusted for possible confounders as already described.
-The p-values of smooth terms, associated with the main effect of corrected age on developmental delay, are shown within each panel.

### 2023-04-23 R/plot_gam.R

@@ -77,6 +77,19 @@ plot_gam <- function(data, x, y, xlab, ylab, round_to = 1, ...) {
 #' @export
 
 gam_confounders <- function(data, var, seed = 1234, legend = TRUE, var_name = NULL, plot = TRUE, ...) {
+  # y_vars <- c("CM" = "comunicacion_total"
+  #             , "GM" = "motora_gruesa_total"
+  #             , "FM" = "motora_fina_total"
+  #             , "CG" = "resolucion_problemas_total"
+  #             , "PS" = "socio_individual_total")
+  #
+  # fig_data <- melt(
+  #   data = data,
+  #   measure.vars = y_vars
+  # )[variable == var]
+  #
+  # fig_data[, variable := `levels<-`(variable, names(y_vars))][]
+
   models = list(
     # Simple model - only smooth term
     `Simple` = "var ~ s(edad_corregida_meses)",
@@ -129,16 +142,14 @@ gam_confounders <- function(data, var, seed = 1234, legend = TRUE, var_name = NU
 
   ylab <- "Predicted score"
   if (!is.null(var_name) && is.character(var_name)) {
-    ylab <- paste("Predicted", var_name, "score")
+    ylab <- paste(var_name, "score")
   }
 
-  plot <- ggplot2::ggplot(predicted_response, ggplot2::aes(edad_corregida_meses, fit)) +
-    ggplot2::facet_grid(rows = ggplot2::vars(sexo_paciente), cols = ggplot2::vars(respondedor_vinculo),
-                        labeller = ggplot2::labeller(
-                          sexo_paciente = c(F = "Female", M = "Male"),
-                          respondedor_vinculo = c(Madre = "Mother", Padre = "Father", `Abuelo/a` = "Grandparent", `Tio/a` = "Uncle")
-                        )) +
-    ggplot2::geom_line(ggplot2::aes(col = `Adjusted for:`, lty = `Adjusted for:`)) +
+  plot_data <- predicted_response[, list(fit = mean(fit)), by = list(edad_corregida_meses ,`Adjusted for:`)]
+
+  plot <- ggplot2::ggplot(plot_data, ggplot2::aes(edad_corregida_meses, fit)) +
+    # ggplot2::geom_count(data = fig_data, ggplot2::aes(edad_corregida_meses, value), alpha = 0.1, na.rm = TRUE) +
+    ggplot2::geom_line(ggplot2::aes(col = `Adjusted for:`, lty = `Adjusted for:`), linewidth = 1) +
[…]

### 2023-04-23 manuscript/figures/fig-1.R

@@ -14,40 +14,20 @@ plots <- mapply(
   data = list(dataset),
   var = params,
   var_name = names(params),
-  legend = c(T,T,T,F,F),
+  legend = c(T,T,F,F,F),
   SIMPLIFY = FALSE
 )
 
-f0 <- ggpubr::ggarrange(plotlist = plots[1:3], nrow = 1, common.legend = TRUE)
-f1 <- ggpubr::ggarrange(plotlist = plots[4:5], nrow = 1)
-figure1 <- ggpubr::ggarrange(f0, f1, ncol = 1)
+f0 <- ggpubr::ggarrange(plotlist = plots[1:2], nrow = 1, common.legend = TRUE)
+f1 <- ggpubr::ggarrange(plotlist = plots[3], nrow = 1)
+f2 <- ggpubr::ggarrange(plotlist = plots[4:5], nrow = 1)
+figure1 <- ggpubr::ggarrange(f0, f1, f2, ncol = 1)
 
 local({
-  pdf("manuscript/figures/fig-1.pdf", width = 15, height = 10);
+  pdf("manuscript/figures/fig-1.pdf", width = 8, height = 10);
+  print(figure1);
+  dev.off()
+  tiff("manuscript/figures/fig-1.tiff", width = 8, height = 10, units = "in", res = 400);
   print(figure1);
   dev.off()
 })
-
-# Comparing models
-mod_comparison <- lapply(models, function(i) {
-  out <- as.data.table(
-    x = do.call(
-    what = AIC,
-    args = unname(i)
-    )
-  )
-  out[, model := names(i)][]
-})
-
-mod_comparison <- rbindlist(mod_comparison, idcol = "Response")
-best_worst_mods <- mod_comparison[, .SD[AIC %in% range(AIC), list(AIC, model)], list(Response)]
-
-best_worst_mods[, Classification := ifelse(max(AIC) == AIC, "Worst", "Best"), Response
-                ][, tapply(paste0(round(AIC), " (", model, ")"), Classification, c, simplify = FALSE), list(Response)]
-
-# Evaluation smooth significance
-lapply(models, function(i) {
-  lapply(i, report_smooth)
-})
-
-models$CM$`Full model` |> plot()

### 2023-04-23 manuscript/figures/fig-2.R

@@ -165,7 +165,10 @@ fig_2b <- ggplot(fig_2b_data, aes(edad_corregida_meses, Coefficient)) +
 fig_2 <- ggpubr::ggarrange(fig_2a, fig_2b, ncol = 2)
 
 if (isTRUE(x = .save_plot)) {
-  pdf("manuscript/figures/fig-2.pdf", width = 12, height = 10);
+  pdf("manuscript/figures/fig-2.pdf", width = 8, height = 10);
+  print(fig_2);
+  dev.off()
+  tiff("manuscript/figures/fig-2.tiff", width = 8, height = 10, units = "in", res = 400);
   print(fig_2);
   dev.off()
 }

### 2023-04-23 manuscript/figures/fig-3.R

@@ -27,7 +27,10 @@ f2 <- ggpubr::ggarrange(plotlist = plots[4:5], nrow = 1)
 figure3 <- ggpubr::ggarrange(f0, f1, f2, ncol = 1)
 
 local({
-  pdf("manuscript/figures/fig-3.pdf", width = 10, height = 12);
+  pdf("manuscript/figures/fig-3.pdf", width = 8, height = 10);
+  print(figure3);
+  dev.off()
+  tiff("manuscript/figures/fig-3.tiff", width = 8, height = 10, units = "in", res = 400);
   print(figure3);
   dev.off()
 })

### 2023-04-23 manuscript/figures/make_plots.R

@@ -1,0 +1,14 @@
+files <- grep(
+  pattern = "make_plots",
+  x = list.files(
+    path = "manuscript/figures",
+    pattern = "\\.R$",
+    full.names = TRUE
+  ),
+  invert = TRUE,
+  value = TRUE
+)
+
+for (file in files) {
+  source(file)
+}

### 2023-04-23 manuscript/make_ms.R

@@ -1,0 +1,43 @@
+## Cargamos librerias
+library(rmarkdown)
+library(officer)
+
+## Generamos los documentos
+files <- list.files(
+  path = "manuscript/",
+  pattern = "\\.Rmd$",
+  ignore.case = TRUE,
+  full.names = TRUE
+)
+
+## Usamos un bucle
+for (file in files) {
+  render(input = file)
+}
+
+## Adjuntamos la tabla al final
+read_docx("manuscript/manuscript.docx") |>
+  body_add_break() |>
+  body_add_docx(src = "manuscript/tables/table-1.docx") |>
+  body_add_break() |>
+  body_add_docx(src = "manuscript/captions.docx") |>
+  print(target = "manuscript/manuscript.docx")
+
+## Volvemos a generar los gráficos
+source("manuscript/figures/make_plots.R")
+
+zip::zip(
+  zipfile = "manuscript/submission_docs/ASQ3-files.zip",
+  files = c(
+    "manuscript/manuscript.docx",
+    "manuscript/supplementary.docx",
+    "manuscript/figures/fig-1.pdf",
+    "manuscript/figures/fig-1.tiff",
+    "manuscript/figures/fig-2.pdf",
+    "manuscript/figures/fig-2.tiff",
+    "manuscript/figures/fig-3.pdf",
+    "manuscript/figures/fig-3.tiff",
+    "manuscript/misc/STROBE checklist.doc"
+  ),
+  include_directories = FALSE
+)

### 2023-04-23 manuscript/misc/setup.R

@@ -4,6 +4,7 @@ library(data.table)
 
 # For GAMS
 library(mgcv)
+library(brms)
 
 # For tables
 library(gt)
@@ -16,7 +17,7 @@ data("dataset", package = "ASQ3")
 # 1. Includes only observations from assessment 1
 # 2. Includes subjects with diagnosis of congenital hypotonia
 # 3. Subjects whose age is not greater than 60 months old (chronological age)
-dataset <- dataset[i = n_evaluacion == 1 & diagnostico %like% "hipotonia" & edad_cronologica_meses <= 60]
+dataset <- dataset[i = n_evaluacion == 1 & diagnostico %like% "hipotonia" & edad_cronologica_meses <= 35]
 
 # Drop levels of unused factors
 fv <- dataset[, names(.SD), .SDcols = is.factor]

### 2023-04-23 manuscript/tables/table.R

@@ -52,7 +52,5 @@ tbl1 <- add_p(
 
 tbl1 <- as_gt(tbl1)
 
-
-
 saveRDS(tbl1, file = "manuscript/tables/table-1.RDS")
-gt::gtsave(tbl1, filename = "manuscript/tables/table-1.rtf")
+gt::gtsave(tbl1, filename = "manuscript/tables/table-1.html")

### 2023-04-23 manuscript/manuscript.Rmd
update on abstract

@@
 **Objective**: To describe and model the relationship between sociodemographics, prematurity and neurodevelopmental levels based on the *Ages and Stages Questionnaire* (ASQ-3) scores in infants diagnosed with congenital hypotonia (CH).
-**Method**: A total of 234 patients diagnosed with CH participated in this study (median age, 8 years; 93 (40.6%) females and 136 (59.4%) males).
+**Method**: A total of 229 patients diagnosed with CH participated in this study (median age, 8 years; 93 (40.6%) females and 136 (59.4%) males).
 Neurodevelopmental status was assessed with the ASQ-3 at admission, as well as sociodemographic and obstetric data obtained through the initial clinical interview.
 **Results**: When modeling the neurodevelopmental status of each domain, an overall negative effect of corrected age on communication and problem-solving skills was observed, whereas the overall effect tends to be positive in gross motor function.
-Fine motor skills did not exhibit a linear relationship with corrected age but a non-linear effect, unlike the personal-social domain, which did not present significant variations across age corrected for prematurity.
+Fine motor skills did not exhibit a linear relationship but a non-linear effect with corrected age, unlike the personal-social domain, which did not present significant variations across age corrected for prematurity.
 **Conclusion**: After adjusting for possible confounders, we found between-subjects fluctuations in neurodevelopmental traits across age in hypotonic infants.
